### 2019-12-12 Docs/Papers/ALifeCities/paper.tex
skeleton_paper

\documentclass[letterpaper]{article}

\usepackage{natbib,alifeconf}  %% The order is important

% *****************
%  Requirements:
% *****************
%
% - All pages sized consistently at 8.5 x 11 inches (US letter size).
% - PDF length <= 8 pages for full papers, <=2 pages for extended
%    abstracts (not including citations).
% - Abstract length <= 250 words.
% - No visible crop marks.
% - Images at no greater than 300 dpi, scaled at 100%.
% - Embedded open type fonts only.
% - All layers flattened.
% - No attachments.
% - All desired links active in the files.

% Note that the PDF file must not exceed 5 MB if it is to be indexed
% by Google Scholar. Additional information about Google Scholar
% can be found here:
% http://www.google.com/intl/en/scholar/inclusion.html.

% If your system does not generate letter format documents by default,
% you can use the following workflow:
% latex example
% bibtex example
% latex example ; latex example
% dvips -o example.ps -t letterSize example.dvi
% ps2pdf example.ps example.pdf

% For pdflatex users:
% The alifeconf style file loads the "graphicx" package, and
% this may lead some users of pdflatex to experience problems.
% These can be fixed by editing the alifeconf.sty file to specify:
% \usepackage[pdftex]{graphicx}
%   instead of
% \usepackage{graphicx}.
% The PDF output generated by pdflatex should match the required
% specifications and obviously the dvips and ps2pdf steps become
% unnecessary.

% Note:  Some laser printers have a serious problem printing TeX
% output. The use of ps type I fonts should avoid this problem.

\title{Cities as they could be: an Artificial Life approach to Urban Systems}
\author{Juste Raimbault$^{1,2,3}$\\
\mbox{}\\
$^1$Center for Advanced Spatial Analysis, University College London\\
$^2$UPS CNRS 3611 ISC-PIF\\
$^3$UMR CNRS 8504 G{\'e}ographie-cit{\'e}s
juste.raimbault@polytechnique.edu} % email of corresponding author

% For several authors from the same institution use the same number to
% refer to one address.
%
% If the names do not fit well on one line use
%         Author 1, Author 2 ... \\ {\Large\bf Author n} ...\\ ...
%
% If the title and author information do not fit in the area
% allocated, place \setlength\titlebox{<new height>} after the
% \documentclass line where <new height> is 2.25in

\begin{document}
\maketitle

\begin{abstract}
% Abstract length should not exceed 250 words

\end{abstract}

\section{Introduction}

\cite{kim2006comprehensive} applications of alife

\cite{pattee1995artificial} epistemic cut: simulation / hard or wet ALife / nature of computation

% urban DNA ?

\cite{torrens2003automata} CA and ABM

\cite{openshaw1995developing} GIS

% broad issues: computing territorial systems ? cultural evol to urban evol ?

\section{Artificial intelligence}

\cite{wu2010artificial} AI applied to land-use dynamics

\cite{white1989artificial}

\section{Artificial life}

\cite{medda2009morphogenetic} Turing equations

\cite{olsen1982urban} urban metabolism and cycles

\cite{kato1998alife} road network / building layout

\cite{kato2000modeling} CA/GA dynamical

% some procedural modeling ?

\section{Bibliometric analysis}

%%%%
% Method
% - keyword req (separate alife / AI ? - size pb) || initial corpus ?
% - cit nw construction
% Q: a priori or a posteriori analysis?

\section{Cities as they could be}

% How ALife concepts and approaches can help understanding / managing / planning / designing urban systems

% - biomimetism
% - morphogenesis / autopoiesis
% - tools: simulation models etc.
% - urban metabolism
% - process understanding: "the real smart"
% - innovation diffusion ? (! not too close from urb evol th: step back? )

\section{Discussion}

% what remains to be done
% / open problems in "urban evolution"
%
%  - proper definition of urban genotype / phenotype (higher dimension of cultural evol. ?)
%  - def and understanding of co-evolution
%. - understanding of morphogenesis - in the sense of urban form and function
%. - understanding of autopoiesis: urban boundaries is an open Q (<- use open Q as arg why analogy is relevant ?)

%\section{Acknowledgements}

\footnotesize
\bibliographystyle{apalike}
\bibliography{biblio} % replace by the name of your .bib file

\end{document}

%\begin{figure}[t]
%\begin{center}
%\includegraphics[width=2.1in,angle=-90]{fig1.eps}
%\caption{``Energies'' (inferiorities) of strings in a first-order
%  phase transition with latent heat $\Delta\epsilon$.}
%\label{fig1}
%\end{center}
%\end{figure}

%
%\begin{table}[h]
%\center{
%\begin{tabular}{|c|c|c|c|}\hline
%Name & Result & Bonus $b_i$ & Difficulty\\ \hline\hline
%Echo & I/O   & 1 & --\\
%Not  & $\neg A$ & 2 & 1 \\
%Nand & $\neg(A\wedge B)$ & 2 & 1 \\
%Not Or & $\neg A \vee B$ & 3 & 2 \\
%And  &  $ A \wedge B $   & 3 & 2 \\
%Or   &  $ A \vee B $     & 4 & 3 \\
%And Not & $A\wedge\neg B$& 4 & 3 \\
%Nor  & $\neg(A\vee B)$   & 5 & 4 \\
%Xor  & $ A\ {\rm xor}\ B$ &   6 & 4 \\
%Equals &$\neg(A\ {\rm xor}\ B)$&6& 4 \\ \hline
%\end{tabular}
%}
%\vskip 0.25cm
%\caption{Logical calculations on random inputs $A$ and $B$ rewarded,
%bonuses, and difficulty (in minimum number of {\tt nand} instructions
%required). Bonuses $b_i$ increase the speed of a CPU by a factor
%$\nu_i=1+2^{b_i-3}$.}
%\end{table}
%

### 2020-02-16 Docs/Papers/ALifeCities/paper.tex
init_corpus_constr

@@
 
-\title{Cities as they could be: an Artificial Life approach to Urban Systems}
+\title{Cities as they could be: Artificial Life and Urban Systems}
 \author{Juste Raimbault$^{1,2,3}$\\
@@
 \section{Introduction}
+
+\cite{batty2009centenary} % + relevant in other paper: morphogenesis, urban evolution?
 
@@
 % Q: a priori or a posteriori analysis?
+
+The keyword request follows \cite{niu2016global} and uses only \texttt{Artificial Intelligence AND Urban} for AI approaches and \texttt{Artificial Life AND Urban} for ALife.
+Using alternative terms such as ``machine learning'' or ``city'' expands the scope too much and yield non relevant results.
+% "smart city" ok should come by itself in the scope - less a priori as possible
+We construct initial seed corpuses using these requests
+% 20 limit already fine for ALife
+% -> add by hand refs? such as \cite{kato1998alife}
+% AI: start with 50 - seems relevant up to 100 at least but strong intersection with CA approaches.
+%
 

### 2020-02-27 Docs/Papers/ALifeCities/paper.tex
paper

@@
 $^2$UPS CNRS 3611 ISC-PIF\\
-$^3$UMR CNRS 8504 G{\'e}ographie-cit{\'e}s
+$^3$UMR CNRS 8504 G{\'e}ographie-cit{\'e}s\\
 juste.raimbault@polytechnique.edu} % email of corresponding author
@@
 
+% \cite{pattee1995artificial} epistemic cut: simulation / hard or wet ALife / nature of computation % not relevant here?
+
 \section{Introduction}
 
-\cite{batty2009centenary} % + relevant in other paper: morphogenesis, urban evolution?
+The understanding of processes driving the growth of cities, or more generally the evolution of urban systems, and approaches for a sustainable design and management of these, have for more than 100 years been tightly linked to concepts borrowed from biology such as evolution \citep{batty2009centenary}.
+Inventing novel ways to design cities, beyond predicting their evolution \citep{batty2018inventing}, is a major asset to tackle climate change and sustainability issues \citep{iepsfc2018science}.
+In that context, Artificial Life (ALife) approaches to urban systems have several advantages, including knowledge transfer from biology and ecology where relevant concepts such as resilience or morphogenesis are thoroughly studied, a strong practice of interdisciplinarity, or methodologies and tools such as agent-based modeling and cellular automata, among others.
+Following the seminal view of ``Life as it could be'' by \cite{langton1986studying}, an explicit ALife take on urban issues would consist in studying ``Cities as they could be''.
+Although not explicitly listed in application domains of ALife by \cite{kim2006comprehensive}, is is included through the use of methods and in relation with economic models.
+Two of archetypal ``would-be'' worlds of \cite{casti1997would} are related to urban systems (transportation and resource exploitation).
+\cite{doi:10.1162/isala00135} recently introduced a conceptual frame considering cities from an organismic perspective, while \cite{doi:10.1162/isala00134} showed that open-endedness has a strong potential to develop sustainable social systems.
 
-\cite{kim2006comprehensive} applications of alife
+Diverse streams of research on urban systems have already linked with ALife.
+From a methodological viewpoint, the use of cellular automata and agent-based models for urban growth and urban dynamics has a long history \cite{torrens2003automata}.
+These can be used for example to generate building layouts and road networks of synthetic cities \citep{kato1998alife} or at the scale of districts \citep{raimbault2019generating}.
+Integrating such models into evolutionary computation algorithms widens the scope of possible synthetic cities \citep{kato2000modeling}.
+Other dimensions such as land prices and residential dynamics can be grasped using agent-based models \citep{takizawa2000simulation}.
+Generative processes can also be used for interactive urban design \citep{openshaw1995developing}.
+The study of urban morphology may be done with methods used to study morphogenesis, and \cite{medda2009morphogenetic} apply reaction-diffusion equation to model the relation between transportation and land-use, while \cite{raimbault2018calibration} shows that the combination of aggregation and diffusion produces realistic urban forms.
+\cite{d2013simulating} explores possible future sustainable urban morphologies.
+\cite{lucic2002transportation} use bio-inspired algorithms to solve difficult transportation planning problems.
+The concept of urban metabolism introduced by \cite{olsen1982urban} also comes as a transfer from biology.
 
-\cite{pattee1995artificial} epistemic cut: simulation / hard or wet ALife / nature of computation
+Moreover, the field of Artificial Intelligence (AI) has also numerous application related to urban systems.
+\cite{wu2010artificial} review AI application to the prediction of land-use dynamics, unveiling a very broad range of methods ranging from evolutionary computation to neural networks, and suggesting that integrative and interdisciplinary approaches still lack for more robust urban applications.
+\cite{white1989artificial} show for example that a neural network trained appropriately can learn to plan transportation infrastructures.
+\cite{zheng2014urban} define the emerging field of Urban Computing as the convergence of ubiquitous urban data with artificial intelligence and new urban services, with varied domains of application including transportation, economy, environment, and planning.
+AI can for example be applied in real-time conditions to manage highway traffic \citep{ma2009real}.
+Other urban dimensions which require accurate predictions with a high spatio-temporal resolution, such as water demand \citep{adamowski2010comparison} or electric vehicles grid management \citep{rigas2014managing}, are other examples where AI is successfully applied.
 
-% urban DNA ?
+Thus, as both ALife and AI have been broadly applied to urban systems, we can first ask what are their respective extent in terms of methodologies, tools, concepts, and application domains, and secondly what are their remaining potentialities to enhance the understanding and management of cities, in other words what research directions and concepts in that particular context remain to be explored.
+This paper proposes to tackle these two questions by means of a systematic literature mapping method based on citation networks.
+More precisely, our contribution (i) constructs and explores a large citation network of around 250,000 papers, to map the respective contributions of ALife and AI to the urban literature, and their relations; and (ii) explores more thoroughly crucial concepts still loosely applied or understood in an urban system perspective.
 
-\cite{torrens2003automata} CA and ABM
-
-\cite{openshaw1995developing} GIS
-
-% broad issues: computing territorial systems ? cultural evol to urban evol ?
-
-\section{Artificial intelligence}
-
-\cite{wu2010artificial} AI applied to land-use dynamics
-
-\cite{white1989artificial}
-
-\section{Artificial life}
-
-\cite{medda2009morphogenetic} Turing equations
-
-\cite{olsen1982urban} urban metabolism and cycles
-
-\cite{kato1998alife} road network / building layout
-
-\cite{kato2000modeling} CA/GA dynamical
-
-% some procedural modeling ?
+The rest of this paper is organized as follows: we develop in the next section the bibliometric analysis based on citation networks to map the scope of artificial life and artificial intelligence approaches to urban systems.
+Building on this systematic mapping, we then review the principal points in which artificial life can significantly inform the study of urban systems.
+We finally discuss research directions opened by taking such a viewpoint of ``Cities as They Could Be''.
 
@@
 
-%%%%
-% Method
-% - keyword req (separate alife / AI ? - size pb) || initial corpus ?
-% - cit nw construction
+\subsection{Method}
+
 % Q: a priori or a posteriori analysis?
 
-The keyword request follows \cite{niu2016global} and uses only \texttt{Artificial Intelligence AND Urban} for AI approaches and \texttt{Artificial Life AND Urban} for ALife.
+Literature mapping and quantitative bibliometrics have been widely used to reinforce knowledge in most disciplines, and are part of a field of study in itself \cite{leydesdorff2001challenge}.
+They are furthermore important to enhance reflexivity which is crucial in disciplines studying socio-technical systems \cite{raimbault2019empowering}.
+In the case of artificial intelligence, several mappings have been proposed, for example from a semantic \citep{van1993neural}, spatialized \citep{niu2016global}, or journal-level \citep{ibanez2011using} viewpoint.
+\cite{squazzoni2013social} analyze the impact of the Journal on Artificial Societies and Social Simulation.
+\cite{aguilar2014past} show the evolution of theme frequency in time for the Artificial Life journal.
+There is however to the best of our knowledge no previous attempt of such an exercise for Artificial Life at a large scale.
+We propose here such a literature mapping approach to both ALife and AI, in the specific context of urban systems applications.
+
+We use therefore a citation network analysis, applying the methods and tools developed by \cite{raimbault2019exploration}.
+In a nutshell, citation networks are constructed by first constituting a seed corpus using a keyword search, and then by collecting papers citing papers in this corpus, recursively to a certain level.
+
+The keyword we use to constitute our seed corpus follows , and we thus consider only the requests \texttt{Artificial Intelligence AND Urban} for AI approaches and \texttt{Artificial Life AND Urban} for ALife.
 Using alternative terms such as ``machine learning'' or ``city'' expands the scope too much and yield non relevant results.
@@
 We construct initial seed corpuses using these requests
-% 20 limit already fine for ALife
+% 20 limit already fine for ALife - NO
 % -> add by hand refs? such as \cite{kato1998alife}
 % AI: start with 50 - seems relevant up to 100 at least but strong intersection with CA approaches.
-%
+% Initial: AI 200; ALife 50.
+Some references which obviously did not fit the scope were manually removed. 
+%sociology / gender studies: WTF?
 
-\section{Cities as they could be}
+\subsection{Results}
 
-% How ALife concepts and approaches can help understanding / managing / planning / designing urban systems
+% - interesting insight: on which communities do AI and ALife intersect (hypothesis: ABM, evolutionary computation?)
 
-% - biomimetism
+\section{Cities as they could be: strengthening ALife concepts in urban science}
+
+We turn now to a more thorough development of some concepts related to ALife which should have either a high theoretical importance for the study of urban systems, or a high potentiality to introduce novel approaches.
+This list stems from the conjunction of open issues in urban science \citep{lobo2020urban} with conclusions from the previous literature mapping on underexplored paths.
+It can be understood as main arguments of why ALife concepts may help understanding, planning, designing and managing in a better and more sustainable way.
+
+% Note: wu2010artificial alife subset of AI? ~ contrary in theory.
+% Insist here that most approaches are "tools"/technical: what is rare and could really make a difference are strong interdisciplinary works, concepts transfers (with proper defs / shared etc.)
+% ! if it is true from the literature review
+
+% - process understanding: "the real smart"
+
+\subsection{Tools and methods}
+
+% - tools: simulation models etc.: already quite mature: which perspectives?
+
+\subsection{Biomimetism}
+
+% - biomimetism (biomimicry?)
+
+\cite{taylor2017art}
+
+\subsection{Urban ecology}
+
+% - urban metabolism
+
+\subsection{Morphogenesis and autopoiesis}
+
 % - morphogenesis / autopoiesis
-% - tools: simulation models etc.
-% - urban metabolism
-% - process understanding: "the real smart"
-% - innovation diffusion ? (! not too close from urb evol th: step back? )
+% form and function: fundamental question. morphogenetic engineering
+% autopoiesis: the question of boundaries; sustainibility.
+
+\subsection{Urban evolutionary theories}
+
+% - proper defs of ``Urban dna'', evolutionary processes
+
+The definition of an ``Urban DNA'', i.e.
+an extension of the gene concept in evolution or of the \emph{meme} concept in cultural evolution, or even an approach combining different types of replicators \cite{bull2000meme}, remains an open question.
+
+\subsection{Innovation in urban systems}
+
+% - innovation diffusion
+
+\subsection{Linking Urban ALife and AI: urban computing}
+
+% broad issues: computing territorial systems ? cultural evol to urban evol ?
 
@@
 %  - proper definition of urban genotype / phenotype (higher dimension of cultural evol. ?)
-%  - def and understanding of co-evolution
+%  - def and understanding of co-evolution -> with open ended, in previous section?
 %. - understanding of morphogenesis - in the sense of urban form and function

### 2020-02-27 Docs/Papers/EvolutionInnovation/paper.tex
paper

--- 2019-12-12 Docs/Papers/ALifeCities/paper.tex
@@
 
-\title{Cities as they could be: an Artificial Life approach to Urban Systems}
+\title{A simple model of urban evolution based on innovation diffusion}
 \author{Juste Raimbault$^{1,2,3}$\\
@@
 
-\cite{kim2006comprehensive} applications of alife
+% https://scholar.google.com/scholar?hl=fr&as_sdt=0%2C5&q=meme+gene+coevolution&btnG=
 
-\cite{pattee1995artificial} epistemic cut: simulation / hard or wet ALife / nature of computation
+\cite{votsis2019urban}
 
-% urban DNA ?
+\section{Urban evolution model}
 
-\cite{torrens2003automata} CA and ABM
+\subsection{Rationale}
 
-\cite{openshaw1995developing} GIS
+The principal idea in the model is to include a concept of ``Urban DNA'', i.e.
+a kind of genome that cities would be exchanging and which would undergo mutation processes.
+A suitable candidate is to build on the concept of \emph{meme} introduced in the field of cultural evolution.
+However, several particularities must be stressed out when working with urban systems.
 
-% broad issues: computing territorial systems ? cultural evol to urban evol ?
+\cite{d2014urban}
 
-\section{Artificial intelligence}
+\subsection{Model description}
 
-\cite{wu2010artificial} AI applied to land-use dynamics
+\cite{favaro2011gibrat}
 
-\cite{white1989artificial}
-
-\section{Artificial life}
-
-\cite{medda2009morphogenetic} Turing equations
-
-\cite{olsen1982urban} urban metabolism and cycles
-
-\cite{kato1998alife} road network / building layout
-
-\cite{kato2000modeling} CA/GA dynamical
-
-% some procedural modeling ?
-
-\section{Bibliometric analysis}
-
-%%%%
-% Method
-% - keyword req (separate alife / AI ? - size pb) || initial corpus ?
-% - cit nw construction
-% Q: a priori or a posteriori analysis?
-
-\section{Cities as they could be}
-
-% How ALife concepts and approaches can help understanding / managing / planning / designing urban systems
-
-% - biomimetism
-% - morphogenesis / autopoiesis
-% - tools: simulation models etc.
-% - urban metabolism
-% - process understanding: "the real smart"
-% - innovation diffusion ? (! not too close from urb evol th: step back? )
+\section{Results}
 
 \section{Discussion}
-
-% what remains to be done
-% / open problems in "urban evolution"
-%
-%  - proper definition of urban genotype / phenotype (higher dimension of cultural evol. ?)
-%  - def and understanding of co-evolution
-%. - understanding of morphogenesis - in the sense of urban form and function
-%. - understanding of autopoiesis: urban boundaries is an open Q (<- use open Q as arg why analogy is relevant ?)
 

### 2020-04-22 Docs/Papers/ALifeCities/paper/paper.tex
paper

\documentclass[letterpaper]{article}

\usepackage{natbib,alifeconf}  %% The order is important

% *****************
%  Requirements:
% *****************
%
% - All pages sized consistently at 8.5 x 11 inches (US letter size).
% - PDF length <= 8 pages for full papers, <=2 pages for extended
%    abstracts (not including citations).
% - Abstract length <= 250 words.
% - No visible crop marks.
% - Images at no greater than 300 dpi, scaled at 100%.
% - Embedded open type fonts only.
% - All layers flattened.
% - No attachments.
% - All desired links active in the files.

% Note that the PDF file must not exceed 5 MB if it is to be indexed
% by Google Scholar. Additional information about Google Scholar
% can be found here:
% http://www.google.com/intl/en/scholar/inclusion.html.

% If your system does not generate letter format documents by default,
% you can use the following workflow:
% latex example
% bibtex example
% latex example ; latex example
% dvips -o example.ps -t letterSize example.dvi
% ps2pdf example.ps example.pdf

% For pdflatex users:
% The alifeconf style file loads the "graphicx" package, and
% this may lead some users of pdflatex to experience problems.
% These can be fixed by editing the alifeconf.sty file to specify:
% \usepackage[pdftex]{graphicx}
%   instead of
% \usepackage{graphicx}.
% The PDF output generated by pdflatex should match the required
% specifications and obviously the dvips and ps2pdf steps become
% unnecessary.

% Note:  Some laser printers have a serious problem printing TeX
% output. The use of ps type I fonts should avoid this problem.

\usepackage{mdframed}
\usepackage{hyperref}
\usepackage{multirow}
\usepackage{changepage}

\title{\vspace{-2cm}Cities as they could be: Artificial Life and Urban Systems}
\author{Juste Raimbault$^{1,2,3}$\\
\mbox{}\\
$^1$Center for Advanced Spatial Analysis, University College London\\
$^2$UPS CNRS 3611 ISC-PIF\\
$^3$UMR CNRS 8504 G{\'e}ographie-cit{\'e}s\\
juste.raimbault@polytechnique.edu} % email of corresponding author

% For several authors from the same institution use the same number to
% refer to one address.
%
% If the names do not fit well on one line use
%         Author 1, Author 2 ... \\ {\Large\bf Author n} ...\\ ...
%
% If the title and author information do not fit in the area
% allocated, place \setlength\titlebox{<new height>} after the
% \documentclass line where <new height> is 2.25in

\begin{document}
\maketitle

\begin{abstract}
% Abstract length should not exceed 250 words
The metaphor of cities as organisms has a long history in urban planning, and a few urban modeling approaches have explicitly been linked to Artificial Life. We propose in that paper to explore the extent of Artificial Life and Artificial Intelligence application to urban issues, by constructing and exploring a citation network of around 225,000 papers. It shows that most of the literature is indeed application of methodologies and a rather strong modularity of approaches. We finally develop ALife concepts which have a strong potential for the development of new urban theories.
\end{abstract}

% \cite{pattee1995artificial} epistemic cut: simulation / hard or wet ALife / nature of computation % not relevant here?

\cite{doi:10.1098/rstb.2019.0329}

\section{Introduction}

The understanding of processes driving the growth of cities, or more generally the evolution of urban systems, and approaches for a sustainable design and management of these, have for more than 100 years been tightly linked to concepts borrowed from biology such as evolution \citep{batty2009centenary}. Inventing novel ways to design cities, beyond predicting their evolution \citep{batty2018inventing}, is a major asset to tackle climate change and sustainability issues \citep{iepsfc2018science}. In that context, Artificial Life (ALife) approaches to urban systems have several advantages, including knowledge transfer from biology and ecology where relevant concepts such as resilience or morphogenesis are thoroughly studied, a strong practice of interdisciplinarity, or methodologies and tools such as agent-based modeling and cellular automata, among others. Following the seminal view of ``Life as it could be'' by \cite{langton1986studying}, an explicit ALife take on urban issues would consist in studying ``Cities as they could be''. Although not explicitly listed in application domains of ALife by \cite{kim2006comprehensive}, is is included through the use of methods and in relation with economic models. Two of archetypal ``would-be'' worlds of \cite{casti1997would} are related to urban systems (transportation and resource exploitation). \cite{doi:10.1162/isala00135} recently introduced a conceptual frame considering cities from an organismic perspective, while \cite{doi:10.1162/isala00134} showed that open-endedness has a strong potential to develop sustainable social systems.

Diverse streams of research on urban systems have already linked with ALife. From a methodological viewpoint, the use of cellular automata and agent-based models for urban growth and urban dynamics has a long history \cite{torrens2003automata}. These can be used for example to generate building layouts and road networks of synthetic cities \citep{kato1998alife} or at the scale of districts \citep{raimbault2019generating}. Integrating such models into evolutionary computation algorithms widens the scope of possible synthetic cities \citep{kato2000modeling}. Other dimensions such as land prices and residential dynamics can be grasped using agent-based models \citep{takizawa2000simulation}. Generative processes can also be used for interactive urban design \citep{openshaw1995developing}. The study of urban morphology may be done with methods used to study morphogenesis, and \cite{medda2009morphogenetic} apply reaction-diffusion equation to model the relation between transportation and land-use, while \cite{raimbault2018calibration} shows that the combination of aggregation and diffusion produces realistic urban forms. \cite{d2013simulating} explores possible future sustainable urban morphologies. \cite{lucic2002transportation} use bio-inspired algorithms to solve difficult transportation planning problems. The concept of urban metabolism introduced by \cite{olsen1982urban} also comes as a transfer from biology.

Moreover, the field of Artificial Intelligence (AI) has also numerous application related to urban systems. \cite{wu2010artificial} review AI application to the prediction of land-use dynamics, unveiling a very broad range of methods ranging from evolutionary computation to neural networks, and suggesting that integrative and interdisciplinary approaches still lack for more robust urban applications. \cite{white1989artificial} show for example that a neural network trained appropriately can learn to plan transportation infrastructures. \cite{zheng2014urban} define the emerging field of Urban Computing as the convergence of ubiquitous urban data with artificial intelligence and new urban services, with varied domains of application including transportation, economy, environment, and planning. AI can for example be applied in real-time conditions to manage highway traffic \citep{ma2009real}. Other urban dimensions which require accurate predictions with a high spatio-temporal resolution, such as water demand \citep{adamowski2010comparison} or electric vehicles grid management \citep{rigas2014managing}, are other examples where AI is successfully applied.

Thus, as both ALife and AI have been broadly applied to urban systems, we can first ask what are their respective extent in terms of methodologies, tools, concepts, and application domains, and secondly what are their remaining potentialities to enhance the understanding and management of cities, in other words what research directions and concepts in that particular context remain to be explored. This paper proposes to tackle these two questions by means of a systematic literature mapping method based on citation networks. More precisely, our contribution (i) constructs and explores a large citation network of around 225,000 papers, to map the respective contributions of ALife and AI to the urban literature, and their relations; and (ii) explores more thoroughly crucial concepts still loosely applied or understood in an urban system perspective.

The rest of this paper is organized as follows: we develop in the next section the bibliometric analysis based on citation networks to map the scope of artificial life and artificial intelligence approaches to urban systems. Building on this systematic mapping, we then review principal points in which artificial life can significantly inform the study of urban systems, which can be understood as research directions opened by taking such a viewpoint of ``Cities as They Could Be''.

\section{Bibliometric analysis}

%\subsection{Method}

% Q: a priori or a posteriori analysis?

Literature mapping and quantitative bibliometrics have been widely used to reinforce knowledge in most disciplines, and are part of a field of study in itself \cite{leydesdorff2001challenge}. They are furthermore important to enhance reflexivity which is crucial in disciplines studying socio-technical systems \cite{raimbault2019empowering}. In the case of artificial intelligence, several mappings have been proposed, for example from a semantic \citep{van1993neural}, spatialized \citep{niu2016global}, or journal-level \citep{ibanez2011using} viewpoint. \cite{squazzoni2013social} analyze the impact of the Journal on Artificial Societies and Social Simulation. \cite{aguilar2014past} show the evolution of theme frequency in time for the Artificial Life journal. There is however to the best of our knowledge no previous attempt of such an exercise for Artificial Life at a large scale. We propose here such a literature mapping approach to both ALife and AI, in the specific context of urban systems applications.

We use therefore a citation network analysis, applying the methods and tools developed by \cite{raimbault2019exploration}. In a nutshell, citation networks are constructed by first constituting a seed corpus using a keyword search, and then by collecting papers citing papers in this corpus, recursively to a certain level. This method is an effective tool to reconstruct endogenous disciplines of citation practice. It is implemented into an open source library collecting and processing data from Google scholar, available at \url{https://github.com/JusteRaimbault/BiblioData}.

% 20 limit already fine for ALife - NO
% -> add by hand refs? such as \cite{kato1998alife} NO
% AI: start with 50 - seems relevant up to 100 at least but strong intersection with CA approaches. NO - 200
% Initial: AI 200; ALife 50.

The keyword we consider for constructing the initial corpus are \texttt{Artificial Intelligence AND Urban} for AI approaches and \texttt{Artificial Life AND Urban} for ALife. Using alternative terms such as ``machine learning'' or ``city'' expands the scope too much and yield non relevant results. A size of 200 papers for each request appears to cover most of relevant literature, the higher rank query results being more and more unrelated. The citation propagation will in any case retrieve possibly missing papers. 25 references in the seed corpus which obviously did not fit the scope were manually removed (sociology or gender studies e.g.), what yields a corpus of 296 papers (due to the overlap between the two requests). From this layer, the backward citation network is retrieved up to depth two. This ensures in particular that all links within the depth one layer are effectively observed.

%\subsection{Results}

Data collection and preprocessing is done in Java using the BiblioData library, and network analysis is done in R with the igraph package. Network visualization is done with the Gephi software. All raw and processed data is available on dataverse at \url{https://doi.org/10.7910/DVN/RXBJ7R}, and source code and results are available on a git repository at \url{https://github.com/JusteRaimbault/UrbanEvolution}.

%%
% full graph after filtering
% IGRAPH 1e32631 DN-- 224510 315829 --

%  components(citation)$csize
% [1] 223476     18    664     60     12     26    114      4      6      6      6     19      4      9     10
%[16]      9      6      5      3      8      6     11      2      1      4      1      5      2      3      5
%[31]      5

% largest component IGRAPH 15a0007 DN-- 223476 314750

% citation core higher citationcorehigher
%IGRAPH 15c490e DN-- 48657 139931

The raw citation network contains $\left|V\right| = 224,510$ papers and $\left|E\right| = 315,829$ citation links. We keep the largest connected component (size $223,476$ against the second of size only $664$), and furthermore extract the core of the network by removing all nodes of degree 1, and repeating this until all nodes have at least a degree of two. We then work with this core network with $48,657$ nodes and $139,931$ edges.

% largest communities
% [1] "31 ; Real-time mapping ; 13.4759644038884"
%[1] "24 ; Autonomous vehicles ; 9.06138890601558"
%[1] "10 ; Traffic forecasting ; 7.11305670304376"
%[1] "8 ; 3D Modeling ; 6.51499270403025"
%[1] "26 ; Complexity/ABM ; 5.34352713895226"
%[1] "23 ; Computer vision ; 4.99619787492036"
%[1] "15 ; Traffic control ; 4.87494091292106"
%[1] "14 ; Urban growth/CA ; 4.87083050742956"
%[1] "22 ; Rescue robots ; 4.86672010193806"
%[1] "20 ; Art/Culture ; 4.28715292763631"

% modularity
% directedmodularity(com$membership,A)
%  0.8411035

%summary(lm(data=data.frame(y=log(sort(comsizes[comsizes>500],decreasing = T)),x=log(1:length(comsizes[comsizes>500]))),y~x))
% x           -0.73953

% hand grouping as themes
% Autonomous vehicles
%. 'Autonomous vehicles', 'Real-time mapping', 'Flying UAV', 'Image capture', 'Power line detection', 'Lidar mapping', 'Computer vision'
% Robots
% 'Rescue robots'
% Environmental science 
% 'Natural hazards', 'Flooding', 'Rainfall', 'Air pollution', 'Biology'
% Traffic and transportation
% 'Traffic control', Traffic forecasting', 'Traffic safety', 'Transit planning', 'Bio-inspired computing'
% Urban growth
% 'Urban growth/CA', 'Remote sensing', 'GIS'
% Social systems
%  'Social intelligence', 'Game theory', 'Fuzzy cognitive maps', 'Crowd simulation'
% Complexity
%   'Complexity/ABM', 'Self-organization'

% Architecture
% 'Living architecture', 'Urbanism', '3D Modeling'
% Humanities 
% 'Art/Culture', 'Smart cities'

\begin{table*}[h!]
\begin{adjustwidth}{-1cm}{-1cm}
\center{
\begin{tabular}{|c|c|c|c|c|c|}\hline
\multicolumn{3}{|c|}{Full network} & \multicolumn{3}{|c|}{ALife subnetwork}\\\hline
Community & Size (\%) & Repr. paper & Community & Size (\%) & Repr. paper\\ \hline\hline
Real-time mapping & 13.5 & \citep{thrun2006graph} & Complexity & 12.7 & \citep{casti1997would} \\
Autonomous vehicles & 9.1  & \citep{levinson2011towards} & Flying UAV & 11.0 & \citep{kendoul2012survey} \\
Traffic forecasting & 7.1  & \citep{vlahogianni2005optimized} & Cellular automata & 9.8 & \citep{white1993cellular} \\
3D Modeling & 6.5  & \citep{akbarzadeh2006towards} & Autonomous vehicles & 9.1 & \citep{gonzalez2015review} \\
Complexity/ABM & 5.3  & \citep{casti1997would} & Culture & 8.6 & \citep{franklin2002embodied} \\
Computer vision & 5.0  & \citep{sen2004robust} & Crowd behavior & 7.2 & \citep{pelechano2007controlling} \\
Traffic control & 4.9  & \citep{chen2010review} & Traffic control & 6.5 & \citep{chen2010review} \\
Urban growth/CA & 4.9  & \citep{batty1994cells} & Traffic forecasting & 4.5 & \citep{lv2014traffic} \\
Rescue robots & 4.8 & \citep{davids2002urban} & Social intelligence & 4.1 & \citep{wang2007toward} \\
Art/Culture & 4.3 & \citep{whitelaw2004metacreation} & Urbanism & 4.0 & \citep{schipperijn2010factors} \\\hline
\end{tabular}
}
\end{adjustwidth}
\caption{Largest citation communities, in the full core network ($\left|V\right| = 48,657$) and in the ALife-induced core network ($\left|V\right|=17,705$). We give the relative share of each community and one representative paper, among the ones with the largest degree.\label{tab:coms}}
\end{table*}

We use the Louvain algorithm for community detection \citep{blondel2008fast} on the undirected equivalent network to reconstruct endogenous scientific fields. We obtain 33 communities, with a directed modularity \citep{nicosia2009extending} of 0.84, what corresponds to highly clustered communities. Their sizes are relatively balanced, since the size distribution with a cutoff of 500 papers has a Zipf exponent (OLS estimation) of 0.73. We associate a ``field'' to each community by inspecting the largest degree papers in each. We describe the ten largest community in Table~\ref{tab:coms} with their label, relative size, and representative paper. The content is broad in terms of disciplines involved, methods used, and applications domains. More precisely, they can be grouped into the following main themes:

\begin{itemize}
	\item Autonomous vehicles, i.e technologies developed for self-driving vehicles in urban environments (Autonomous vehicles, Real-time mapping, Flying UAV, Image capture, Power line detection, Lidar mapping, Computer vision);
	\item Robots: use rescue robots in urban disasters;
	\item Environmental science: application of mostly AI methods to the estimation of environmental variables or natural hazards (Natural hazards, Flooding, Rainfall, Air pollution, Biology);
	\item Traffic and transportation, in terms of management and planning (Traffic control, Traffic forecasting, Traffic safety, Transit planning, Bio-inspired computing applied to transit);
	\item Urban growth: analysis and modeling of urban growth (Urban growth/Cellular automata, Remote sensing, GIS);
	\item Social systems: simulation of artificial societies and agent-based modeling (Social intelligence, Game theory, Fuzzy cognitive maps, Crowd simulation, Smart cities);
	\item Complexity: concepts and methods linked to complexity, mostly from ALife (Complexity/Agent-based Modeling, Self-organization)
	\item Architecture: bio-inspired and computational architecture and urbanism (Living architecture, Urbanism, 3D Modeling);
	\item Humanities: art and cultural studies such as generative art and studies of digital culture (Art/Culture). The emergence of this community is interesting in itself, as it shows the importance of artistic approaches in the ALife landscape, actually witnessed in recent ALife conferences with the Art track, and more generally by the crucial role of epistemology and philosophy in complexity science.
\end{itemize}

%%%%%%%%%%
\begin{figure*}
\begin{center}
\begin{mdframed}
\includegraphics[width=\linewidth]{figures/core.png}
\end{mdframed}
\begin{mdframed}
\includegraphics[width=\linewidth]{figures/corealife_zoom.png}
\end{mdframed}
\caption{Citation networks for Artificial Life and Artificial Intelligence applications to urban systems. Full core citation network (top panel) and ALife subnetwork (bottom panel). Network visualization is done with the Gephi software, spatialization with a Force Atlas 2 algorithm, node label size corresponds to node degree. Colors correspond to communities obtained with the Louvain algorithm.\label{fig:citnw}}
\end{center}
\end{figure*}
%%%%%%%%%%

Note that these main themes do not necessarily correspond to the actual organization of communities and their interactions. To grasp these in a first order simple manner, we can inspect the network visualization shown in Fig.~\ref{fig:citnw} (top panel). In the network spatialization, we observe a big cluster of communities formed by studies on autonomous vehicles and robotics (top left of the network). As real-time mapping for such technologies is closely related to 3D reconstruction used in architecture, this last community is far from other architecture communities which are closer to ALife approaches on the bottom. An other large and relatively isolated cluster corresponds to the transportation theme (top right cluster), while more sparse communities, mostly related to ALife (complexity, urban growth, artificial societies, art and culture), gather on the bottom of the network. These include Urban growth modeling, Remote sensing and GIS, but also the complexity and Art/Culture community. Finally, Environmental science communities, such as pollution or landslides, are mostly quite isolated (e.g. the two on the most right) as they are highly specific to the application. The ``bridge'' between ALife approaches on the bottom of the network and the approaches more dominated by AI on the top (transportation and autonomous vehicles) is actually done by communities related to social systems: Crowd simulation, Smart cities and game theories. In a sense, ALife focuses more on generative processes for the built environment, while AI on technical issues related to urban technologies, what makes the ``human-in-the-middle'' a logical mediator between the two. Indeed, urban systems are complex socio-technical systems, in which human agents interact between themselves, and multiple social and technical layers.

%%%%%%%%%%
\begin{figure*}
\vspace{-1cm}
\begin{center}
\includegraphics[width=0.49\linewidth]{figures/citation_all.png}
\includegraphics[width=0.49\linewidth]{figures/citation_alife.png}
\caption{Citation flows between largest communities, for the full network (Left) and the ALife subnetwork (Right). Scale color is given by a renormalized logarithmic scale (the major share of citations being internal to communities), the maximal flow being 0.95 (resp. 0.98) for the full network (resp. the ALife subnetwork) and the minimal flow $4.8\cdot 10^{-5}$ (resp. $1.9 \cdot 10^{-4}$). Dendograms show a hierarchical clustering following the proximity matrix.\label{fig:flows}}
\end{center}
\end{figure*}
%%%%%%%%%%

To confirm which are the communities typical to ALife but not AI approaches, we study the subnetwork originating from the ALife keyword request only. In terms of size, the core of this network has only $\left|V\right|=17,705$ papers, which is a bit more than a third of the full network. Artificial Intelligence is thus larger than ALife for paper quantity in urban applications, what is consistent with the respective size of their literature and communities. We also give the 10 largest ALife citation communities in Table~\ref{tab:coms}, and find quite a high overlap in term of content. Not surprisingly, the largest is the conceptual community related to complexity and agent-based modeling, dominated by \cite{casti1997would} which surveyed two typical artificial societies. Autonomous vehicles and robotics are still present, what is consistent with the ``hard'' branch of ALife linked to self-assembly and self-organizing robots. When inspecting the spatialized subnetwork (Fig.~\ref{fig:citnw}, bottom panel), we find that the Complexity community is central, and that social modeling and transportation are this time making the bridge with autonomous vehicles, which are weakly linked to the rest of the network. The Remote Sensing community has disappeared, what is consistent with the fact that this aspect of urban studies corresponds mostly to machine learning for unsupervised learning. Similarly, Smart cities do not appear in the ALife subnetwork: indeed, most of these approaches do not deal with slow time scales nor with generative processes, being in a sense ``superficial'' since they do not capture underlying processes governing the evolution of urban systems, but aim at exploiting a new technical layer (consisting of sensors e.g.) put on top of existing urban structures.

% ! should also map AI only communities to confirm which one are alife only.

%%% Alife communities

% corealife
%IGRAPH 57fd0b0 DN-- 17705 36497

% directedmodularity(comalife$membership,A)
%[1] 0.8645957

% "20 ; Complexity ; 12.7478113527252"
%[1] "7 ; Flying UAV ; 10.9743010449026"
%[1] "15 ; Cellular automata ; 9.80513979101949"
%[1] "30 ; Autonomous vehicles ; 9.10477266308952"
%[1] "26 ; Culture ; 8.64162665913584"
%[1] "12 ; Crowd behavior ; 7.15617057328438"
%[1] "42 ; Traffic control ; 6.5009884213499"
%[1] "14 ; Traffic forecasting ; 4.45636825755436"
%[1] "38 ; Social intelligence ; 4.10053657158995"
%[1] "24 ; Urbanism ; 3.99887037560011"
%[1] "10 ; Rescue robots ; 3.66563117763344"
%[1] "4 ; Art ; 3.46229878565377"

Finally, we can quantify the relation between the citation communities in terms of citation flows. More formally, if $c_{ij}$ is the citation network adjacency matrix, the directed flow from community $k_1$ to community $k_2$ is the share of citation made by $k_1$ which go to the other community, what reads $\phi_{i \rightarrow j} = \frac{\sum_{i\in k_1,j\in k_2} c_{ij}}{\sum_{i\in k_1,j} c_{ij}}$. These flows are shown as heatmaps in Fig.~\ref{fig:flows}. We observe consistent clustering of communities, following previous results from network spatialization and main themes: for example, Autonomous vehicles, 3D Modeling, Computer vision and Real-time mapping, form a strong cluster in the full network, as traffic control and traffic forecasting do. Note that flows are not symmetric: for example, Art/culture cite a small proportion of literature related to traffic, but traffic is totally blind to it. A similar asymmetry is observed between rescue robots citing traffic without reciprocity. We find that the ALife subnetwork is more modular, with less flows between far communities, but still a progressive shift from humanities (Urbanism) to technical disciplines (Autonomous vehicles), with successive sub-blocks on the diagonal corresponding to clusters of communities. In a nutshell, we confirm the relations unveiled previously, and that domains remain isolated in absolute terms (we recall that heatmap have a logarithmic scale).

From this systematic literature mapping and exploration, we can conclude the following main stylized facts: (i) AI and ALife have overlapping applications for urban systems, but still remain complementary altogether; (ii) the contributions are mostly methodological and technical, since the only community explicitly including conceptual contributions (Complexity) represents only 5.3\% of the network. Similarly, approaches dealing with generative urban processes on slow time scales, typically urban morphogenesis, which dominated our introducing literature review which was driven by ALife concepts, is a relatively restreint part of the full corpus (around 10\% by including CA and ABM).

%%%

% interesting to look at: quantif of non-reciprocity

% - interesting insight: on which communities do AI and ALife intersect (hypothesis: ABM, evolutionary computation?)
%  -- later -- or not enough space?

\section{Discussion: Cities as they could be, strengthening ALife concepts in urban science}

We turn now to a more thorough development of some concepts related to ALife which should have either a high theoretical importance for the study of urban systems, or a high potentiality to introduce novel approaches. This list stems from the conjunction of open issues in urban science \citep{lobo2020urban} with conclusions from the previous literature mapping on underexplored paths. It can be understood as main arguments of why ALife concepts may help understanding, planning, designing and managing in a better and more sustainable way.

% Note: wu2010artificial alife subset of AI? ~ contrary in theory.
% Insist here that most approaches are "tools"/technical: what is rare and could really make a difference are strong interdisciplinary works, concepts transfers (with proper defs / shared etc.)
% ! if it is true from the literature review

% - process understanding: "the real smart"

\subsection{Tools and methods}

% - tools: simulation models etc.: already quite mature: which perspectives?

We saw that tools and methods, in particular Artificial Intelligence methods, are already widely applied. However, new methods from Artificial Life, such as modeling frameworks which would be appropriate to simulate open-ended evolution \citep{banzhaf2016defining}, could have an important role in exploring potential urban futures by means of simulation.

\subsection{Biomimicry and urban ecology}

% - biomimetism (biomimicry?)
% - urban metabolism

\cite{taylor2017art} has put forward biomimicry as a powerful approach for an innovative urban design, indeed in a way similar to evolutionary computation and bio-inspired computing or engineering following natural design principles. Building on the broad and old literature on metaphors of the city as an organism, these approaches remain mostly to be explored and tested, preferentially using simulation models, and transferring existing knowledge from ALife for this. Urban metabolism enters this frame, and \cite{raimbault2020spatial} for example systematically explore an agent-based model for industrial symbiosis.

\subsection{Morphogenesis and autopoiesis}

% - morphogenesis / autopoiesis
% form and function: fundamental question. morphogenetic engineering
% autopoiesis: the question of boundaries; sustainibility.
% %. - understanding of morphogenesis - in the sense of urban form and function
%. - understanding of autopoiesis: urban boundaries is an open Q (<- use open Q as arg why analogy is relevant ?)

The link between urban form and urban function remains an open fundamental question in the study of urban systems. The field of morphogenetic engineering provides a framework to engineer this relation from the bottom-up \citep{doursat2012morphogenetic}, which fits particularly well issues in urban planning and design. How it can be concretely applied is however still unknown. Furthermore, beyond the concept of morphogenesis and urban morphogenesis models, the concept of autopoiesis could play a crucial role in the study of urban systems. An autopoietic system can be understood as a network of processes with well-defined boundaries and which self-sustains itself \citep{bourgine2004autopoiesis}. The issue of urban systems boundaries is crucial in the definition of urban systems (see e.g. \citep{cottineau2017diverse} in the case of scaling laws), and investigating the autopoietic nature of cities should be a promising research direction.

\subsection{Urban evolutionary theories}

% - proper defs of ``Urban dna'', evolutionary processes
%  - proper definition of urban genotype / phenotype (higher dimension of cultural evol. ?)
%  - def and understanding of co-evolution -> with open ended, in previous section?

The definition of an ``Urban DNA'', i.e. an extension of the gene concept in evolution or of the \emph{meme} concept in cultural evolution, or even an approach combining different types of replicators \citep{bull2000meme}, remains an open question. Previous attempts such as \citep{votsis2019urban} are not satisfying since they rely more on the extraction of phenotypical parameters based on urban form. Fundamental processes to effectively have evolution (such as transmission, exchanges and mutation) work well in the case of cultural evolution, but urban systems have also slow time scale artefacts and infrastructures which may follow other types of rules. Defining and understanding co-evolution processes in urban systems is also an open research avenue strongly related to ALife. In that context, \cite{raimbault2019modeling} proposed for example a co-evolution approach to the particular case of modeling interactions between transportation networks and territories.

\subsection{Innovation in urban systems}

% - innovation diffusion

In relation to the previous point, \cite{pumain2018evolutionary} proposes an urban evolutionary theory based on the diffusion of innovation between cities. It is evolutionary in the sense that cities are complex adaptive systems of systems. Defining innovation and characterizing it in artificial systems is one cornerstone of ALife research, in particular through studies of open-ended evolution. In that sense, as innovation is indeed central to urban systems (technological, but also social and cultural), a better understanding of it is needed.

\subsection{Linking Urban ALife and AI: urban computing}

% broad issues: computing territorial systems ? cultural evol to urban evol ?

Finally, a last point possibly joining AI and ALife potentialities for urban systems, is the concept of urban computing. As currently defined, it relates more to AI approaches. However, in the same way that biological systems can compute, the notion of social and collective computation \citep{flack2011challenges} relates to social intelligence and collective decision-making. To what extent are these processes modified when embedded into territorial systems, in particular through the geographical constraints, and to what extent new hybrid systems integrating sensors and crowd-sourced data can actually compute and make cities ``smarter'', i.e. in a sense how to achieve territorial intelligence, has still to be investigated.

%\section{Discussion} % -> not needed
% what remains to be done
% / open problems in "urban evolution"
%

\section{Conclusion}

We have developed a literature mapping of AI and ALife approaches to urban systems. It appears that most application remain methodological and technical, and that many conceptual research avenues stemming from Artificial Life remain to be developed as new theories of urban systems. We believe these have a strong potential for sustainable urban planning and management.

%\section{Acknowledgements}

\footnotesize
\bibliographystyle{apalike}
\bibliography{biblio} % replace by the name of your .bib file

\end{document}

%\begin{figure}[t]
%\begin{center}
%\includegraphics[width=2.1in,angle=-90]{fig1.eps}
%\caption{``Energies'' (inferiorities) of strings in a first-order
%  phase transition with latent heat $\Delta\epsilon$.}
%\label{fig1}
%\end{center}
%\end{figure}

%
%\begin{table}[h]
%\center{
%\begin{tabular}{|c|c|c|c|}\hline
%Name & Result & Bonus $b_i$ & Difficulty\\ \hline\hline
%Echo & I/O   & 1 & --\\
%Not  & $\neg A$ & 2 & 1 \\
%Nand & $\neg(A\wedge B)$ & 2 & 1 \\
%Not Or & $\neg A \vee B$ & 3 & 2 \\
%And  &  $ A \wedge B $   & 3 & 2 \\
%Or   &  $ A \vee B $     & 4 & 3 \\
%And Not & $A\wedge\neg B$& 4 & 3 \\
%Nor  & $\neg(A\vee B)$   & 5 & 4 \\
%Xor  & $ A\ {\rm xor}\ B$ &   6 & 4 \\
%Equals &$\neg(A\ {\rm xor}\ B)$&6& 4 \\ \hline
%\end{tabular}
%}
%\vskip 0.25cm
%\caption{Logical calculations on random inputs $A$ and $B$ rewarded,
%bonuses, and difficulty (in minimum number of {\tt nand} instructions
%required). Bonuses $b_i$ increase the speed of a CPU by a factor
%$\nu_i=1+2^{b_i-3}$.}
%\end{table}
%

### 2020-04-22 Docs/Papers/EvolutionInnovation/paper.tex
paper

@@
 
+% (i) interest to understand urban dynamics, link to ALife
+
+Urban systems 
+\cite{batty2009centenary}
+
+% (ii) Urban Evolution / Cultural evolution
+%  - urban dna not related to evolutionary processes in existing literature
+
 % https://scholar.google.com/scholar?hl=fr&as_sdt=0%2C5&q=meme+gene+coevolution&btnG=
 
-\cite{votsis2019urban}
+\cite{Eletreby5664} diffusion/evolving
+
+\cite{votsis2019urban} use the concept of Urban DNA to characterize morphological properties of cities such as density or the role of the road network.
+Similarly, \cite{kaya2017urban} describe cities based on their morphological properties
+
+\cite{wu2011urban}
+
+% (iii) Innovation diffusion - in ALife studies ; - in urban systems
+
+% (iv) contribution
 
@@
 
-The principal idea in the model is to include a concept of ``Urban DNA'', i.e.
+The principal idea in the model is to build on a concept of ``Urban DNA'' which would capture evolution processes as in biological evolution and cultural evolution, i.e.
 a kind of genome that cities would be exchanging and which would undergo mutation processes.
@@
 
+\cite{batty2009digital}
+
+\cite{blommestein1987adoption} describes a model of innovation diffusion and urban dynamics with endogenous demand for innovations, but in which the spatial component only influences prices of innovations.
+
+\cite{deffuant2005individual} give an example of an elaborated model for adoption dynamics at the microscopic level.
+
+Effective channels for the diffusion of innovations are multiple, and can for example be urban firm linkages \citep{rozenblat2007firm}.
+
 \subsection{Model description}
 
-\cite{favaro2011gibrat}
+% - urban genome as a matrix (one for each city)
+% - rows: % of each innovation (to simplify, sums to 1), columns: dimension of "urban life": transportation, economic production, leisure, built environment, governance, etc.
+% - mutations: new innovations: new row.
+"endogenous innovation": add new column (dimension that did not exist before: new info technologies, induce niches, new needs, new dimensions in themselves.
+too complicated, do not include this aspect for the sake of simplicity, consider "fundamental dimensions" only
+% - interdependance between dimensions? -> should yield "technological lockdowns". Q: how to implement? SBM? additional correlation matrix/ parameter?
+% - innovation utility matrix: stochastic, not necessarily strongly increasing (should be similar though)
+% - crossovers is hierarchical spatial diffusion: spatial interaction model
+% - ! no network evolving (too complicated
+% - evolves pop following Favaro-Pumain. P(innov) = scaling (pop) / utility ? ! not pop, gdp? (ok for eco innovation and sectors - should be multi-dimensional indeed)
+
+Our model is inspired from the urban dynamics model of \cite{favaro2011gibrat}.
+
+\paragraph{Model dynamics}
+
+% $\rightarrow$ Work under Gibrat independence assumptions, i.e.
+$\Covb{P_i(t)}{P_j(t)}=0$.
+If $\vec{P}(t+1)=\mathbf{R}\cdot \vec{P}(t)$ where $\mathbf{R}$ is also independent, then $\Eb{\vec{P}(t+1)}=\Eb{\mathbf{R}}\cdot\Eb{\vec{P}}(t)$.
+Consider expectancies only (higher moments computable similarly)
+%$\rightarrow$ With $\vec{\mu}(t)=\Eb{\vec{P}(t)}$, we generalize this approach by taking $\vec{\mu}(t+1)=f(\vec{\mu}(t))$
+
+%Let $\vec{\mu}(t)=\Eb{\vec{P}(t)}$ cities population and $(d_{ij})$ distance matrix
+
+%Model specified by
+%\[
+%f(\vec{\mu}) = r_0\cdot \mathbf{Id}\cdot \vec{\mu} + \mathbf{G}\cdot \mathbf{1}
+%\]
+
+Innovation occur along dimensions $1 \leq d \leq D$, and are indexed by their order of apparition $c$.
+
+The crossover between urban genomes relies on spatial processes of innovation diffusion, following a spatial interaction model given by
+
+\[
+\delta_{c,i,t} = \frac{\sum_j p_{c,j,t-1}^{s_c} \exp{(-\lambda_s d_{ij})}}{\sum_c \sum_j p_{c,j,t-1}^{s_c} \exp{(-\lambda_s d_{ij})}}
+\]
+
+The sizes of cities evolve according to their performance in terms of innovation, i.e. more innovative cities are more attractive, with $G_{ij} = w_G\cdot \frac{V_{ij}}{<V_{ij}>}$ such that
+\[
+V_{ij}= \frac{p_i p_j}{(\sum_k p_k)^2} \exp{(-\lambda_m d_{ij} \prod_c \delta_{c,i}^{\phi_c})}
+\]
+with $\phi_c = \sum_i p_{i,c}/\sum_{i,c} p_{i,c}$
+
+Mutation corresponds to the introduction of new innovations with utility $s_{c+1} = g_0 \cdot s_c$ in a randomly chosen city with a hierarchy parameter $\alpha_I$, if global adoption share $\phi_c$ is larger than a threshold $\theta_I$.
+Initial utility $s_0$ is a parameter.
+New innovation has an initial penetration rate $r_I$ in the city.
+
+\paragraph{Synthetic setup}
+
+\paragraph{Indicators}
+
+% - Indicators: innovation diversity (on cities profiles); overall utility (optimal config as a function of lambda - non trivial?) - with path-dep/correlations?; existence of cycles?
+%.   compute utility for one city: along each dimension, weight utility of innovations. then aggregate? ~
 

### 2020-04-22 Models/QuantEpistemo/citation.R

@@ -1,0 +1,224 @@
+setwd(paste0(Sys.getenv('CS_HOME'),'/UrbanEvolution/Models/QuantEpistemo'))
+
+library(dplyr)
+library(igraph)
+library(glue)
+library(reshape2)
+library(ggplot2)
+library(heatmaply)
+
+source(paste0(Sys.getenv('CS_HOME'),'/Organisation/Models/Utils/R/plots.R'))
+source('functions.R')
+
+
+edge_file = 'export/alife_links.csv'
+node_file = 'export/alife.csv'
+
+edges <- read.csv(edge_file,sep=";",header=F,colClasses = c('character','character'))
+nodes <- as.tbl(read.csv(node_file,sep=";",stringsAsFactors = F,quote = '"',colClasses = rep('character',9)))
+
+
+# numeric hdepth values (incl. NA)
+for(j in 8:9){nodes[,j]=as.numeric(unlist(nodes[,j]))}
+kws = names(nodes)[8:9]
+
+# remove edges which have been filtered
+edges = edges[edges$V1%in%nodes$id&edges$V2%in%nodes$id,]
+
+nodes = cbind(nodes,numHorizontalDepth=apply(nodes[,8:9],1,function(r){min(r,na.rm=T)}))
+nodes = nodes[is.finite(nodes$numHorizontalDepth),]
+
+citation <- graph_from_data_frame(edges,vertices = nodes)
+
+# connected components
+components(citation)$csize
+
+citation = induced_subgraph(citation,which(components(citation)$membership==1))
+
+#V(citation)$reduced_title = sapply(V(citation)$title,function(s){paste0(substr(s,1,30),"...")})
+#V(citation)$reduced_title = ifelse(degree(citation)>50,V(citation)$reduced_title,rep("",vcount(citation)))
+
+citationcorehigher = induced_subgraph(citation,which(degree(citation)>1))
+while(length(which(degree(citationcorehigher)==1))>0){citationcorehigher = induced_subgraph(citationcorehigher,which(degree(citationcorehigher)>1))}
+
+# csv export
+export_gml(citationcorehigher,'processed/core.gml')
+save(citationcorehigher,file='processed/core.RData')
+save(citation,file='processed/full.RData')
+
+# export subgraphs alife / ai
+coreai = induced_subgraph(citationcorehigher,!is.na(V(citationcorehigher)$ai))
[…]

### 2020-04-22 Models/QuantEpistemo/functions.R

@@ -1,0 +1,171 @@
+library(Matrix)
+
+
+
+getLogCitationFlows <- function(g,com,comindices,comnames){
+  A=as_adjacency_matrix(g,sparse = T)
+  citprops = matrix(0,length(comindices),length(comindices))
+  for(i in 1:length(comindices)){
+    alloutcits = sum(A[com$membership==comindices[i],])
+    for(j in 1:length(comindices)){
+      show(paste0(i,",",j))
+      citprops[i,j] = sum(A[com$membership==comindices[i],com$membership==comindices[j]])/alloutcits
+    }
+  }
+  show(max(citprops))
+  show(min(citprops[citprops>0]))
+    
+  lp=log(citprops)
+  lp[is.infinite(lp)]=NA
+  lp= (lp - min(lp,na.rm = T)) / (max(lp,na.rm = T) - min(lp,na.rm = T))
+  lp[is.na(lp)]=0 # not super clean
+  rownames(lp)<-unlist(comnames[as.character(comindices)])
+  colnames(lp)<-unlist(comnames[as.character(comindices)])
+  return(lp)
+}
+
+
+
+nwProperties <- function(citnw){
+  A=as_adjacency_matrix(citnw,sparse = T)
+  M = A+t(A)
+  undir_citnw = graph_from_adjacency_matrix(M,mode="undirected")
+  coms = cluster_louvain(undir_citnw)
+  return(c(
+    modularity = modularity(coms),
+    #directedModularity = directedmodularity(membership(coms),A), # long to compute and very close to undir
+    communitiesnumber = length(unique(membership(coms))),
+    avgDegree = mean(degree(citnw,mode = 'all')),
+    avgInDegree = mean(degree(citnw,mode = 'in')),
+    avgOutDegree = mean(degree(citnw,mode = 'in')),
+    alphaDegree = hierarchy(degree(citnw,mode = 'all')),
+    alphaInDegree = hierarchy(degree(citnw,mode = 'in')),
+    alphaOutDegree = hierarchy(degree(citnw,mode = 'out')),
+    ecount = ecount(citnw),
+    vcount = vcount(citnw)
+  ))
+}
+
+export_gml <- function(citnw,exportfile){
+  V(citnw)$reduced_title = sapply(V(citnw)$title,function(s){paste0(substr(s,1,50),"...")})
[…]

### 2020-04-29 Docs/Papers/EvolutionInnovation/paper.tex
run_model

@@
 
-The principal idea in the model is to build on a concept of ``Urban DNA'' which would capture evolution processes as in biological evolution and cultural evolution, i.e.
+The core idea of the model is to build on a concept of ``Urban DNA'' which would capture evolution processes as in biological evolution and cultural evolution, i.e.
 a kind of genome that cities would be exchanging and which would undergo mutation processes.

### 2020-04-29 Models/EvolutionInnovation/src/main/scala/urbanevolution/UrbanEvolution.scala

@@ -1,0 +1,51 @@
+
+package urbanevolution
+
+
+import org.openmole.spatialdata.model.urbandynamics._
+import org.openmole.spatialdata.utils.math.Statistics
+
+import scala.util.Random
+
+
+case class UrbanEvolution(
+                          model: Innovation
+                         ) {
+
+  def run(): (Double,Double,Double,Double) = {
+    val result = Innovation.run(model)
+    val finalHierarchy = Statistics.slope(result.macroResult.simulatedPopulation.getCol(model.dates.length).flatValues)._1
+    (result.averageDiversity,result.averageUtility,result.averageInnovation,finalHierarchy)
+  }
+
+}
+
+object UrbanEvolution {
+
+  def apply(
+            gravityDecay: Double,
+            innovationDecay: Double,
+            mutationRate: Double,
+            newInnovationHierarchy: Double,
+            earlyAdoptersRate: Double,
+            utilityStd: Double,
+            utilityDistribution: String,
+            syntheticCities: Int,
+            syntheticHierarchy: Double,
+            syntheticMaxPop: Double,
+            finalTime: Int,
+            growthRate: Double,
+            innovationWeight: Double,
+            seed: Int
+           ): UrbanEvolution = {
+
+    implicit val rng: Random = new Random(seed)
+
+    UrbanEvolution(
+      Innovation(syntheticCities, syntheticHierarchy, syntheticMaxPop, finalTime, seed, growthRate, innovationWeight, gravityDecay,
+        innovationDecay, mutationRate, newInnovationHierarchy, earlyAdoptersRate, utilityStd, utilityDistribution)
+    )
+
+  }
+
+}

### 2020-04-29 Models/EvolutionInnovation/src/main/scala/urbanevolution/CSV.scala
reverse_to_scala2.12

@@ -1,0 +1,59 @@
+package urbanevolution
+
+import java.io.{BufferedReader, File, FileReader}
+import com.github.tototoshi.csv._
+
+import scala.collection.mutable.ArrayBuffer
+
+object CSV {
+
+
+  /**
+    * read a numerical matrix
+    * @param file
+    * @param sep
+    * @return
+    */
+  def readMat(file: String,sep: String=",",naformat: String = "NA"): Array[Array[Double]] = {
+    val r = new BufferedReader(new FileReader(new File(file)))
+    val res = new ArrayBuffer[Array[Double]]
+    var currentline = r.readLine()
+    while(currentline!=null){
+      res.append(currentline.split(sep).map{s => if(s.equals("NA")) Double.NaN else s.toDouble})
+      currentline = r.readLine()
+    }
+    res.toArray
+  }
+
+  /**
+    * read csv (file name)
+    * @param filename
+    * @param sep
+    * @param withHeader
+    * @return
+    */
+  def readCSV(filename: String, sep: String = ",", withHeader: Boolean =true): Map[String,Seq[String]] = readCSVFile(new File(filename),sep, withHeader)
+
+
+  /**
+    * read csv file
+    * @param file
+    * @param sep
+    * @return
+    */
+  def readCSVFile(file: File,sep: String=",", withHeader: Boolean =true): Map[String,Seq[String]] = {
+
+    implicit val readerFormat = new DefaultCSVFormat {
+      override val delimiter = sep.charAt(0)
+      override val quoteChar: Char = '"'
+    }
+    val (header,content): (Seq[String], Seq[Map[String, String]]) = if (withHeader) CSVReader.open(file)(readerFormat).allWithOrderedHeaders()
+    else {
+      val raw = CSVReader.open(file)(readerFormat).all()
+      val h =(0 until raw(0).length).map("V"+_)
+      (h,raw.map{row => row.zip(h).toMap})
+    }
+    header.map{case s => (s,content.map{_(s)})}.toMap
+  }
+
+}

### 2020-04-29 Models/EvolutionInnovation/src/main/scala/urbanevolution/Matrix.scala

@@ -1,0 +1,349 @@
+package urbanevolution
+
+import org.apache.commons.math3.linear
+import org.apache.commons.math3.linear.{LUDecomposition, MatrixUtils}
+
+import scala.util.Random
+
+
+/**
+  * Generic methods for double matrices
+  *   rq: could make generic in Numeric type?
+  */
+sealed trait Matrix {
+
+  /**
+    * rq: in immutable terms, setting is not efficient (should clone the underlying matrix?)
+    * @param i row
+    * @param j column
+    * @param v value
+    * @return cloned matrix with value set
+    */
+  def set(i: Int, j: Int, v: Double): Matrix
+
+  /**
+    * Mutable set
+    * @param i row
+    * @param j column
+    * @param v value
+    */
+  def setM(i: Int, j: Int, v: Double): Unit
+
+  /**
+    * get element
+    * @param i row
+    * @param j column
+    * @return value
+    */
+  def get(i: Int, j: Int): Double
+
+  /**
+    * get submatrix
+    * @param i starting row
+    * @param j starting column
+    * @param nrows number of rows
+    * @param ncols number of columns
+    * @return submatrix
+    */
+  def getSubmat(i: Int, j: Int, nrows: Int, ncols: Int): Matrix
+
+  def getRow(i: Int): Matrix = getSubmat(i,0,1,ncols)
+  def getCol(j: Int): Matrix = getSubmat(0,j,nrows,1)
+
+  /**
+    * unoptimal mutable submat set
+    * @param i starting row
+    * @param j starting col
+    * @param a values
+    */
+  def setMSubmat(i: Int, j: Int, a: Array[Array[Double]]): Unit = {
+    a.zipWithIndex.foreach{case (row,k) =>
+      row.zipWithIndex.foreach{case (v,l) =>
+        if(i+k<nrows&&j+l<ncols) setM(i+k,j+l,v)
+      }
+    }
+  }
+
+  /**
+    * map element by element
+    * @param f function to map
+    * @return
+    */
+  def map(f: Double => Double): Matrix
+
+  // !! mutable
+  def values: Array[Array[Double]]
+
+  def flatValues: Array[Double]
+
+  def nrows: Int
+  def ncols: Int
+
+  //basic ring operations (R style for the notation)
+  def %*%(m: Matrix): Matrix
+
+  // scalar operations
[…]

### 2020-04-29 Models/EvolutionInnovation/src/main/scala/urbanevolution/RandomPointsGenerator.scala

@@ -1,0 +1,31 @@
+package urbanevolution
+
+import scala.util.Random
+
+
+/**
+  * Random points
+  *
+  * @param npoints
+  */
+case class RandomPointsGenerator(
+                                  npoints: Int,
+                                  xmin: Double = 0.0,
+                                  xmax: Double = 1.0,
+                                  ymin: Double = 0.0,
+                                  ymax: Double = 1.0
+                                )  {
+  def generatePoints(implicit rng: Random): Vector[(Double,Double)] =
+    RandomPointsGenerator.randomPoints(this)(rng)
+}
+
+
+
+object RandomPointsGenerator {
+
+  def randomPoints(generator: RandomPointsGenerator)(implicit rng: Random): Vector[(Double,Double)] =
+  //Array.fill(generator.npoints){(math.digits(generator.xmin + (generator.xmax - generator.xmin)*rng.nextDouble,4),
+  //  math.digits(generator.ymin + (generator.ymax - generator.ymin)*rng.nextDouble,4))}
+    Vector.fill(generator.npoints){(generator.xmin + (generator.xmax - generator.xmin)*rng.nextDouble,
+      generator.ymin + (generator.ymax - generator.ymin)*rng.nextDouble)}
+}

### 2020-04-29 Models/EvolutionInnovation/src/main/scala/urbanevolution/RunUrbanEvolution.scala

@@ -1,0 +1,35 @@
+package urbanevolution
+
+import scala.util.Random
+
+object RunUrbanEvolution extends App {
+
+  implicit val rng: Random = new Random
+
+  val model = UrbanEvolution(
+    syntheticCities = 30,
+    syntheticHierarchy = 1.0,
+    syntheticMaxPop =  100000.0,
+    finalTime = 50,
+    seed = rng.nextInt(),
+    growthRate = 0.0,
+    innovationWeight = 0.005,
+    gravityDecay = 0.5,
+    innovationDecay = 0.3,
+    mutationRate = 0.4,
+    newInnovationHierarchy = 0.5,
+    earlyAdoptersRate = 0.2,
+    utilityStd = 1.0,
+    utilityDistribution = "log-normal" // "normal"
+  )
+
+  val result = UrbanEvolution.run(model)
+
+  println("Average diversity = "+result.averageDiversity)
+  println("Average utility = "+result.averageUtility)
+  println("Average innovation = "+result.averageInnovation)
+  val finalHierarchy = Statistics.slope(result.populations.getCol(50).flatValues)._1
+  println("Final pop hierarchy = "+finalHierarchy)
+  println("Delta hierarchy pop = "+(finalHierarchy - Statistics.slope(result.populations.getCol(0).flatValues)._1))
+
+}

### 2020-04-29 Models/EvolutionInnovation/src/main/scala/urbanevolution/Spatstat.scala

@@ -1,0 +1,27 @@
+package urbanevolution
+
+import org.apache.commons.math3.linear.MatrixUtils
+
+object Spatstat {
+
+  /**
+    * Euclidian distance matrix
+    *  rq: unoptimal, + already coded with great circle dist in geotools
+    * @param pi points
+    * @return
+    *
+    */
+  def euclidianDistanceMatrix(pi: Array[(Double,Double)]): Array[Array[Double]] = {
+    val n = pi.length
+    val xcoords = MatrixUtils.createRealMatrix(Array.fill(n)(pi.map(_._1)))
+    val ycoords = MatrixUtils.createRealMatrix(Array.fill(n)(pi.map(_._2)))
+    MatrixUtils.createRealMatrix(
+      xcoords.subtract(xcoords.transpose()).getData.map(_.map( x => x*x))
+    ).add(
+      MatrixUtils.createRealMatrix(
+        ycoords.subtract(ycoords.transpose()).getData.map(_.map( x => x*x))
+      )
+    ).getData.map(_.map(math.sqrt))
+  }
+
+}

### 2020-04-29 Models/EvolutionInnovation/src/main/scala/urbanevolution/Statistics.scala

@@ -1,0 +1,28 @@
+package urbanevolution
+
+import org.apache.commons.math3.stat.regression.SimpleRegression
+
+import scala.math.{Ordering, log}
+
+object Statistics {
+
+  /**
+    * Distribute a variable following a rank size
+    * @param size number of elements
+    * @param alpha slope
+    * @param pmax max value
+    * @return
+    */
+  def rankSizeDistribution(size: Int,alpha: Double, pmax: Double): Vector[Double] =
+    (1 to size by 1).map{i => pmax*math.pow(1.0/i.toDouble,alpha)}.toVector
+
+
+  def slope(values: Array[Double]): (Double,Double) = {
+    def distribution: Array[Double] = values.filter(_ > 0).sorted(Ordering.Double.reverse)
+    def distributionLog: Array[Array[Double]] = distribution.zipWithIndex.map { case (q, i) => Array(log(i + 1), log(q)) }
+    val simpleRegression = new SimpleRegression(true)
+    simpleRegression.addData(distributionLog)
+    (simpleRegression.getSlope, simpleRegression.getRSquare)
+  }
+
+}

### 2020-04-29 Models/EvolutionInnovation/src/main/scala/urbanevolution/Stochastic.scala

@@ -1,0 +1,34 @@
+package urbanevolution
+
+import scala.util.Random
+
+object Stochastic {
+
+  sealed trait Distribution {
+    def rng: Random
+    def draw: Double
+  }
+
+  case class NormalDistribution(mu: Double, sigma: Double)(implicit localRng: Random) extends Distribution {
+    override def rng: Random = localRng
+    override def draw: Double = mu + sigma*rng.nextGaussian()
+  }
+
+  case class LogNormalDistribution(mu: Double, sigma: Double)(implicit localRng: Random) extends Distribution {
+    override def rng: Random = localRng
+    override def draw: Double = math.exp(mu + sigma*rng.nextGaussian())
+  }
+  object LogNormalDistribution {
+    def fromMoments(average: Double, std: Double)(implicit localRng: Random): LogNormalDistribution = {
+      assert(average>0,"Average of a log-normal should be strictly positive")
+      assert(std>0,"Std of a log-normal should be strictly positive")
+      assert(math.log(std)<math.log(average)-0.5,"Std of log normal too low regarding average")
+      LogNormalDistribution(
+        mu = 2 * math.log(average) - math.log(std) - 0.5,
+        sigma = math.sqrt(1 - 2*(math.log(average) - math.log(std)))
+      )
+    }
+  }
+
+
+}

### 2020-04-29 Models/EvolutionInnovation/src/main/scala/urbanevolution/UrbanEvolution.scala

@@ -2,50 +2,471 @@
 package urbanevolution
 
 
-import org.openmole.spatialdata.model.urbandynamics._
-import org.openmole.spatialdata.utils.math.Statistics
+import urbanevolution.Matrix.MatrixImplementation
 
+import scala.collection.mutable.ArrayBuffer
 import scala.util.Random
 
 
-case class UrbanEvolution(
-                          model: Innovation
-                         ) {
 
-  def run(): (Double,Double,Double,Double) = {
-    val result = Innovation.run(model)
-    val finalHierarchy = Statistics.slope(result.macroResult.simulatedPopulation.getCol(model.dates.length).flatValues)._1
-    (result.averageDiversity,result.averageUtility,result.averageInnovation,finalHierarchy)
+/**
+  *
+  * Generalization of the Favaro-Pumain model for urban systems dynamics based on the diffusion of innovation
+  *
+  *   Favaro, J. M., & Pumain, D. (2011). Gibrat Revisited: An Urban Growth Model Incorporating Spatial Interaction and Innovation Cycles. Geographical Analysis, 43(3), 261-286.
+  *
+  * @param populationMatrix        Real population matrix
+  * @param distanceMatrix          Distance matrix
+  * @param dates                   Dates
+  * @param rng                     Model has its own rng
+  * @param growthRate              Gibrat growth rate
+  * @param innovationWeight        weight of innovation induced growth rate
+  * @param gravityDecay            Decay of gravity interaction
+  * @param innovationDecay         Decay of innovation diffusion
+  */
+case class UrbanEvolution(
+                       populationMatrix: Matrix,
+                       distanceMatrix: Matrix,
+                       dates: Array[Double],
+                       rng : Random,
+                       growthRate: Double,
+                       innovationWeight: Double,
+                       gravityDecay: Double,
+                       innovationDecay: Double,
+                       newInnovation: (Seq[Double],Seq[Double],Seq[Seq[Double]]) => (Boolean,Seq[Double],Seq[Seq[Double]]),
+                       initialInnovationUtility: Double
+                     ) {
+
+  /**
+    * run for model calibration
+    * @return
+    */
+  def run: (Double, Double, Double, Double) = {
+    val result = UrbanEvolution.run(this)
[…]

### 2020-04-30 Docs/Papers/EvolutionInnovation/paper.tex
exploration_results

@@
 
-\title{A simple model of urban evolution based on innovation diffusion}
+% Modeling urban evolution with the diffusion of innovations
+% A simple model of urban evolution based on innovation diffusion
+% A model of urban evolution based on innovation diffusion
+% A model of urban evolution and innovation diffusion
+% A model of urban evolution based on the diffusion of innovations
+
+\title{A model of urban evolution based on innovation diffusion}
 \author{Juste Raimbault$^{1,2,3}$\\
@@
 
-Urban systems 
-\cite{batty2009centenary}
+Urban systems are complex as they combine technical artefacts with socio-economic dynamics at multiple temporal and spatial scales.
+An understanding of processes driving their dynamics is an important aspect for shaping sustainable policies, and a sustainable design and management of cities \citep{lobo2020urban}.
+Multiple disciplines and viewpoints have proposed such insights from a complexity perspective \citep{pumain2020conclusion}, and Artificial Life approaches have shown promising results to study urban systems, both through conceptual contributions such as interpreting cities through biological metaphors \citep{batty2009centenary}, but also through modeling and simulation for urban growth including cellular automata and evolutionary computation \citep{raimbault2020cities}.
 
 % (ii) Urban Evolution / Cultural evolution
-%  - urban dna not related to evolutionary processes in existing literature
 
-% https://scholar.google.com/scholar?hl=fr&as_sdt=0%2C5&q=meme+gene+coevolution&btnG=
-
-\cite{Eletreby5664} diffusion/evolving
-
-\cite{votsis2019urban} use the concept of Urban DNA to characterize morphological properties of cities such as density or the role of the road network.
-Similarly, \cite{kaya2017urban} describe cities based on their morphological properties
-
-\cite{wu2011urban}
+A stream of research tightly linked to ALife relates to concepts of \emph{Urban Evolution}.
+These build in a sense on Cultural Evolution \citep{mesoudi2001cultural}, which aims at understanding changes in social knowledge as evolutionary processes involving replication, mutation, selection.
+It bears similarities with biological evolution but is not reducible to it, and for example uses the concept of \emph{meme} as transmission units comparable to genes.
+Both biological and cultural evolution can be linked into common frameworks and models, implying the coupling of different timescales \citep{bull2000meme}.
+In terms of urban studies, the concept of urban evolution is less formalized than cultural evolution and can be understood in multiple ways.
+\cite{votsis2019urban} use the concept of Urban DNA to characterize morphological properties of cities such as population density or the role of the road network.
+Similarly, \cite{kaya2017urban} describe cities based on their morphological properties as a product of their dynamics.
+\cite{wu2011urban} link the parameters of a cellular automaton model of urban growth to intrinsic properties of urban regions, which should play a role in their overall evolution.
+In Economic Geography, the concept of co-evolution is used mostly for urban agents such as firms and stakeholders \citep{gong2019co}.
+At the macroscopic scale of urban systems, \cite{pumain2018evolutionary} proposes an evolutionary theory to study systems of cities as complex adaptive systems, in which interactions between cities play a crucial role.
+Corresponding simulation models were proposed with different thematic focus, for example by \cite{cottineau2015modular} with economic exchanges and by \cite{raimbault2020indirect} with infrastructure networks.
+There still however remains a lack of models which would operationalize the concept of urban evolution in a way close to its biological and cultural counterparts.
 
 % (iii) Innovation diffusion - in ALife studies ; - in urban systems
+%\cite{Eletreby5664} diffusion/evolving
+
+Besides, a central concept bridging ALife and evolution with the study of urban systems is the concept of innovation diffusion.
+
+\cite{hagerstrand1968innovation}
 
@@
 
+\cite{kiesling2012agent}
+
 Effective channels for the diffusion of innovations are multiple, and can for example be urban firm linkages \citep{rozenblat2007firm}.
@@
 
-Our model is inspired from the urban dynamics model of \cite{favaro2011gibrat}.
+Our model is inspired from the urban dynamics model of \cite{favaro2011gibrat}, for which crossover and mutation processes were explicited and
 
@@
 
+The model is implemented in \texttt{scala} and integrated into the OpenMOLE software \citep{reuillon2013openmole} for numerical experiments.
+OpenMOLE allows embedding models in any language as black boxes, provides a transparent access to high performance computing infrastructures, and model exploration and validation methods such as sensitivity analysis, design of experiments, and calibration methods.
+Experiments are designed through workflows using a Domain Specific Language \citep{passerat2017reproducible}.
+
+Source code of the model and analysis, and results are available on the open git repository of the project at \url{}.
+Simulation data files are available on the dataverse repository at \url{https://doi.org/10.7910/DVN/Q5GKZ0}.
+
+Parameter explored in experiments are
+
+\subsection{Model exploration}
+
+% summary(reldistance("meanAverageDiversity","sdAverageDiversity"))
+%   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+% 0.0000  0.6035  1.2685  1.4709  2.1557  6.1199 
+%> summary(reldistance("medianAverageDiversity","sdAverageDiversity"))
+%   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+% 0.0000  0.5733  1.2115  1.4425  2.0941  6.5835 
+% summary(reldistance("meanAverageInnovation","sdAverageInnovation"))
+%   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+%  0.000   1.839   4.136   4.768   7.011  15.799 
+%> summary(reldistance("medianAverageInnovation","sdAverageInnovation"))
+%   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+%  0.000   1.841   4.176   4.783   7.100  15.891 
+%summary(reldistance("meanAverageUtility","sdAverageUtility"))
+%   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+%  0.000   0.509   1.103   1.320   1.923   6.350 
+%> summary(reldistance("medianAverageUtility","sdAverageUtility"))
+%   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+% 0.0000  0.4942  1.0857  1.3013  1.8863  6.3479 
+%summary(reldistance("meanFinalHierarchy","sdFinalHierarchy"))
+%   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+%  0.000   2.488   6.066  11.424  14.133  96.805 
+%> summary(reldistance("medianFinalHierarchy","sdFinalHierarchy"))
+%   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+%  0.000   2.508   6.074  11.449  14.171  96.789 
+% idea: on this "relative dist": index similar to Saltelli to tell distinguishability across the all parameter space?
+
+The sharpe ratios estimated on repetitions as sample standard deviation relative to absolute sample mean have high values for all indicators (min.
+3.9 and median 12.1 for diversity; min.
+3.0 and median 6.2 for innovation; min.
+1.7 and median 3.6 for utility; min.
+26 and median 257.3 for population hierarchy), what means that stochastic noise is not an issue for interpreting indicator values.
+We also study the distance between points
+
 \section{Discussion}
 
-%\section{Acknowledgements}
+Several extensions and applications would be possible to this first model exploration.
+First, the innovation space in our model remained unidimensional.
+\citep{hidalgo2007product} \citep{bergeaud2017classifying}
+Second, applying the model to real system of cities \cite{raimbault:halshs-01880492}
+Finally, % multiscalar?
+
+\section{Acknowledgements}
+
+Results obtained in this paper were computed on the vo.complex-system.eu virtual organization of the European Grid Infrastructure ( http://www.egi.eu ).
+We thank the European Grid Infrastructure and its supporting National Grid Initiatives (France-Grilles in particular) for providing the technical support and infrastructure.
+This work was funded by the Urban Dynamics Lab grant EPSRC EP/M023583/1.
 

### 2020-04-30 Models/EvolutionInnovation/analysis.R

@@ -1,0 +1,125 @@
+
+setwd(paste0(Sys.getenv('CS_HOME'),'/UrbanEvolution/Models/EvolutionInnovation'))
+
+library(dplyr)
+library(ggplot2)
+
+source(paste0(Sys.getenv('CS_HOME'),'/Organisation/Models/Utils/R/plots.R'))
+
+params<-c("gravityDecay","innovationDecay","mutationRate","newInnovationHierarchy","utilityStd","earlyAdoptersRate","utilityDistribution","initialHierarchy")
+paramnames=list(gravityDecay=expression(d[G]),innovationDecay=expression(d[I]),mutationRate=expression(beta),newInnovationHierarchy=expression(alpha[I]),utilityStd=expression(sigma[U]),earlyAdoptersRate=expression(delta[0]),utilityDistribution="Distribution")
+indics<-c("averageDiversity","averageInnovation","averageUtility","finalHierarchy")
+indicnames = list(averageDiversity="Diversity",averageInnovation="Innovation",averageUtility="Utility",finalHierarchy="Population hierarchy")
+
+# stochasticity
+resprefix = '20200429_150801_STOCHASTICITY_GRID'
+resdir = paste0(Sys.getenv('CS_HOME'),'/UrbanEvolution/Results/EvolutionInnovation/',resprefix,'/');dir.create(resdir)
+
+res <- as.tbl(read.csv(file=paste0('openmole/exploration/',resprefix,'.csv')))
+
+# filter NAs and absurb utility values
+res=res[res$averageUtility<quantile(res$averageUtility,c(0.99),na.rm = T)&!is.na(res$averageUtility),]
+
+# sample hists
+#seed=0
+seed=42
+set.seed(seed)
+res$id = paste0(res$gravityDecay,res$innovationDecay,res$mutationRate,res$newInnovationHierarchy,res$utilityStd,res$earlyAdoptersRate,res$utilityDistribution)
+intid = 1:length(unique(res$id));names(intid)<-unique(res$id)
+res$id = as.character(intid[res$id])
+ids = sample(unique(res$id),size = 20,replace = F)
+
+for(indic in indics){
+  g=ggplot(res[res$id%in%ids,],aes_string(x=indic,color="id",group="id"))
+  g+geom_density()+facet_wrap(~utilityDistribution)+scale_color_discrete()+stdtheme
+  ggsave(file=paste(resdir,indic,'_seed',seed,'.png'),width=20,height=15,units='cm')
+}
+
+
[…]

### 2020-04-30 Models/EvolutionInnovation/src/main/scala/urbanevolution/UrbanEvolution.scala

@@ -43,7 +43,7 @@ case class UrbanEvolution(
     */
   def run: (Double, Double, Double, Double) = {
     val result = UrbanEvolution.run(this)
-    (result.averageDiversity,result.averageUtility,result.averageInnovation,Statistics.slope(result.populations.getCol(dates.length).flatValues)._1)
+    (result.averageDiversity,result.averageUtility,result.averageInnovation,Statistics.slope(result.populations.getCol(dates.length - 1).flatValues)._1)
   }
 
 
@@ -210,8 +210,9 @@ object UrbanEvolution {
              earlyAdoptersRate: Double,
              utilityStd: Double,
              utilityDistribution: String
-           )(implicit rng: Random): UrbanEvolution = {
+           ): UrbanEvolution = {
     implicit val m: MatrixImplementation = Matrix.defaultImplementation
+    implicit val rng: Random = new Random
     rng.setSeed(seed.toLong)
 
     val dmat = Matrix(Spatstat.euclidianDistanceMatrix(RandomPointsGenerator(syntheticCities).generatePoints.toArray))

### 2020-04-30 Models/EvolutionInnovation/analysis.R
calibration

@@ -6,7 +6,7 @@ library(ggplot2)
 
 source(paste0(Sys.getenv('CS_HOME'),'/Organisation/Models/Utils/R/plots.R'))
 
-params<-c("gravityDecay","innovationDecay","mutationRate","newInnovationHierarchy","utilityStd","earlyAdoptersRate","utilityDistribution","initialHierarchy")
+params<-c("gravityDecay","innovationDecay","mutationRate","newInnovationHierarchy","utilityStd","earlyAdoptersRate","utilityDistribution")#,"initialHierarchy")
 paramnames=list(gravityDecay=expression(d[G]),innovationDecay=expression(d[I]),mutationRate=expression(beta),newInnovationHierarchy=expression(alpha[I]),utilityStd=expression(sigma[U]),earlyAdoptersRate=expression(delta[0]),utilityDistribution="Distribution")
 indics<-c("averageDiversity","averageInnovation","averageUtility","finalHierarchy")
 indicnames = list(averageDiversity="Diversity",averageInnovation="Innovation",averageUtility="Utility",finalHierarchy="Population hierarchy")
@@ -70,6 +70,7 @@ resdir = paste0(Sys.getenv('CS_HOME'),'/UrbanEvolution/Results/EvolutionInnovati
 
 res <- as.tbl(read.csv(file=paste0('openmole/exploration/',resprefix,'.csv')))
 res=res[res$averageUtility<quantile(res$averageUtility,c(0.95),na.rm = T)&!is.na(res$averageUtility),]
+res1=res
 
 for(indic in indics){
   
@@ -95,6 +96,7 @@ resdir = paste0(Sys.getenv('CS_HOME'),'/UrbanEvolution/Results/EvolutionInnovati
 
 res <- as.tbl(read.csv(file=paste0('openmole/exploration/',resprefix,'.csv')))
 res=res[res$averageUtility<quantile(res$averageUtility,c(0.95),na.rm = T)&!is.na(res$averageUtility),]
+res2=res
 
 for(indic in indics){
   for(distrib in unique(res$utilityDistribution)){
@@ -107,15 +109,66 @@ for(indic in indics){
           xlab(expression(d[G]))+ylab(indicnames[[indic]])+scale_color_continuous(name=expression(d[I]))+stdtheme
[…]

### 2020-04-30 Docs/Papers/EvolutionInnovation/arxiv/paper.tex
papers

\documentclass[letterpaper]{article}

\usepackage{natbib,alifeconf}  %% The order is important
\usepackage{hyperref}

% *****************
%  Requirements:
% *****************
%
% - All pages sized consistently at 8.5 x 11 inches (US letter size).
% - PDF length <= 8 pages for full papers, <=2 pages for extended
%    abstracts (not including citations).
% - Abstract length <= 250 words.
% - No visible crop marks.
% - Images at no greater than 300 dpi, scaled at 100%.
% - Embedded open type fonts only.
% - All layers flattened.
% - No attachments.
% - All desired links active in the files.

% Note that the PDF file must not exceed 5 MB if it is to be indexed
% by Google Scholar. Additional information about Google Scholar
% can be found here:
% http://www.google.com/intl/en/scholar/inclusion.html.

% If your system does not generate letter format documents by default,
% you can use the following workflow:
% latex example
% bibtex example
% latex example ; latex example
% dvips -o example.ps -t letterSize example.dvi
% ps2pdf example.ps example.pdf

% For pdflatex users:
% The alifeconf style file loads the "graphicx" package, and
% this may lead some users of pdflatex to experience problems.
% These can be fixed by editing the alifeconf.sty file to specify:
% \usepackage[pdftex]{graphicx}
%   instead of
% \usepackage{graphicx}.
% The PDF output generated by pdflatex should match the required
% specifications and obviously the dvips and ps2pdf steps become
% unnecessary.

% Note:  Some laser printers have a serious problem printing TeX
% output. The use of ps type I fonts should avoid this problem.

% Modeling urban evolution with the diffusion of innovations
% A simple model of urban evolution based on innovation diffusion
% A model of urban evolution based on innovation diffusion
% A model of urban evolution and innovation diffusion
% A model of urban evolution based on the diffusion of innovations

\title{A model of urban evolution based on innovation diffusion}
\author{Juste Raimbault$^{1,2,3}$\\
\mbox{}\\
$^1$Center for Advanced Spatial Analysis, University College London\\
$^2$UPS CNRS 3611 ISC-PIF\\
$^3$UMR CNRS 8504 G{\'e}ographie-cit{\'e}s\\
juste.raimbault@polytechnique.edu} % email of corresponding author

% For several authors from the same institution use the same number to
% refer to one address.
%
% If the names do not fit well on one line use
%         Author 1, Author 2 ... \\ {\Large\bf Author n} ...\\ ...
%
% If the title and author information do not fit in the area
% allocated, place \setlength\titlebox{<new height>} after the
% \documentclass line where <new height> is 2.25in

\begin{document}
\maketitle

\begin{abstract}
% Abstract length should not exceed 250 words
The dynamics of urban systems can be understood from an evolutionary perspective, in some sense extending biological and cultural evolution. Models for systems of cities implementing elementary evolutionary processes remain however to be investigated. We propose here such a model for urban dynamics at the macroscopic scale, in which the diffusion of innovations between cities captures transformation processes (mutations) and transmission processes (diffusion), using two coupled spatial interaction models. Explorations of the model on synthetic systems of cities show the role of spatial interaction and innovation diffusion ranges on measures of diversity and utility, and the existence of intermediate ranges yielding an optimal utility. Multi-objective optimization shows how the model produces a compromise between utility and diversity. This model paves the way towards more elaborated formalizations of urban evolution.
\end{abstract}

\section{Introduction}

% (i) interest to understand urban dynamics, link to ALife

Urban systems are complex as they combine technical artefacts with socio-economic dynamics at multiple temporal and spatial scales. An understanding of processes driving their dynamics is an important aspect for shaping sustainable policies, and a sustainable design and management of cities \citep{lobo2020urban}. Multiple disciplines and viewpoints have proposed such insights from a complexity perspective \citep{pumain2020conclusion}, and Artificial Life approaches have shown promising results to study urban systems, both through conceptual contributions such as interpreting cities through biological metaphors \citep{batty2009centenary}, but also through modeling and simulation for urban growth including cellular automata and evolutionary computation \citep{raimbault2020cities}.

% (ii) Urban Evolution / Cultural evolution

A stream of research tightly linked to ALife relates to concepts of \emph{Urban Evolution}. These build in a sense on Cultural Evolution \citep{mesoudi2001cultural}, which aims at understanding changes in social knowledge as evolutionary processes involving replication, mutation, selection. It bears similarities with biological evolution but is not reducible to it, and for example uses the concept of \emph{meme} as transmission units comparable to genes. Both biological and cultural evolution can be linked into common frameworks and models, implying the coupling of different timescales \citep{bull2000meme}. In terms of urban studies, the concept of urban evolution is less formalized than cultural evolution and can be understood in multiple ways. \cite{votsis2019urban} use the concept of Urban DNA to characterize morphological properties of cities such as population density or the role of the road network. Similarly, \cite{kaya2017urban} describe cities based on their morphological properties as a product of their dynamics. \cite{wu2011urban} link the parameters of a cellular automaton model of urban growth to intrinsic properties of urban regions, which should play a role in their overall evolution. \cite{d2014urban} propose an urban genetic code as the way agents cooperate and compete, cities emerging as dynamical equilibria from these interactions. In Economic Geography, the concept of co-evolution is used mostly for urban agents such as firms and stakeholders \citep{gong2019co}. At the macroscopic scale of urban systems, \cite{pumain2018evolutionary} proposes an evolutionary theory to study systems of cities as complex adaptive systems, in which interactions between cities play a crucial role. Corresponding simulation models were proposed with different thematic focus, for example by \cite{cottineau2015modular} with economic exchanges and by \cite{raimbault2020indirect} with infrastructure networks. In Urban Design, \cite{batty2009digital} introduce evolutionary computation to explore possible urban forms. There still however remains a lack of models which would operationalize the concept of urban evolution in a way close to its biological and cultural counterparts, i.e. capturing explicitly the fundamental processes of transmission and transformation within differentiating subsystems \citep{durham1991coevolution}.

% (iii) Innovation diffusion - in ALife studies ; - in urban systems
%\cite{Eletreby5664} diffusion/evolving

Besides, a central concept bridging ALife and evolution with the study of urban systems is the concept of innovation diffusion. Within artificially evolved systems, understanding how innovation emerges and how it diffuses in the population is essential \citep{bedau2000open}. One aspect of open-ended evolution are mechanisms endogeneously producing novelty. In the case of urban systems, \cite{hagerstrand1968innovation} had already highlighted the role of a hierarchical diffusion of innovation between cities in their trajectories. The aforementioned evolutionary urban theory suggests that innovation cycles and their hierarchical diffusion is a possible explanation for the properties of urban scaling laws \citep{pumain2006evolutionary}. Evolutionary economics also study regional systems of innovation, how it diffuses and thee potential existence of spatial spillovers \citep{uyarra2010evolutionary}. Thus, the diffusion of innovation is a privileged entry into urban evolutionary processes.

% (iv) contribution

This paper proposes to tackle the issue of modeling urban evolution by using innovation diffusion processes to capture elementary evolutionary processes. We investigate thus the question to what extent simple models of urban evolution integrating an urban genome can be used to simulate urban dynamics. Our contribution relies on the following points: (i) we describe a relatively simple model for systems of cities at the macroscopic scales, based on population growth and the diffusion of innovation between cities; (ii) we systematically explore this model on synthetic systems of cities to extract stylized facts on different indicators including global diversity and utility.

The rest of this paper is organized as follows: in the next section we give the model context and describe it formally. We then present numerical experiments, including internal statistical validation, exploration of the parameter space, and optimization using a multi-objective genetic algorithm. We finally discuss implications of our results and possible future extensions.

\section{Urban evolution model}

\subsection{Rationale}

The core idea of the model is to build on a concept of ``Urban DNA'' which would capture evolution processes as in biological evolution and cultural evolution, i.e. a kind of genome that cities would be exchanging and which would undergo mutation processes. A suitable candidate is to build on the concept of \emph{meme} introduced in the field of cultural evolution. However, several particularities must be stressed out when working with urban systems. First, they are multi-dimensional implying very different types of agents, including physical and technical artefacts (e.g. infrastructures) but also social and economical structures. A comprehensive urban genome would include these very different dimensions. Then, they are multi-scalar in time and space, meaning that evolution processes, if they exist, may occur at different paces and through different elementary carriers. Finally, they are embedded in the geographical space, what structures the way transmission and mutation can occur, but also can strongly change properties of underlying processes \citep{raimbault2019space}. We choose to focus on the last point, considering a model at the macroscopic scale where cities are agents and with a one-dimensional genome, but in which spatial interactions are crucial for the dynamics. More particularly, this model combines two spatial interaction models \citep{fotheringham1989spatial}.

Several models have been proposed to simulate the diffusion of innovation at microscopic and mesoscopic scales \citep{kiesling2012agent}. \cite{blommestein1987adoption} describe a model of innovation diffusion and urban dynamics with endogenous demand for innovations, but in which the spatial component only influences prices of innovations. \cite{deffuant2005individual} give an example of an elaborated model for adoption dynamics at the microscopic level. Effective channels for the diffusion of innovations are multiple, and can for example be urban firm linkages \citep{rozenblat2007firm}. \cite{pumain2017simpoplocal} describe a model based on innovation diffusion to explain the emergence of the first cities. \cite{favaro2011gibrat} study an urban growth model including innovation diffusion at the scale of the urban system. Our model builds on this last framework, extending and adapting it to an urban evolution context. It is however significantly different in the way population are updated deterministically and in the innovation process.

In our model, cities are characterized by their size in terms of population, and city size evolve following a spatial interaction model in which city attractivity is included. This attractivity is determined by how innovative cities are. Transformation processes are included as mutations, when random innovation appear in cities. Transmission processes (spatial crossover) are included by diffusing innovations between cities. Finally, to existence of subsystems in which evolution can occur is natural through the spatial aspect of the model, through which different regions in space may behave and evolve differently, possibly resulting in the emergence of co-evolution niches \citep{holland2012signals,raimbault2018co}.

\subsection{Model description}

% - urban genome as a matrix (one for each city)
% - rows: % of each innovation (to simplify, sums to 1), columns: dimension of "urban life": transportation, economic production, leisure, built environment, governance, etc.
% - mutations: new innovations: new row. "endogenous innovation": add new column (dimension that did not exist before: new info technologies, induce niches, new needs, new dimensions in themselves. too complicated, do not include this aspect for the sake of simplicity, consider "fundamental dimensions" only
% - interdependance between dimensions? -> should yield "technological lockdowns". Q: how to implement? SBM? additional correlation matrix/ parameter?
% - innovation utility matrix: stochastic, not necessarily strongly increasing (should be similar though)
% - crossovers is hierarchical spatial diffusion: spatial interaction model
% - ! no network evolving (too complicated
% - evolves pop following Favaro-Pumain. P(innov) = scaling (pop) / utility ? ! not pop, gdp? (ok for eco innovation and sectors - should be multi-dimensional indeed)

We now formally describe the urban evolution model. A number $N$ of cities are located in the geographical space, and described in time by their size $P_i (t)$ (which generally corresponds to population). The geography is captured with a distance matrix $d_{ij}$ between cities. We consider a one-dimensional technological space in which innovations can be introduced, indexed by their order of apparition $c$. Cities are then also characterized by their innovation profile $\delta_{c,i} (t) \in \left[0;1\right]$ which represents the proportion of population having adopted innovation $c$ in the city $i$ at time $t$. We assume exclusivity in the adoption of innovation, i.e. that we always have $\sum_c \delta_{c,i} (t) = 1$. This innovation profile corresponds to the urban genome. Each innovation has furthermore an utility $u_c >0$ which will influence its diffusion dynamics.

\paragraph{Model dynamics}

% $\rightarrow$ Work under Gibrat independence assumptions, i.e. $\Covb{P_i(t)}{P_j(t)}=0$. If $\vec{P}(t+1)=\mathbf{R}\cdot \vec{P}(t)$ where $\mathbf{R}$ is also independent, then $\Eb{\vec{P}(t+1)}=\Eb{\mathbf{R}}\cdot\Eb{\vec{P}}(t)$. Consider expectancies only (higher moments computable similarly)
%$\rightarrow$ With $\vec{\mu}(t)=\Eb{\vec{P}(t)}$, we generalize this approach by taking $\vec{\mu}(t+1)=f(\vec{\mu}(t))$
%Let $\vec{\mu}(t)=\Eb{\vec{P}(t)}$ cities population and $(d_{ij})$ distance matrix
%Model specified by
%\[
%f(\vec{\mu}) = r_0\cdot \mathbf{Id}\cdot \vec{\mu} + \mathbf{G}\cdot \mathbf{1}
%\]

Starting from an initial configuration, the simulation model is iterated in time with the following steps at each time tick:

\begin{enumerate}
	\item The crossover between urban genomes relies on spatial processes of innovation diffusion. This means that existing innovations are propagated between cities following a spatial interaction model given by

\begin{equation}
\delta_{c,i,t} = \frac{\sum_j p_{c,j,t-1}^{\frac{1}{u_c}} \cdot \exp{(-\frac{d_{ij}}{d_I})}}{\sum_c \sum_j p_{c,j,t-1}^{\frac{1}{u_c}} \cdot \exp{(-\frac{d_{ij}}{d_I})}}
\end{equation}

where $p_{c,i,t} = \delta_{c,i,t} \cdot \frac{P_{i}(t)}{\sum_k P_k (t)}$ is the share of total population adopting $c$ in the city $i$ at time $t$; $d_I$ is the characteristic distance in the spatial innovation model, for which an increase will correspond to a broader spatial diffusion. As the population share is between 0 and 1, higher utilities will effectively diffuse faster as the exponent $1/u_c$ is used. In other terms, the inverse of the utility is the coefficient of population in the underlying spatial interaction model, and using an exponent is more relevant in that sense.

	\item The sizes of cities evolve according to their performance in terms of innovation, i.e. more innovative cities are more attractive. This stage is also deterministic, following  with $P_i(t) - P_i(t-1) = w_I\cdot \sum_j \frac{V_{ij}}{<V_{ij}>}$ where $w_I$ is a fixed growth rate fixed to $w_I = 0.005$ in experiments, and where the interaction potential is defined by
\begin{equation}
V_{ij}= \frac{P_{i}(t-1) \cdot P_{j}(t-1)}{(\sum_k P_k(t-1))^2} \cdot \exp{\left(-\frac{d_{ij}}{d_G} \cdot \prod_c \delta_{c,i,t}^{\phi_{c,t}}\right)}
\end{equation}
where $\phi_{c,t} = \sum_i \delta_{i,c,t}\cdot P_i(t-1) /\sum_{i,c} \delta_{i,c,t}\cdot P_{i}(t-1)$ is the macroscopic adoption level (globally more adopted innovations will have a higher attractivity); $d_G$ is the characteristic distance for this second spatial interaction model. This model for population dynamics was proposed in such a setting by \citep{raimbault2020indirect} without the additional innovation attractivity term.

 \item Finally, mutations in the urban genome (transformation process) corresponds to the introduction of new innovations. Each city has an innovation activity $\beta$ independent of its size, which corresponds to an intrinsic mutation rate; and will have a probability to innovate function of its size, similar to a Gibrat model \citep{pumain2012theorie}, as $\beta \cdot \left(P_i (t) / \max_k P_k (t)\right)^{\alpha_I}$ where $\alpha_I$ is a hierarchy exponent biasing the innovation towards larger cities. The utility for a new innovation is drawn stochastically, following either a normal or log-normal distribution, such that (i) its average corresponds to the current mean of utilities for existing innovations, and (ii) its standard deviation is a fixed parameter $\sigma_U$, which allows controlling the ``disruptivity'' of innovations. Urban genomes are adapted such that the new innovation as an initial penetration rate $r_0$ in the innovative city (previous innovation shares are rescaled for this city).
\end{enumerate}

The urban evolution is stopped after a fixed number of steps $t_f$.

\paragraph{Synthetic setup}

Although \cite{raimbault:halshs-01880492} proposed to apply the model of \cite{favaro2011gibrat} on real systems of cities, we focus our model study here on synthetic systems of cities, in order to isolate intrinsic effects independently of geographical contingencies. Cities are located randomly in an uniform square space. City sizes follow a Zipf law with exponent $\alpha_0 = 1$, i.e. such that $P_i (0) = \frac{P_{max}}{i^{\alpha_0}}$ \citep{pumain2012theorie}. We do not modify this initial hierarchy in our experiments. Distances between cities $d_{ij}$ are computed as euclidian distances. We take $N=30$ cities and $t_f = 50$ in our experiments, corresponding to a regional or national system of cities, on timescales of the order of a century.

\paragraph{Indicators}

% - Indicators: innovation diversity (on cities profiles); overall utility (optimal config as a function of lambda - non trivial?) - with path-dep/correlations?; existence of cycles?
%.   compute utility for one city: along each dimension, weight utility of innovations. then aggregate? ~

The behavior of the model is quantified through indicators at the macroscopic scale. These should allow extracting stylized facts from the model exploration. We consider the following indicators:
\begin{itemize}
	\item Average diversity, defined as an Herfindhal diversity index over innovation shares across cities, averaged in time as
	\begin{equation}
		D = \frac{1}{t_f + 1} \sum_{t=0}^{t_f} \left(1 - \sum_{i,c} \left(p_{c,i,t}\right)^2 \right)	
	\end{equation}
	Note that other diversity indices could be applied, or a similar be computed on macroscopic adoption shares or within each city, and would yield different results. This one has the advantage to combine the diversity within and across cities.
	\item Average utility, given by the weighted average of innovation utilities, averaged in time, as
	\begin{equation}
		U = \frac{1}{t_f + 1} \sum_{t=0}^{t_f} \sum_{i,c} \delta_{c,i,t} u_c 
	\end{equation}
	\item Innovativity, given by the average number of innovations per city and unit of time $I = \frac{\max c}{N\cdot (t_f + 1)}$.
	\item To quantify the trajectories of city populations, many indicators can be used \citep{raimbault2020unveiling}. We choose simply to quantify the level of hierarchy at final time, estimated by fitting the rank-size law on populations using an Ordinary Least Squares on logarithms. This captures if the urban systems has become more unequal in terms of population balance.
\end{itemize}

%%%%%%%%%%%%
\begin{figure*}[t]
	\centering
	\includegraphics[width=0.49\linewidth]{averageDiversity-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_distriblog-normal.png}
	\includegraphics[width=0.49\linewidth]{averageInnovation-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_distriblog-normal.png}
	\includegraphics[width=0.49\linewidth]{averageUtility-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_distriblog-normal.png}
	\includegraphics[width=0.49\linewidth]{finalHierarchy-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_distriblog-normal.png}
	\caption{Values of indicators obtained with the grid model exploration. Each plot give each indicator among diversity $D$, utility $U$, innovation $I$ and population hierarchy $\alpha_P$, at fixed parameters $\alpha_I = 1$, $\sigma_U = 1$ and for the log-normal utility law. Indicators are plotted as a function of gravity interaction range $d_G$, for varying values of innovation diffusion range $d_I$ (color). Sub-panels show varying values of mutation rate $\beta$ (rows) and of early adoption rate $r_0$ (columns).\label{fig:exploration}}
\end{figure*}
%%%%%%%%%%%%

\paragraph{Model parameters}

Parameter explored in experiments are the spatial interaction parameters $d_G,d_I \in \left[0;2\right]$ which correspond to the crossover mechanism; and the innovation parameter which correspond to the mutation mechanism: innovation rate $\beta \in \left[0;1\right]$, innovation hierarchy $\alpha_I \in \left[0;2\right]$, innovation utility standard deviation $\sigma_U \in \left[ 0.7;2\right]$ (the lower bound is a constraint for the existence of the log-normal with log mean 0), initial penetration rate $r_0 \in \left[0.1;0.9\right]$, and type of innovation utility as normal or log-normal.

\section{Results}

The model is implemented in \texttt{scala} and integrated into the OpenMOLE software \citep{reuillon2013openmole} for numerical experiments. OpenMOLE allows embedding models in any language as black boxes, provides a transparent access to high performance computing infrastructures, and model exploration and validation methods such as sensitivity analysis, design of experiments, and calibration methods. Experiments are designed through workflows using a Domain Specific Language \citep{passerat2017reproducible}.

Source code of the model and analysis, and results are available on the open git repository of the project at \url{https://github.com/JusteRaimbault/UrbanEvolution}. Simulation data files are available on the dataverse repository at \url{https://doi.org/10.7910/DVN/Q5GKZ0}.

\subsection{Internal validation}

% summary(reldistance("meanAverageDiversity","sdAverageDiversity"))
%   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
% 0.0000  0.6035  1.2685  1.4709  2.1557  6.1199 
%> summary(reldistance("medianAverageDiversity","sdAverageDiversity"))
%   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
% 0.0000  0.5733  1.2115  1.4425  2.0941  6.5835 
% summary(reldistance("meanAverageInnovation","sdAverageInnovation"))
%   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
%  0.000   1.839   4.136   4.768   7.011  15.799 
%> summary(reldistance("medianAverageInnovation","sdAverageInnovation"))
%   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
%  0.000   1.841   4.176   4.783   7.100  15.891 
%summary(reldistance("meanAverageUtility","sdAverageUtility"))
%   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
%  0.000   0.509   1.103   1.320   1.923   6.350 
%> summary(reldistance("medianAverageUtility","sdAverageUtility"))
%   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
% 0.0000  0.4942  1.0857  1.3013  1.8863  6.3479 
%summary(reldistance("meanFinalHierarchy","sdFinalHierarchy"))
%   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
%  0.000   2.488   6.066  11.424  14.133  96.805 
%> summary(reldistance("medianFinalHierarchy","sdFinalHierarchy"))
%   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
%  0.000   2.508   6.074  11.449  14.171  96.789 
% idea: on this "relative dist": index similar to Saltelli to tell distinguishability across the all parameter space?

As the model is stochastic through the mutation process, first experiments are needed to assess the ``internal validity'' of the model, i.e. to what extent studied indicators are robust to noise and how much stochastic repetitions are needed to significantly distinguish between central values of distributions for different parameter points. We sample thus 100 parameter points using a Latin Hypercube Sampling, and run 1000 model repetitions for each point, in order to estimate the statistical properties across different points in the parameter space.

The sharpe ratios estimated on repetitions as sample standard deviation relative to absolute sample mean have high values for all indicators (min. 3.9 and median 12.1 for diversity; min. 3.0 and median 6.2 for innovation; min. 1.7 and median 3.6 for utility; min. 26 and median 257.3 for population hierarchy), what means that stochastic noise is not an issue for interpreting indicator values.

We also study the distance between points averages relative to their standard deviation, in order to know the significance in comparing averages. This distance is defined by $\Delta_{ij} = 2\frac{\left|\mu_i - \mu_j \right|}{\sigma_i + \sigma_j}$ if $\mu_i$ are estimated central values (mean or median) and $\sigma_i$ estimated standard deviations. These distances have an average and median larger than one for all couples, and a first quartile larger than one for innovation and population hierarchy, while the first quartile is 0.5 for utility and 0.6 for diversity, what means that these two last indicators need more statistical precision to be interpreted. As in the case of normal distributions, a 95\% confidence interval of size $\sigma_2$ is achieved for $n\simeq 64$ runs, a number of 50 runs is satisfying for further experiments.

\subsection{Model exploration}

We then explore a grid of the parameter space consisting of 23,168 parameter points, with a finer step on spatial parameters $d_G$ and $d_I$, and with 50 model replications for each parameter point. We find that the type of distribution for utility has small effects for all indicators and across most parameter values, but particular cases are more interesting in the log-normal case, which we comment further. Similarly, the effect of innovation hierarchy $\alpha_I$ and of innovation utility standard deviation $\sigma_U$ do not induce strong qualitative differences, so we discuss model behavior obtained at $\alpha_I = 1$ and $\sigma_U = 1$.

The Fig.~\ref{fig:exploration} shows the behavior of indicators with remaining parameters varying, namely spatial interaction parameters $d_G$ and $d_I$, mutation rate $\beta$ and innovation adoption $r_0$. Regarding the behavior of diversity (top left panel of Fig.~\ref{fig:exploration}), we find that it always increase for larger population growth spatial interaction distances, meaning that broader population exchanges will induce a higher diversity in innovations, probably through the higher innovation dynamics induced. This increase however reaches a plateau around half of the width of the world ($d_G = 0.5$) which is more obvious for higher values of mutation rates (bottom raw): higher intrinsic innovation dampens the role of spatial interactions. An increase of innovation interaction distance $d_I$ on the contrary monotonously decreases diversity, consistently with the idea of local innovation niches which can be overridden by competing innovations of higher utility coming from further away. Finally, mutation rate has a strong quantitative impact on total diversity as expected, while higher early adoption rate will change diversity behavior only when mutation rates are also high.

Two indicators which behavior is less rich are innovation and population hierarchy (top right and bottom right panels of Fig.~\ref{fig:exploration}), which could be expected as population dynamics are deterministic and strongly related to space, while innovation is directly related to population. We find that these increase both with $d_G$ and $d_I$, meaning that more interactions and more diffusion will foster local innovation. Regarding the final distribution of population, it means that systems where interaction are more global will be less unequal. This is however not always the case in such urban dynamics model as explored by \cite{raimbault2020hierarchy}. The mutation rate $\beta$ here only fosters the role of small $d_I$, yielding higher hierarchies and less innovations for these. The effect of initial adoption rate is small for these indicators. Note that although they appear highly correlated in this region of the parameter space, they are not the indicators with the highest correlation as shown further (see Fig.~\ref{fig:cormat}).

Finally, the indicator with a more interesting behavior is the utility (bottom left panel of Fig.~\ref{fig:exploration}). For low values of mutation rates, with find a piecewise linear behavior as a function of $d_G$ and small effects of $d_I$, meaning that low innovation settings yield regimes where broader interactions are systematically desirable for the all system. However, when $\beta$ is higher, we witness a maximum of utility as a function of $d_G$, consistent across different values of $d_I$ and of $r_0$. This corresponds to an intermediate regional regime where local innovation regimes are more beneficial than a global integration. This peak is the strongest when innovation diffuse at broader range, what may mean that this regime is due to the emergence of regional ensemble of comparable competitivity and thus a higher chance of high utility innovation, while a globalized system would concentrate on a single dominating city and be overall less performant.

% correlations
%%%%%%%%%%%%
\begin{figure}[t]
	\centering
	\includegraphics[width=\linewidth,trim={6cm 6cm 0 6cm}]{corrmat_indics.png}
	\caption{Correlation matrix between indicators, estimated on the full grid experiment. Confidence intervals for correlations estimated with the Fisher method are tiny and not distinguishable in the plot.\label{fig:cormat}}
\end{figure}
%%%%%%%%%%%%

% summary(prcomp(m))
%Importance of components:
%                          PC1    PC2    PC3     PC4
%Standard deviation     0.2925 0.2259 0.1560 0.11976
%Proportion of Variance 0.4882 0.2910 0.1389 0.08183
%Cumulative Proportion  0.4882 0.7793 0.9182 1.00000

We also compute the correlation matrix between indicators across the full grid experiment. This shows that correlations that one could visually extrapolate from studying a part of the parameter space as commented in Fig.~\ref{fig:exploration} do not correspond to the actual correlations on the broader parameter space. Note that local correlation matrices could be estimated for a more thorough discussion. We show the correlation matrix in Fig.~\ref{fig:cormat}. Confidence intervals estimated with the Fisher method are of negligible width compared to correlation values. We find that diversity is finally the indicator with highest correlations. The correlation between innovation and population is low, meaning that the similar curves observed before are a particular case. Innovation and utility have a correlation lower than 0.5, and are thus rather independent. Regarding the effective dimension of the indicator space, a principal component analysis on normalized indicators gives 48\% of variance on the first component, 77\% of cumulated variance on the second and 91\% on the third, confirming that even if correlation exist the behavior of indicators are rather independent.

\subsection{Model optimization}

%%%%%%%%%%%%
\begin{figure}[t]
	\centering
	\includegraphics[width=\linewidth]{paretoDiversity-Utility_colorinnovationDecay.png}
	\caption{Pareto front for the contradictory indicator of utility and diversity obtained with the NSGA2 algorithm. Point color gives innovation interaction range $d_I$ while point size gives the number of stochastic samples.\label{fig:pareto}}
\end{figure}
%%%%%%%%%%%%

The last experiment we perform is the application of a multi-objective optimization procedure to the model. More particularly, one could want to optimize simultaneously the global utility but also the diversity to ensure a certain resilience in the urban system. We apply thus a bi-objective genetic algorithm on the model, trying to maximize simultaneously utility and diversity. The algorithm is the NSGA2 algorithm \citep{deb2002fast}, which we run for 10,000 generations with a population of 200 individuals on a Island distribution scheme. The algorithm is integrated into the OpenMOLE software, and the stochastic aspect is internally tackled by using an embedding strategy, i.e. by adding the number of repetitions as an additional objective to find compromise points between the number of run needed and their statistical accuracy.

We show in Fig.~\ref{fig:pareto} the obtained points, which are close to a Pareto front. We find two regimes, the first corresponding to the upper points in the plot for higher utilities but lower diversities, which should correspond to the intermediate optimal regimes identified before, and in which some kind of linear compromize between utility and diversity exists: increasing global utility is done at the cost of reducing diversity. Within this first regime, two very different parameter setting coexist, one with high innovation diffusion (light blue, points with higher utility), the other in dark blue with a more local innovation diffusion. This means that reducing the span of innovation diffusion will increase diversity as one could expect. A second part of the Pareto front, obtained after a sharp transition, allows increasing the diversity but at the price of a much lower utility (points on the bottom right). These points are all close to equivalent, and one may prefer in the compromise the points just before the transition which correspond to a local diffusion setting. Thus, local diffusion correspond to intermediate compromises, while broad diffusion corresponds to extremes in the Pareto front. In a nutshell, this optimization exercise is interesting both to show how the model produces compromises, but also how it could be used in practice for innovation policies applied to systems of cities.

\section{Discussion}

Our simple formulation of an urban genome and associated evolutionary processes, implemented by the diffusion of innovation, already capture complex urban dynamics, as shown for example by the emergence of an intermediate optimal regime corresponding to local innovation niches which diffuse far. Although much more empirical work would be needed to compare these stylized facts to real world settings, our model suggests in this particular setting that too much integration is not always optimal, what corresponds to the theoretical fact that complex systems are generally modular at different scales \citep{ethiraj2004modularity}. More generally, regarding the implications of our results for possible formalizations and theories of urban evolution which would closely build on biological and cultural evolution by extending them, we have demonstrated how a particular instance of an urban genome can be used to simulated urban dynamics, including fundamental processes needed to effectively have evolution. To what extent this approach relates to existing approaches of urban evolution which use other definitions remains to be investigated.

Several extensions and applications would be possible to this first model exploration. First, the innovation space in our model remained unidimensional. \citep{hidalgo2007product} show in terms of industrial production by countries that these industrial spaces are highly dimensional in terms of product produced and interdependencies between countries and types of production, increasing the path-dependency in economic trajectories. The investigation of patent data from a semantic viewpoint also shows the highly dimensional nature of technological innovation \citep{bergeaud2017classifying}. A direct extension of our model would consist in having a matrix genome instead of a vector one. Innovations would occur across several dimensions which can correspond to industrial or technological domains, but also social and cultural innovation and innovations in terms of infrastructures which condition the way systems of cities evolve. Each dimension should have particular innovation rules depending on its nature, and dependencies across dimensions could be introduced, implementing the possible emergence of technological co-evolution niches beside the spatial co-evolution niches.

Second, applying the model to real system of cities, both in terms of initial parametrization and of empirical laws for innovation processes, would allow bringing this approach closer to possible policy applications. \cite{raimbault:halshs-01880492} benchmarked several models of urban growth based on interactions between cities, and integrated the model of \cite{favaro2011gibrat} on which this work was based. It however only included accurate initial populations and distance matrix. This application to real systems of cities also allows testing the performance of the model in predicting possible trajectories for these systems.

Finally, our approach was rather restricted in the sense that even a broad geographical range is taken into account, a single ontological scale is included in the model, i.e. the macroscopic scale since cities are the basic agents. Similarly, a single temporal scale was included, although two dynamics of spatial interaction and innovation diffusion are effectively combined. A multi-scale approach of urban evolution would be necessary to fully account for the complexity of these systems. \cite{raimbault:halshs-02351722} introduced a multi-scalar model for population growth with upward and downward strong feedbacks between the mesoscopic scale (urban form) and the macroscopic scale (spatial interaction model). Adapting this approach in the case of urban evolution through innovation diffusion would be an interesting potential development.

\section{Conclusion}

We have introduced a simple model of urban evolution integrating effectively the evolutionary processes of transmission, transformation and evoluting sub-systems, through innovation diffusion and spatial interactions. Model exploration yield complex behavior while multi-objective optimization shows the potentiality for the model to produce compromises between utility and diversity in the system of cities. This work is thus a first step towards more elaborated models of urban evolution.

\bigskip

\section{Acknowledgements}

Results obtained in this paper were computed on the vo.complex-system.eu virtual organization of the European Grid Infrastructure ( http://www.egi.eu ). We thank the European Grid Infrastructure and its supporting National Grid Initiatives (France-Grilles in particular) for providing the technical support and infrastructure. This work was funded by the Urban Dynamics Lab grant EPSRC EP/M023583/1.

\bigskip

\footnotesize
%\bibliographystyle{apalike}
%\bibliography{biblio} % replace by the name of your .bib file

\begin{thebibliography}{}

\bibitem[Batty, 2009]{batty2009digital}
Batty, M. (2009).
\newblock A digital breeder for designing cities.
\newblock {\em Architectural Design}, 79(4):46--49.

\bibitem[Batty and Marshall, 2009]{batty2009centenary}
Batty, M. and Marshall, S. (2009).
\newblock Centenary paper: The evolution of cities: Geddes, abercrombie and the
  new physicalism.
\newblock {\em Town Planning Review}, 80(6):551--574.

\bibitem[Bedau et~al., 2000]{bedau2000open}
Bedau, M.~A., McCaskill, J.~S., Packard, N.~H., Rasmussen, S., Adami, C.,
  Green, D.~G., Ikegami, T., Kaneko, K., and Ray, T.~S. (2000).
\newblock Open problems in artificial life.
\newblock {\em Artificial life}, 6(4):363--376.

\bibitem[Bergeaud et~al., 2017]{bergeaud2017classifying}
Bergeaud, A., Potiron, Y., and Raimbault, J. (2017).
\newblock Classifying patents based on their semantic content.
\newblock {\em PloS one}, 12(4):e0176310.

\bibitem[Blommestein and Nijkamp, 1987]{blommestein1987adoption}
Blommestein, H. and Nijkamp, P. (1987).
\newblock Adoption and diffusion of innovations and the evolution of spatial
  systems.
\newblock In {\em Economic Evolution and Structural Adjustment}, pages
  368--380. Springer.

\bibitem[Bull et~al., 2000]{bull2000meme}
Bull, L., Holland, O., and Blackmore, S. (2000).
\newblock On meme--gene coevolution.
\newblock {\em Artificial life}, 6(3):227--235.

\bibitem[Cottineau et~al., 2015]{cottineau2015modular}
Cottineau, C., Reuillon, R., Chapron, P., Rey-Coyrehourcq, S., and Pumain, D.
  (2015).
\newblock A modular modelling framework for hypotheses testing in the
  simulation of urbanisation.
\newblock {\em Systems}, 3(4):348--377.

\bibitem[D'Acci, 2014]{d2014urban}
D'Acci, L. (2014).
\newblock Urban dna for cities evolutions. cities as physical expression of
  dynamic equilibriums between competitive and cooperative forces.
\newblock {\em arXiv preprint arXiv:1408.2874}.

\bibitem[Deb et~al., 2002]{deb2002fast}
Deb, K., Pratap, A., Agarwal, S., and Meyarivan, T. (2002).
\newblock A fast and elitist multiobjective genetic algorithm: Nsga-ii.
\newblock {\em IEEE transactions on evolutionary computation}, 6(2):182--197.

\bibitem[Deffuant et~al., 2005]{deffuant2005individual}
Deffuant, G., Huet, S., and Amblard, F. (2005).
\newblock An individual-based model of innovation diffusion mixing social value
  and individual benefit.
\newblock {\em American Journal of Sociology}, 110(4):1041--1069.

\bibitem[Durham, 1991]{durham1991coevolution}
Durham, W.~H. (1991).
\newblock {\em Coevolution: Genes, culture, and human diversity}.
\newblock Stanford University Press.

\bibitem[Ethiraj and Levinthal, 2004]{ethiraj2004modularity}
Ethiraj, S.~K. and Levinthal, D. (2004).
\newblock Modularity and innovation in complex systems.
\newblock {\em Management science}, 50(2):159--173.

\bibitem[Favaro and Pumain, 2011]{favaro2011gibrat}
Favaro, J.-M. and Pumain, D. (2011).
\newblock Gibrat revisited: An urban growth model incorporating spatial
  interaction and innovation cycles. 再评吉尔布瑞特:
  顾及空间相互作用和创新周期的城市增长模型.
\newblock {\em Geographical Analysis}, 43(3):261--286.

\bibitem[Fotheringham and O'Kelly, 1989]{fotheringham1989spatial}
Fotheringham, A.~S. and O'Kelly, M.~E. (1989).
\newblock {\em Spatial interaction models: formulations and applications},
  volume~1.
\newblock Kluwer Academic Publishers Dordrecht.

\bibitem[Gong and Hassink, 2019]{gong2019co}
Gong, H. and Hassink, R. (2019).
\newblock Co-evolution in contemporary economic geography: Towards a
  theoretical framework.
\newblock {\em Regional Studies}, 53(9):1344--1355.

\bibitem[Hagerstrand et~al., 1968]{hagerstrand1968innovation}
Hagerstrand, T. et~al. (1968).
\newblock Innovation diffusion as a spatial process.
\newblock {\em Innovation diffusion as a spatial process.}

\bibitem[Hidalgo et~al., 2007]{hidalgo2007product}
Hidalgo, C.~A., Klinger, B., Barab{\'a}si, A.-L., and Hausmann, R. (2007).
\newblock The product space conditions the development of nations.
\newblock {\em Science}, 317(5837):482--487.

\bibitem[Holland, 2012]{holland2012signals}
Holland, J.~H. (2012).
\newblock {\em Signals and boundaries: Building blocks for complex adaptive
  systems}.
\newblock Mit Press.

\bibitem[Kaya and B{\"o}len, 2017]{kaya2017urban}
Kaya, H.~S. and B{\"o}len, F. (2017).
\newblock Urban dna: morphogenetic analysis of urban pattern.

\bibitem[Kiesling et~al., 2012]{kiesling2012agent}
Kiesling, E., G{\"u}nther, M., Stummer, C., and Wakolbinger, L.~M. (2012).
\newblock Agent-based simulation of innovation diffusion: a review.
\newblock {\em Central European Journal of Operations Research},
  20(2):183--230.

\bibitem[Lobo et~al., 2020]{lobo2020urban}
Lobo, J., Alberti, M., Allen-Dumas, M., Arcaute, E., Barthelemy, M.,
  Bojorquez~Tapia, L.~A., Brail, S., Bettencourt, L., Beukes, A., Chen, W.-Q.,
  et~al. (2020).
\newblock Urban science: Integrated theory from the first cities to sustainable
  metropolises.

\bibitem[Mesoudi, 2001]{mesoudi2001cultural}
Mesoudi, A. (2001).
\newblock Cultural evolution.
\newblock {\em eLS}, pages 1--8.

\bibitem[Passerat-Palmbach et~al., 2017]{passerat2017reproducible}
Passerat-Palmbach, J., Reuillon, R., Leclaire, M., Makropoulos, A., Robinson,
  E.~C., Parisot, S., and Rueckert, D. (2017).
\newblock Reproducible large-scale neuroimaging studies with the openmole
  workflow management system.
\newblock {\em Frontiers in neuroinformatics}, 11:21.

\bibitem[Pumain, 2018]{pumain2018evolutionary}
Pumain, D. (2018).
\newblock An evolutionary theory of urban systems.
\newblock In {\em International and transnational perspectives on urban
  systems}, pages 3--18. Springer.

\bibitem[Pumain et~al., 2012]{pumain2012theorie}
Pumain, D. et~al. (2012).
\newblock Une th{\'e}orie g{\'e}ographique pour la loi de zipf.
\newblock {\em R{\'e}gion et d{\'e}veloppement}, 36(31-54):125--150.

\bibitem[Pumain et~al., 2006]{pumain2006evolutionary}
Pumain, D., Paulus, F., Vacchiani-Marcuzzo, C., and Lobo, J. (2006).
\newblock An evolutionary theory for interpreting urban scaling laws.
\newblock {\em Cybergeo: European Journal of Geography}.

\bibitem[Pumain and Raimbault, 2020]{pumain2020conclusion}
Pumain, D. and Raimbault, J. (2020).
\newblock Conclusion: Perspectives on urban theories.
\newblock In {\em Theories and Models of Urbanization}, pages 303--330.
  Springer.

\bibitem[Pumain and Reuillon, 2017]{pumain2017simpoplocal}
Pumain, D. and Reuillon, R. (2017).
\newblock The simpoplocal model.
\newblock In {\em Urban Dynamics and Simulation Models}, pages 21--35.
  Springer.

\bibitem[Raimbault, 2018a]{raimbault:halshs-01880492}
Raimbault, J. (2018a).
\newblock {A systematic comparison of interaction models for systems of
  cities}.
\newblock In {\em {Conference on Complex Systems 2018}}, Thessaloniki, Greece.

\bibitem[Raimbault, 2018b]{raimbault2018co}
Raimbault, J. (2018b).
\newblock Co-evolution and morphogenetic systems.
\newblock {\em arXiv preprint arXiv:1803.11457}.

\bibitem[Raimbault, 2019]{raimbault:halshs-02351722}
Raimbault, J. (2019).
\newblock {A multi-scalar model for system of cities}.
\newblock In {\em {Conference on Complex Systems 2019}}, Singapore, Singapore.

\bibitem[Raimbault, 2020a]{raimbault2020cities}
Raimbault, J. (2020a).
\newblock Cities as they could be: Artificial life and urban systems.
\newblock {\em arXiv preprint arXiv:2002.12926}.

\bibitem[Raimbault, 2020b]{raimbault2020hierarchy}
Raimbault, J. (2020b).
\newblock Hierarchy and co-evolution processes in urban systems.
\newblock {\em arXiv preprint arXiv:2001.11989}.

\bibitem[Raimbault, 2020c]{raimbault2020indirect}
Raimbault, J. (2020c).
\newblock Indirect evidence of network effects in a system of cities.
\newblock {\em Environment and Planning B: Urban Analytics and City Science},
  47(1):138--155.

\bibitem[Raimbault, 2020d]{raimbault2020unveiling}
Raimbault, J. (2020d).
\newblock Unveiling co-evolutionary patterns in systems of cities: a systematic
  exploration of the simpopnet model.
\newblock In {\em Theories and Models of Urbanization}, pages 261--278.
  Springer.

\bibitem[Raimbault et~al., 2019]{raimbault2019space}
Raimbault, J., Cottineau, C., Le~Texier, M., Le~Nechet, F., and Reuillon, R.
  (2019).
\newblock Space matters: Extending sensitivity analysis to initial spatial
  conditions in geosimulation models.
\newblock {\em Journal of Artificial Societies and Social Simulation},
  22(4):10.

\bibitem[Reuillon et~al., 2013]{reuillon2013openmole}
Reuillon, R., Leclaire, M., and Rey-Coyrehourcq, S. (2013).
\newblock Openmole, a workflow engine specifically tailored for the distributed
  exploration of simulation models.
\newblock {\em Future Generation Computer Systems}, 29(8):1981--1990.

\bibitem[Rozenblat and Pumain, 2007]{rozenblat2007firm}
Rozenblat, C. and Pumain, D. (2007).
\newblock Firm linkages, innovation and the evolution of urban systems.
\newblock {\em Cities in globalization: Practices, policies, theories}, pages
  130--156.

\bibitem[Uyarra, 2010]{uyarra2010evolutionary}
Uyarra, E. (2010).
\newblock What is evolutionary about ‘regional systems of innovation’?
  implications for regional policy.
\newblock {\em Journal of evolutionary economics}, 20(1):115.

\bibitem[Votsis and Haavisto, 2019]{votsis2019urban}
Votsis, A. and Haavisto, R. (2019).
\newblock Urban dna and sustainable cities: A multi-city comparison.
\newblock {\em Frontiers in Environmental Science}, 7:4.

\bibitem[Wu and Silva, 2011]{wu2011urban}
Wu, N. and Silva, E.~A. (2011).
\newblock Urban dna: exploring the biological metaphor of urban evolution with
  dg-abc model.
\newblock In {\em Proceedings The 14th AGILE International Conference on
  Geographic Information Science. Utrecht University}.

\end{thebibliography}

\end{document}

%\begin{figure}[t]
%\begin{center}
%\includegraphics[width=2.1in,angle=-90]{fig1.eps}
%\caption{``Energies'' (inferiorities) of strings in a first-order
%  phase transition with latent heat $\Delta\epsilon$.}
%\label{fig1}
%\end{center}
%\end{figure}

%
%\begin{table}[h]
%\center{
%\begin{tabular}{|c|c|c|c|}\hline
%Name & Result & Bonus $b_i$ & Difficulty\\ \hline\hline
%Echo & I/O   & 1 & --\\
%Not  & $\neg A$ & 2 & 1 \\
%Nand & $\neg(A\wedge B)$ & 2 & 1 \\
%Not Or & $\neg A \vee B$ & 3 & 2 \\
%And  &  $ A \wedge B $   & 3 & 2 \\
%Or   &  $ A \vee B $     & 4 & 3 \\
%And Not & $A\wedge\neg B$& 4 & 3 \\
%Nor  & $\neg(A\vee B)$   & 5 & 4 \\
%Xor  & $ A\ {\rm xor}\ B$ &   6 & 4 \\
%Equals &$\neg(A\ {\rm xor}\ B)$&6& 4 \\ \hline
%\end{tabular}
%}
%\vskip 0.25cm
%\caption{Logical calculations on random inputs $A$ and $B$ rewarded,
%bonuses, and difficulty (in minimum number of {\tt nand} instructions
%required). Bonuses $b_i$ increase the speed of a CPU by a factor
%$\nu_i=1+2^{b_i-3}$.}
%\end{table}
%

### 2020-04-30 Docs/Papers/EvolutionInnovation/paper/paper.tex
papers

--- 2020-04-30 Docs/Papers/EvolutionInnovation/arxiv/paper.tex
@@
 \end{abstract}
+
+% Urban Systems; Urban Evolution; Innovation Diffusion
 
@@
 	\centering
-	\includegraphics[width=0.49\linewidth]{averageDiversity-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_distriblog-normal.png}
-	\includegraphics[width=0.49\linewidth]{averageInnovation-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_distriblog-normal.png}
-	\includegraphics[width=0.49\linewidth]{averageUtility-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_distriblog-normal.png}
-	\includegraphics[width=0.49\linewidth]{finalHierarchy-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_distriblog-normal.png}
+	\includegraphics[width=0.49\linewidth]{figures/averageDiversity-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_distriblog-normal.png}
+	\includegraphics[width=0.49\linewidth]{figures/averageInnovation-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_distriblog-normal.png}
+	\includegraphics[width=0.49\linewidth]{figures/averageUtility-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_distriblog-normal.png}
+	\includegraphics[width=0.49\linewidth]{figures/finalHierarchy-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_distriblog-normal.png}
 	\caption{Values of indicators obtained with the grid model exploration.
@@
 	\centering
-	\includegraphics[width=\linewidth,trim={6cm 6cm 0 6cm}]{corrmat_indics.png}
+	\includegraphics[width=\linewidth,trim={6cm 6cm 0 6cm}]{figures/corrmat_indics.png}
 	\caption{Correlation matrix between indicators, estimated on the full grid experiment.
@@
 	\centering
-	\includegraphics[width=\linewidth]{paretoDiversity-Utility_colorinnovationDecay.png}
+	\includegraphics[width=\linewidth]{figures/paretoDiversity-Utility_colorinnovationDecay.png}
 	\caption{Pareto front for the contradictory indicator of utility and diversity obtained with the NSGA2 algorithm.

### 2020-06-18 Docs/Papers/EvolutionInnovation/paper/paper.tex
calib_script
.

@@
 
+% Review 1 
+%  evaluation:	
+%2: (accept)
+%Promising work. For the future, it will be interesting to see how the model can reproduce observed properties of different cities.
+
+% Review 2
+%1: (weak accept)
+%This paper presents a model of urban evolution based on innovation diffusion.
+The paper describes a simple model for systems of cities at the macroscopic level using population growth and innovation diffusion, which is then explored on synthethic city systems.
+
+%The paper is well written and the results promising.
+It would have been very interesting to see how this would work for a model of a real city - there are an abundance of datasets and models (even cited by the author) that would not have been difficult to integrate.
+
+%In addition, it would be very useful to include a discussion on how modeling innovation at a microscopic scale (or in-between, eg at a level of suburbs) changes the results of the simulations.
+
+% Review 3
+
+% 2: (accept)
+%A model of urban evolution based on innovation diffusion
+
+%I think you have really interesting ideas in this article particularly the research methodology followed.
+I found the paper to be very comprehensive with an interesting presentation of findings.
+The objectives are clear and literature review well done.
+The supporting theory was put together partly based on empirical preliminary studies carried out however, there are few parameters used that need to be justified adequately.
+The general concerns are that the paper is too hard to read because of typos and number of assumptions that need to be verified and qualified as stated below.
+
+%Typos:
+%1. Evolutionary economics also study regional systems of innovation, how it diffuses and thee potential existence of spatial spillovers (Uyarra, 2010).
+%2. Finally, to existence of subsystems in which evolution can occur is natural through the spatial aspect of the model, ….
+%3. It is however significantly different in the way population are updated deterministically
+%4. Transformation processes are included as mutations, when random innovation appear in cities.
+%5. In our model, cities are characterized by their size in terms of population, and city size evolve following a spatial
+%6. Urban genomes are adapted such that the new innovation as an initial penetration rate ..
+%7. We choose simply to quantify the level of hierarchy at final time, estimated by fitting the rank-size law on populations using an Ordinary Least Squares on logarithms.
+%8. what means that stochastic noise is not an issue for interpreting indicator values … what means that these two last indicators need more statistical precision
+%9. we find that it always increase for larger population
+%10. it means that systems where i nteraction are more global will be less unequal. This is however not always the case in such urban dynamics model as
+%11. For low values of mutation rates, with find a piecewise linear
+%12.
+This peak is the strongest when innovation diffuse at broader range , what may mean that this regime is due to the emergence of regional ensemble of comparable competitivity and thus a higher chance of high utility innovation,
+%13. cumulated variance on the second and 91% on the third, confirming that even if correlation exist the behavior of
+%14. one may prefer in the compromise the points just before the transition which correspond to a local diffusion setting. Thus, local diffusion correspond to intermediate
+%15. urban genome can be used to simulated urban dynamics
+%16. model to real system …. and evoluting sub-systems
+%Other issues:
+%1. You need to deal with text formatting as some texts are being cut out at the end of the line (observed on paragraph 1 on page 5).
+%2. Use of compromise and compromize in the same article, (kind of linear compromize between utility … prefer in the compromise the …)
+%3. It is not clear why the growth rate is fixed to 0.005 and the cited paper (Raimbault, 2020c) does not justify the use of that value.
+%4.
+Fig 1 is a bit confusing, you will need to find a better way of presenting it than this.To enhance readability, the mutation rate and adoption rate should be clearly associated with those rows and columns you mentioned.
+As it is, it's hard to read.
+%5.
+The statement “This increase however reaches a plateau around half of the width of the world (dG = 0.5) which is more obvious for higher values of mutation rates (bottom raw) …” is also confusing as it is not clear which plotting it refers to as it does not appear to apply to all plottings.
+I don’t think ‘(bottom raw) ’ is correct either.
+%Your arguments are strong and I hope my comments will help to fine-tune your article to make it more readable. Good luck with your submission.
+
 \begin{document}
@@
 % Urban Systems; Urban Evolution; Innovation Diffusion
+
+\cite{zhang2019empirically}
 
@@
 This innovation profile corresponds to the urban genome.
-Each innovation has furthermore an utility $u_c >0$ which will influence its diffusion dynamics.
+Each innovation has furthermore a utility $u_c >0$ which will influence its diffusion dynamics.
 
@@
 % correlations
-%%%%%%%%%%%%
-\begin{figure}[t]
-	\centering
-	\includegraphics[width=\linewidth,trim={6cm 6cm 0 6cm}]{figures/corrmat_indics.png}
-	\caption{Correlation matrix between indicators, estimated on the full grid experiment.
-Confidence intervals for correlations estimated with the Fisher method are tiny and not distinguishable in the plot.\label{fig:cormat}}
-\end{figure}
-%%%%%%%%%%%%
 
@@
 Regarding the effective dimension of the indicator space, a principal component analysis on normalized indicators gives 48\% of variance on the first component, 77\% of cumulated variance on the second and 91\% on the third, confirming that even if correlation exist the behavior of indicators are rather independent.
+
+%%%%%%%%%%%%
+\begin{figure}[t]
+	\centering
+	\includegraphics[width=\linewidth,trim={6cm 6cm 0 6cm}]{figures/corrmat_indics.png}
+	\caption{Correlation matrix between indicators, estimated on the full grid experiment.
+Confidence intervals for correlations estimated with the Fisher method are tiny and not distinguishable in the plot.\label{fig:cormat}}
+\end{figure}
+%%%%%%%%%%%%
 
@@
 \footnotesize
-%\bibliographystyle{apalike}
-%\bibliography{biblio} % replace by the name of your .bib file
-
-\begin{thebibliography}{}
-
-\bibitem[Batty, 2009]{batty2009digital}
-Batty, M. (2009).
-\newblock A digital breeder for designing cities.
-\newblock {\em Architectural Design}, 79(4):46--49.
-
-\bibitem[Batty and Marshall, 2009]{batty2009centenary}
-Batty, M. and Marshall, S. (2009).
-\newblock Centenary paper: The evolution of cities: Geddes, abercrombie and the
-  new physicalism.
-\newblock {\em Town Planning Review}, 80(6):551--574.
-
-\bibitem[Bedau et~al., 2000]{bedau2000open}
-Bedau, M.~A., McCaskill, J.~S., Packard, N.~H., Rasmussen, S., Adami, C.,
-  Green, D.~G., Ikegami, T., Kaneko, K., and Ray, T.~S. (2000).
-\newblock Open problems in artificial life.
-\newblock {\em Artificial life}, 6(4):363--376.
-
-\bibitem[Bergeaud et~al., 2017]{bergeaud2017classifying}
-Bergeaud, A., Potiron, Y., and Raimbault, J. (2017).
-\newblock Classifying patents based on their semantic content.
-\newblock {\em PloS one}, 12(4):e0176310.
-
-\bibitem[Blommestein and Nijkamp, 1987]{blommestein1987adoption}
-Blommestein, H. and Nijkamp, P. (1987).
-\newblock Adoption and diffusion of innovations and the evolution of spatial
-  systems.
-\newblock In {\em Economic Evolution and Structural Adjustment}, pages
-  368--380. Springer.
-
-\bibitem[Bull et~al., 2000]{bull2000meme}
-Bull, L., Holland, O., and Blackmore, S. (2000).
-\newblock On meme--gene coevolution.
-\newblock {\em Artificial life}, 6(3):227--235.
-
-\bibitem[Cottineau et~al., 2015]{cottineau2015modular}
-Cottineau, C., Reuillon, R., Chapron, P., Rey-Coyrehourcq, S., and Pumain, D.
-  (2015).
-\newblock A modular modelling framework for hypotheses testing in the
-  simulation of urbanisation.
-\newblock {\em Systems}, 3(4):348--377.
-
-\bibitem[D'Acci, 2014]{d2014urban}
-D'Acci, L. (2014).
-\newblock Urban dna for cities evolutions. cities as physical expression of
-  dynamic equilibriums between competitive and cooperative forces.
-\newblock {\em arXiv preprint arXiv:1408.2874}.
-
-\bibitem[Deb et~al., 2002]{deb2002fast}
-Deb, K., Pratap, A., Agarwal, S., and Meyarivan, T. (2002).
-\newblock A fast and elitist multiobjective genetic algorithm: Nsga-ii.
-\newblock {\em IEEE transactions on evolutionary computation}, 6(2):182--197.
-
-\bibitem[Deffuant et~al., 2005]{deffuant2005individual}
-Deffuant, G., Huet, S., and Amblard, F. (2005).
-\newblock An individual-based model of innovation diffusion mixing social value
-  and individual benefit.
-\newblock {\em American Journal of Sociology}, 110(4):1041--1069.
-
-\bibitem[Durham, 1991]{durham1991coevolution}
-Durham, W.~H. (1991).
-\newblock {\em Coevolution: Genes, culture, and human diversity}.
-\newblock Stanford University Press.
-
-\bibitem[Ethiraj and Levinthal, 2004]{ethiraj2004modularity}
-Ethiraj, S.~K. and Levinthal, D. (2004).
-\newblock Modularity and innovation in complex systems.
-\newblock {\em Management science}, 50(2):159--173.
-
-\bibitem[Favaro and Pumain, 2011]{favaro2011gibrat}
-Favaro, J.-M. and Pumain, D. (2011).
-\newblock Gibrat revisited: An urban growth model incorporating spatial
-  interaction and innovation cycles. 再评吉尔布瑞特:
-  顾及空间相互作用和创新周期的城市增长模型.
-\newblock {\em Geographical Analysis}, 43(3):261--286.
-
-\bibitem[Fotheringham and O'Kelly, 1989]{fotheringham1989spatial}
-Fotheringham, A.~S. and O'Kelly, M.~E. (1989).
-\newblock {\em Spatial interaction models: formulations and applications},
-  volume~1.
-\newblock Kluwer Academic Publishers Dordrecht.
-
-\bibitem[Gong and Hassink, 2019]{gong2019co}
-Gong, H. and Hassink, R. (2019).
-\newblock Co-evolution in contemporary economic geography: Towards a
-  theoretical framework.
-\newblock {\em Regional Studies}, 53(9):1344--1355.
-
-\bibitem[Hagerstrand et~al., 1968]{hagerstrand1968innovation}
-Hagerstrand, T. et~al. (1968).
-\newblock Innovation diffusion as a spatial process.
-\newblock {\em Innovation diffusion as a spatial process.}
-
-\bibitem[Hidalgo et~al., 2007]{hidalgo2007product}
-Hidalgo, C.~A., Klinger, B., Barab{\'a}si, A.-L., and Hausmann, R. (2007).
-\newblock The product space conditions the development of nations.
-\newblock {\em Science}, 317(5837):482--487.
-
-\bibitem[Holland, 2012]{holland2012signals}
-Holland, J.~H. (2012).
-\newblock {\em Signals and boundaries: Building blocks for complex adaptive
-  systems}.
-\newblock Mit Press.
-
-\bibitem[Kaya and B{\"o}len, 2017]{kaya2017urban}
-Kaya, H.~S. and B{\"o}len, F. (2017).
-\newblock Urban dna: morphogenetic analysis of urban pattern.
-
-\bibitem[Kiesling et~al., 2012]{kiesling2012agent}
-Kiesling, E., G{\"u}nther, M., Stummer, C., and Wakolbinger, L.~M. (2012).
-\newblock Agent-based simulation of innovation diffusion: a review.
-\newblock {\em Central European Journal of Operations Research},
-  20(2):183--230.
-
-\bibitem[Lobo et~al., 2020]{lobo2020urban}
-Lobo, J., Alberti, M., Allen-Dumas, M., Arcaute, E., Barthelemy, M.,
-  Bojorquez~Tapia, L.~A., Brail, S., Bettencourt, L., Beukes, A., Chen, W.-Q.,
-  et~al. (2020).
-\newblock Urban science: Integrated theory from the first cities to sustainable
-  metropolises.
-
-\bibitem[Mesoudi, 2001]{mesoudi2001cultural}
-Mesoudi, A. (2001).
-\newblock Cultural evolution.
-\newblock {\em eLS}, pages 1--8.
-
-\bibitem[Passerat-Palmbach et~al., 2017]{passerat2017reproducible}
-Passerat-Palmbach, J., Reuillon, R., Leclaire, M., Makropoulos, A., Robinson,
-  E.~C., Parisot, S., and Rueckert, D. (2017).
-\newblock Reproducible large-scale neuroimaging studies with the openmole
-  workflow management system.
-\newblock {\em Frontiers in neuroinformatics}, 11:21.
-
-\bibitem[Pumain, 2018]{pumain2018evolutionary}
-Pumain, D. (2018).
-\newblock An evolutionary theory of urban systems.
-\newblock In {\em International and transnational perspectives on urban
-  systems}, pages 3--18. Springer.
-
-\bibitem[Pumain et~al., 2012]{pumain2012theorie}
-Pumain, D. et~al. (2012).
-\newblock Une th{\'e}orie g{\'e}ographique pour la loi de zipf.
-\newblock {\em R{\'e}gion et d{\'e}veloppement}, 36(31-54):125--150.
-
-\bibitem[Pumain et~al., 2006]{pumain2006evolutionary}
-Pumain, D., Paulus, F., Vacchiani-Marcuzzo, C., and Lobo, J. (2006).
-\newblock An evolutionary theory for interpreting urban scaling laws.
-\newblock {\em Cybergeo: European Journal of Geography}.
-
-\bibitem[Pumain and Raimbault, 2020]{pumain2020conclusion}
-Pumain, D. and Raimbault, J. (2020).
-\newblock Conclusion: Perspectives on urban theories.
-\newblock In {\em Theories and Models of Urbanization}, pages 303--330.
-  Springer.
-
-\bibitem[Pumain and Reuillon, 2017]{pumain2017simpoplocal}
-Pumain, D. and Reuillon, R. (2017).
-\newblock The simpoplocal model.
-\newblock In {\em Urban Dynamics and Simulation Models}, pages 21--35.
-  Springer.
-
-\bibitem[Raimbault, 2018a]{raimbault:halshs-01880492}
-Raimbault, J. (2018a).
-\newblock {A systematic comparison of interaction models for systems of
-  cities}.
-\newblock In {\em {Conference on Complex Systems 2018}}, Thessaloniki, Greece.
-
-\bibitem[Raimbault, 2018b]{raimbault2018co}
-Raimbault, J. (2018b).
-\newblock Co-evolution and morphogenetic systems.
-\newblock {\em arXiv preprint arXiv:1803.11457}.
-
-\bibitem[Raimbault, 2019]{raimbault:halshs-02351722}
-Raimbault, J. (2019).
-\newblock {A multi-scalar model for system of cities}.
-\newblock In {\em {Conference on Complex Systems 2019}}, Singapore, Singapore.
-
-\bibitem[Raimbault, 2020a]{raimbault2020cities}
-Raimbault, J. (2020a).
-\newblock Cities as they could be: Artificial life and urban systems.
-\newblock {\em arXiv preprint arXiv:2002.12926}.
-
-\bibitem[Raimbault, 2020b]{raimbault2020hierarchy}
-Raimbault, J. (2020b).
-\newblock Hierarchy and co-evolution processes in urban systems.
-\newblock {\em arXiv preprint arXiv:2001.11989}.
-
-\bibitem[Raimbault, 2020c]{raimbault2020indirect}
-Raimbault, J. (2020c).
-\newblock Indirect evidence of network effects in a system of cities.
-\newblock {\em Environment and Planning B: Urban Analytics and City Science},
-  47(1):138--155.
-
-\bibitem[Raimbault, 2020d]{raimbault2020unveiling}
-Raimbault, J. (2020d).
-\newblock Unveiling co-evolutionary patterns in systems of cities: a systematic
-  exploration of the simpopnet model.
-\newblock In {\em Theories and Models of Urbanization}, pages 261--278.
-  Springer.
-
-\bibitem[Raimbault et~al., 2019]{raimbault2019space}
-Raimbault, J., Cottineau, C., Le~Texier, M., Le~Nechet, F., and Reuillon, R.
-  (2019).
-\newblock Space matters: Extending sensitivity analysis to initial spatial
-  conditions in geosimulation models.
-\newblock {\em Journal of Artificial Societies and Social Simulation},
-  22(4):10.
-
-\bibitem[Reuillon et~al., 2013]{reuillon2013openmole}
-Reuillon, R., Leclaire, M., and Rey-Coyrehourcq, S. (2013).
-\newblock Openmole, a workflow engine specifically tailored for the distributed
-  exploration of simulation models.
-\newblock {\em Future Generation Computer Systems}, 29(8):1981--1990.
-
-\bibitem[Rozenblat and Pumain, 2007]{rozenblat2007firm}
-Rozenblat, C. and Pumain, D. (2007).
-\newblock Firm linkages, innovation and the evolution of urban systems.
-\newblock {\em Cities in globalization: Practices, policies, theories}, pages
-  130--156.
-
-\bibitem[Uyarra, 2010]{uyarra2010evolutionary}
-Uyarra, E. (2010).
-\newblock What is evolutionary about ‘regional systems of innovation’?
-  implications for regional policy.
-\newblock {\em Journal of evolutionary economics}, 20(1):115.
-
-\bibitem[Votsis and Haavisto, 2019]{votsis2019urban}
-Votsis, A. and Haavisto, R. (2019).
-\newblock Urban dna and sustainable cities: A multi-city comparison.
-\newblock {\em Frontiers in Environmental Science}, 7:4.
-
-\bibitem[Wu and Silva, 2011]{wu2011urban}
-Wu, N. and Silva, E.~A. (2011).
-\newblock Urban dna: exploring the biological metaphor of urban evolution with
-  dg-abc model.
-\newblock In {\em Proceedings The 14th AGILE International Conference on
-  Geographic Information Science. Utrecht University}.
-
-\end{thebibliography}
+\bibliographystyle{apalike}
+\bibliography{biblio} % replace by the name of your .bib file
 

### 2020-06-18 Docs/Papers/EvolutionInnovation/paper/paper.tex
.

@@
 
-% Modeling urban evolution with the diffusion of innovations
-% A simple model of urban evolution based on innovation diffusion
-% A model of urban evolution based on innovation diffusion
-% A model of urban evolution and innovation diffusion
-% A model of urban evolution based on the diffusion of innovations
-
 \title{A model of urban evolution based on innovation diffusion}
@@
 %Promising work. For the future, it will be interesting to see how the model can reproduce observed properties of different cities.
+%TODO discussion real system
 
@@
 It would have been very interesting to see how this would work for a model of a real city - there are an abundance of datasets and models (even cited by the author) that would not have been difficult to integrate.
+%TODO apply model to real city (more discussion)
 
 %In addition, it would be very useful to include a discussion on how modeling innovation at a microscopic scale (or in-between, eg at a level of suburbs) changes the results of the simulations.
+%TODO discussion on microscopic scale
 
@@
 %1. Evolutionary economics also study regional systems of innovation, how it diffuses and thee potential existence of spatial spillovers (Uyarra, 2010).
+% typo - OK
 %2. Finally, to existence of subsystems in which evolution can occur is natural through the spatial aspect of the model, ….
+% typo - OK
 %3. It is however significantly different in the way population are updated deterministically
+%TODO typo
 %4. Transformation processes are included as mutations, when random innovation appear in cities.
+% typo - OK
 %5. In our model, cities are characterized by their size in terms of population, and city size evolve following a spatial
+%TODO typo
 %6. Urban genomes are adapted such that the new innovation as an initial penetration rate ..
+%TODO typo
 %7. We choose simply to quantify the level of hierarchy at final time, estimated by fitting the rank-size law on populations using an Ordinary Least Squares on logarithms.
+%TODO typo
 %8. what means that stochastic noise is not an issue for interpreting indicator values … what means that these two last indicators need more statistical precision
+%TODO typo
 %9. we find that it always increase for larger population
+%TODO typo
 %10. it means that systems where i nteraction are more global will be less unequal. This is however not always the case in such urban dynamics model as
+%TODO typo
 %11. For low values of mutation rates, with find a piecewise linear
+%TODO typo
 %12.
 This peak is the strongest when innovation diffuse at broader range , what may mean that this regime is due to the emergence of regional ensemble of comparable competitivity and thus a higher chance of high utility innovation,
+%TODO typo
 %13. cumulated variance on the second and 91% on the third, confirming that even if correlation exist the behavior of
+%TODO typo
 %14. one may prefer in the compromise the points just before the transition which correspond to a local diffusion setting. Thus, local diffusion correspond to intermediate
+%TODO typo
 %15. urban genome can be used to simulated urban dynamics
+%TODO typo
 %16. model to real system …. and evoluting sub-systems
+%TODO typo
 %Other issues:
 %1. You need to deal with text formatting as some texts are being cut out at the end of the line (observed on paragraph 1 on page 5).
+%TODO pb cropping fig2 - OK?
 %2. Use of compromise and compromize in the same article, (kind of linear compromize between utility … prefer in the compromise the …)
+%TODO typo
 %3. It is not clear why the growth rate is fixed to 0.005 and the cited paper (Raimbault, 2020c) does not justify the use of that value.
+%TODO justify fixed growth rate
 %4.
@@
 As it is, it's hard to read.
+%TODO facet names
 %5.
@@
 I don’t think ‘(bottom raw) ’ is correct either.
+%TODO clarify
 %Your arguments are strong and I hope my comments will help to fine-tune your article to make it more readable. Good luck with your submission.
@@
 % Urban Systems; Urban Evolution; Innovation Diffusion
-
-\cite{zhang2019empirically}
 
@@
 The aforementioned evolutionary urban theory suggests that innovation cycles and their hierarchical diffusion is a possible explanation for the properties of urban scaling laws \citep{pumain2006evolutionary}.
-Evolutionary economics also study regional systems of innovation, how it diffuses and thee potential existence of spatial spillovers \citep{uyarra2010evolutionary}.
+Evolutionary economics also study regional systems of innovation, how it diffuses and the potential existence of spatial spillovers \citep{uyarra2010evolutionary}.
+At the microscopic scale, several agent-based models of innovation diffusion highlight the role of social and geographical networks, both of which are embedded into urban systems \citep{zhang2019empirically}.
 Thus, the diffusion of innovation is a privileged entry into urban evolutionary processes.
@@
 
-In our model, cities are characterized by their size in terms of population, and city size evolve following a spatial interaction model in which city attractivity is included.
+In our model, cities are characterized by their size in terms of population.
+City sizes evolve following a spatial interaction model taking into account the attractivity of cities.
 This attractivity is determined by how innovative cities are.
-Transformation processes are included as mutations, when random innovation appear in cities.
+Transformation processes are included as mutations, when random innovations appear in cities.
 Transmission processes (spatial crossover) are included by diffusing innovations between cities.
-Finally, to existence of subsystems in which evolution can occur is natural through the spatial aspect of the model, through which different regions in space may behave and evolve differently, possibly resulting in the emergence of co-evolution niches \citep{holland2012signals,raimbault2018co}.
+Finally, the existence of subsystems in which evolution can occur comes naturally with the spatial aspect of the model: different regions in space may behave and evolve differently, possibly resulting in the emergence of co-evolution niches \citep{holland2012signals,raimbault2018co}.
 
@@
 Second, applying the model to real system of cities, both in terms of initial parametrization and of empirical laws for innovation processes, would allow bringing this approach closer to possible policy applications.
-\cite{raimbault:halshs-01880492} benchmarked several models of urban growth based on interactions between cities, and integrated the model of \cite{favaro2011gibrat} on which this work was based.
-It however only included accurate initial populations and distance matrix.
-This application to real systems of cities also allows testing the performance of the model in predicting possible trajectories for these systems.
+\cite{raimbault:halshs-01880492} and later \cite{2020arXiv200510007R} benchmarked several models of urban growth based on interactions between cities, and integrated the model of \cite{favaro2011gibrat} on which this work was based.
+It however only included accurate initial populations and distance matrix, but no empirical data on innovation waves which were also synthetically constructed.
+One important aspect of this application to real systems of cities is thus a test of the performance of this model in predicting possible trajectories for systems of cities.
 
@@
 the macroscopic scale since cities are the basic agents.
+One can expect processes of innovation diffusion to behave differently at the microscopic scale or at intermediate scales of companies or innovation clusters % lit spillovers?
 Similarly, a single temporal scale was included, although two dynamics of spatial interaction and innovation diffusion are effectively combined.
@@
 Adapting this approach in the case of urban evolution through innovation diffusion would be an interesting potential development.
+The aspects to be clarified
 

### 2020-06-18 Models/EvolutionInnovation/analysis.R

@@ -98,6 +98,9 @@ res <- as.tbl(read.csv(file=paste0('openmole/exploration/',resprefix,'.csv')))
 res=res[res$averageUtility<quantile(res$averageUtility,c(0.95),na.rm = T)&!is.na(res$averageUtility),]
 res2=res
 
+res$mutationRateF = paste0('beta *"="*',res$mutationRate)
+res$earlyAdoptersRateF = paste0('r[0]*"="*',res$earlyAdoptersRate)
+
 for(indic in indics){
   for(distrib in unique(res$utilityDistribution)){
       for(newInnovationHierarchy in unique(res$newInnovationHierarchy)){
@@ -105,7 +108,7 @@ for(indic in indics){
         g=ggplot(res[res$utilityDistribution==distrib&res$newInnovationHierarchy==newInnovationHierarchy,],
                  aes_string(x="gravityDecay",y=indic,group="innovationDecay",color="innovationDecay")
         )
-        g+geom_smooth(se = F)+facet_grid(mutationRate~earlyAdoptersRate,scales = 'free')+
+        g+geom_smooth(se = F)+facet_grid(mutationRateF~earlyAdoptersRateF,scales = 'free', labeller =label_parsed)+
           xlab(expression(d[G]))+ylab(indicnames[[indic]])+scale_color_continuous(name=expression(d[I]))+stdtheme
         ggsave(filename = paste0(resdir,indic,'-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy',newInnovationHierarchy,'_distrib',distrib,'.png'),width=30,height=20,units='cm')
         

### 2020-06-19 Models/EvolutionInnovation/openmole/run.sh
new_explo

@@ -1,0 +1,7 @@
+
+for i in `seq 1 10`
+do
+  echo $i
+  /home/ubuntu/ComplexSystems/OpenMole/openmole-11.0-SNAPSHOT/openmole/openmole --mem 15G --password-file omlpwd -p /home/ubuntu/ComplexSystems/UrbanEvolution/Models/EvolutionInnovation/target/scala-2.12/urbanevolution_2.12-0.1-SNAPSHOT.jar --script Exploration.oms 
+done
+

### 2020-06-20 Docs/Papers/EvolutionInnovation/paper/paper.tex
.

@@
 %3. It is however significantly different in the way population are updated deterministically
-%TODO typo
+% typo - OK
 %4. Transformation processes are included as mutations, when random innovation appear in cities.
@@
 %5. In our model, cities are characterized by their size in terms of population, and city size evolve following a spatial
-%TODO typo
+%typo - OK
 %6. Urban genomes are adapted such that the new innovation as an initial penetration rate ..
-%TODO typo
+%typo - OK
 %7. We choose simply to quantify the level of hierarchy at final time, estimated by fitting the rank-size law on populations using an Ordinary Least Squares on logarithms.
-%TODO typo
+% typo - OK
 %8. what means that stochastic noise is not an issue for interpreting indicator values … what means that these two last indicators need more statistical precision
-%TODO typo
+% typo - OK
 %9. we find that it always increase for larger population
-%TODO typo
+% typo - OK
 %10. it means that systems where i nteraction are more global will be less unequal. This is however not always the case in such urban dynamics model as
-%TODO typo
+% typo - OK
 %11. For low values of mutation rates, with find a piecewise linear
-%TODO typo
+% typo - OK
 %12.
 This peak is the strongest when innovation diffuse at broader range , what may mean that this regime is due to the emergence of regional ensemble of comparable competitivity and thus a higher chance of high utility innovation,
-%TODO typo
+% typo - OK
 %13. cumulated variance on the second and 91% on the third, confirming that even if correlation exist the behavior of
-%TODO typo
+% typo - OK
 %14. one may prefer in the compromise the points just before the transition which correspond to a local diffusion setting. Thus, local diffusion correspond to intermediate
@@
 %15. urban genome can be used to simulated urban dynamics
-%TODO typo
+% typo - OK
 %16. model to real system …. and evoluting sub-systems
-%TODO typo
+% typo - OK
 %Other issues:
@@
 Our model builds on this last framework, extending and adapting it to an urban evolution context.
-It is however significantly different in the way population are updated deterministically and in the innovation process.
+We keep the innovation diffusion and spatial interaction processes unchanged, but take a deterministic update of populations.
+The introduction of new innovations is modified to better fit the urban evolution context.
+Finally, the introduction of indicators to quantify model behavior regarding different dimensions is a significant extension to the initial model.
 
@@
 The utility for a new innovation is drawn stochastically, following either a normal or log-normal distribution, such that (i) its average corresponds to the current mean of utilities for existing innovations, and (ii) its standard deviation is a fixed parameter $\sigma_U$, which allows controlling the ``disruptivity'' of innovations.
-Urban genomes are adapted such that the new innovation as an initial penetration rate $r_0$ in the innovative city (previous innovation shares are rescaled for this city).
+Urban genomes are modified such that the new innovation has an initial penetration rate $r_0$ in the innovative city (previous innovation shares are rescaled for this city).
 \end{enumerate}
@@
 Distances between cities $d_{ij}$ are computed as euclidian distances.
-We take $N=30$ cities and $t_f = 50$ in our experiments, corresponding to a regional or national system of cities, on timescales of the order of a century.
+We take $N=30$ cities, $P_{max} = 10^5$ and $t_f = 50$ in our experiments, corresponding to a regional or small national system of cities, on timescales of the order of a century.
 
@@
 	\item To quantify the trajectories of city populations, many indicators can be used \citep{raimbault2020unveiling}.
-We choose simply to quantify the level of hierarchy at final time, estimated by fitting the rank-size law on populations using an Ordinary Least Squares on logarithms.
-This captures if the urban systems has become more unequal in terms of population balance.
+We choose to simply quantify the level of hierarchy of final populations.
+It is estimated by adjusting a rank-size law for populations, using an Ordinary Least Squares regression on logarithms of populations as a function of logarithms of ranks.
+This captures if the urban systems has become more unequal in terms of population balance, since the initial hierarchy is fixed.
 \end{itemize}
@@
 
-Parameter explored in experiments are the spatial interaction parameters $d_G,d_I \in \left[0;2\right]$ which correspond to the crossover mechanism; and the innovation parameter which correspond to the mutation mechanism: innovation rate $\beta \in \left[0;1\right]$, innovation hierarchy $\alpha_I \in \left[0;2\right]$, innovation utility standard deviation $\sigma_U \in \left[ 0.7;2\right]$ (the lower bound is a constraint for the existence of the log-normal with log mean 0), initial penetration rate $r_0 \in \left[0.1;0.9\right]$, and type of innovation utility as normal or log-normal.
+Parameter explored in experiments are the spatial interaction parameters $d_G,d_I \in \left[0;2\right]$ which correspond to the crossover mechanism; and the innovation parameters which correspond to the mutation mechanism: innovation rate $\beta \in \left[0;1\right]$, innovation hierarchy $\alpha_I \in \left[0;2\right]$, innovation utility standard deviation $\sigma_U \in \left[ 0.7;2\right]$ (the lower bound is a constraint for the existence of the log-normal with log mean 0), initial penetration rate $r_0 \in \left[0.1;0.9\right]$, and type of innovation utility as normal or log-normal.
+All model parameters are summarized in Table~\ref{tab:params}, with their range, default value and associated process.
+
+%%%%%%%%%%%%
+\begin{table}
+\caption{Summary of model parameters.
+We first describe synthetic setup parameters which are fixed during our experiments, and then parameters varied during experiments.
+We give associated process, indicative range and default value for each parameter.\label{tab:params}}
+	\hspace{-0.5cm}
+	\footnotesize
+	\begin{tabular}{|l|l|l|l|l|l|}
+	\hline
+	Parameter & Not. & Process & Range & Def. \\ \hline
+	Number of cities & $N$ & Spatial scale & $[10 ; 100]$ & $30$\\
+	Initial hierarchy & $\alpha_0$ & System of cities & $[0.5 ; 2.0]$ & $1$\\
+	Initial population & $P_{max}$ & System of cities & $[10^4 ; 10^7]$ & $10^5$\\
+	Simulation steps & $t_f$ & Temporal scale & $[10 ; 100]$ & $50$\\\hline
+	Gravity range & $d_G$ & Crossover & $[0 ; 2]$ & $1$\\
+	Innovation range & $d_I$ & Crossover & $[0 ; 2]$ & $1$\\
+	Innovation rate & $\beta$ & Mutation & $[0 ; 1]$ & $0.5$\\
+	Innovation hierarchy & $\alpha_I$ & Mutation & $[0 ; 2]$ & $1$\\
+	Innov. utility std. & $\sigma_U$ & Mutation & $\left[ 0.7;2\right]$ & $1$\\
+	Penetration rate & $r_0$ & Mutation & $\left[0.1;0.9\right]$ & $0.5$\\
+	Utility type & - & Mutation & \{n;ln\}& ln \\
+	\hline
+	\end{tabular}
+\end{table}
+%%%%%%%%%%%%
 
@@
 
-The sharpe ratios estimated on repetitions as sample standard deviation relative to absolute sample mean have high values for all indicators (min.
+The Sharpe ratios estimated on repetitions as sample standard deviation relative to absolute sample mean have high values for all indicators (min.
 3.9 and median 12.1 for diversity; min.
@@
 1.7 and median 3.6 for utility; min.
-26 and median 257.3 for population hierarchy), what means that stochastic noise is not an issue for interpreting indicator values.
+26 and median 257.3 for population hierarchy).
+This implies that stochastic noise is not an issue for interpreting indicator values.
 
@@
 This distance is defined by $\Delta_{ij} = 2\frac{\left|\mu_i - \mu_j \right|}{\sigma_i + \sigma_j}$ if $\mu_i$ are estimated central values (mean or median) and $\sigma_i$ estimated standard deviations.
-These distances have an average and median larger than one for all couples, and a first quartile larger than one for innovation and population hierarchy, while the first quartile is 0.5 for utility and 0.6 for diversity, what means that these two last indicators need more statistical precision to be interpreted.
+These distances have an average and median larger than one for all couples, and a first quartile larger than one for innovation and population hierarchy, while the first quartile is 0.5 for utility and 0.6 for diversity.
+The most constraining indicators in terms of statistical precision are thus utility and diversity.
 As in the case of normal distributions, a 95\% confidence interval of size $\sigma_2$ is achieved for $n\simeq 64$ runs, a number of 50 runs is satisfying for further experiments.
+%TODO new number of runs
 
@@
 The Fig.~\ref{fig:exploration} shows the behavior of indicators with remaining parameters varying, namely spatial interaction parameters $d_G$ and $d_I$, mutation rate $\beta$ and innovation adoption $r_0$.
-Regarding the behavior of diversity (top left panel of Fig.~\ref{fig:exploration}), we find that it always increase for larger population growth spatial interaction distances, meaning that broader population exchanges will induce a higher diversity in innovations, probably through the higher innovation dynamics induced.
+Regarding the behavior of diversity (top left panel of Fig.~\ref{fig:exploration}), we find that it always increases for larger population growth spatial interaction distances, meaning that broader population exchanges will induce a higher diversity in innovations, probably through the higher innovation dynamics induced.
 This increase however reaches a plateau around half of the width of the world ($d_G = 0.5$) which is more obvious for higher values of mutation rates (bottom raw): higher intrinsic innovation dampens the role of spatial interactions.
@@
 We find that these increase both with $d_G$ and $d_I$, meaning that more interactions and more diffusion will foster local innovation.
-Regarding the final distribution of population, it means that systems where interaction are more global will be less unequal.
+Regarding the final distribution of populations, this implies that systems where interactions are more global will be less unequal.
 This is however not always the case in such urban dynamics model as explored by \cite{raimbault2020hierarchy}.
@@
 The effect of initial adoption rate is small for these indicators.
-Note that although they appear highly correlated in this region of the parameter space, they are not the indicators with the highest correlation as shown further (see Fig.~\ref{fig:cormat}).
+Note that although they appear highly correlated in this region of the parameter space, they are not the indicators with the highest correlation as shown further (see Fig.~\ref{fig:cormat} for a correlation matrix between indicators).
 
 Finally, the indicator with a more interesting behavior is the utility (bottom left panel of Fig.~\ref{fig:exploration}).
-For low values of mutation rates, with find a piecewise linear behavior as a function of $d_G$ and small effects of $d_I$, meaning that low innovation settings yield regimes where broader interactions are systematically desirable for the all system.
+For low values of mutation rates, we find a piecewise linear behavior as a function of $d_G$ and small effects of $d_I$, meaning that low innovation settings yield regimes where broader interactions are systematically desirable for the all system.
 However, when $\beta$ is higher, we witness a maximum of utility as a function of $d_G$, consistent across different values of $d_I$ and of $r_0$.
 This corresponds to an intermediate regional regime where local innovation regimes are more beneficial than a global integration.
-This peak is the strongest when innovation diffuse at broader range, what may mean that this regime is due to the emergence of regional ensemble of comparable competitivity and thus a higher chance of high utility innovation, while a globalized system would concentrate on a single dominating city and be overall less performant.
+This peak is the strongest when innovations diffuse at a broader range, which could be consistent with this regime being due to the emergence of regional ensembles of comparable competitivity.
+Such a setting would yield a higher chance of introducing innovations with a high utility, while a globalized system would concentrate on a single dominating city and be overall less performant.
 
@@
 Innovation and utility have a correlation lower than 0.5, and are thus rather independent.
-Regarding the effective dimension of the indicator space, a principal component analysis on normalized indicators gives 48\% of variance on the first component, 77\% of cumulated variance on the second and 91\% on the third, confirming that even if correlation exist the behavior of indicators are rather independent.
+Regarding the effective dimension of the indicator space, a principal component analysis on normalized indicators gives 48\% of variance on the first component, 77\% of cumulated variance on the second and 91\% on the third.
+This confirms that even if correlations exist, the behavior of indicators are rather independent.
 
@@
 Although much more empirical work would be needed to compare these stylized facts to real world settings, our model suggests in this particular setting that too much integration is not always optimal, what corresponds to the theoretical fact that complex systems are generally modular at different scales \citep{ethiraj2004modularity}.
-More generally, regarding the implications of our results for possible formalizations and theories of urban evolution which would closely build on biological and cultural evolution by extending them, we have demonstrated how a particular instance of an urban genome can be used to simulated urban dynamics, including fundamental processes needed to effectively have evolution.
+More generally, regarding the implications of our results for possible formalizations and theories of urban evolution which would closely build on biological and cultural evolution by extending them, we have demonstrated how a particular instance of an urban genome can be used to simulate urban dynamics, including fundamental processes needed to effectively have evolution.
 To what extent this approach relates to existing approaches of urban evolution which use other definitions remains to be investigated.
@@
 
-Second, applying the model to real system of cities, both in terms of initial parametrization and of empirical laws for innovation processes, would allow bringing this approach closer to possible policy applications.
+Second, applying the model to real systems of cities, both in terms of initial parametrization and of empirical laws for innovation processes, would allow bringing this approach closer to possible policy applications.
 \cite{raimbault:halshs-01880492} and later \cite{2020arXiv200510007R} benchmarked several models of urban growth based on interactions between cities, and integrated the model of \cite{favaro2011gibrat} on which this work was based.
@@
 
-We have introduced a simple model of urban evolution integrating effectively the evolutionary processes of transmission, transformation and evoluting sub-systems, through innovation diffusion and spatial interactions.
+We have introduced a simple model of urban evolution integrating effectively the evolutionary processes of transmission, transformation and evolving sub-systems, through innovation diffusion and spatial interactions.
 Model exploration yield complex behavior while multi-objective optimization shows the potentiality for the model to produce compromises between utility and diversity in the system of cities.

### 2020-06-20 Models/EvolutionInnovation/analysis.R

@@ -91,11 +91,23 @@ for(indic in indics){
 }
 
 #
-resprefix= '20200429_215249_GRID_GRID'
+#resprefix= '20200429_215249_GRID_GRID'
+#resfiles=c('20200429_215249_GRID_GRID')
+resprefix= '20200619_GRID_GRID'
+resfiles=c(#'20200429_215249_GRID_GRID',
+           '20200618_140526_GRID_GRID','20200618_164030_GRID_GRID','20200618_190830_GRID_GRID','20200618_213402_GRID_GRID',
+           '20200619_001006_GRID_GRID','20200619_025115_GRID_GRID','20200619_052000_GRID_GRID','20200619_084724_GRID_GRID',
+           '20200619_112519_GRID_GRID')
+
 resdir = paste0(Sys.getenv('CS_HOME'),'/UrbanEvolution/Results/EvolutionInnovation/',resprefix,'/');dir.create(resdir)
 
-res <- as.tbl(read.csv(file=paste0('openmole/exploration/',resprefix,'.csv')))
-res=res[res$averageUtility<quantile(res$averageUtility,c(0.95),na.rm = T)&!is.na(res$averageUtility),]
+res <- as.tbl(read.csv(file=paste0('openmole/exploration/',resfiles[1],'.csv')))
+if(is.null(res[['initialHierarchy']])){res[['initialHierarchy']]=rep(1,nrow(res))}
+for(resfile in resfiles[2:length(resfiles)]){
+res <- rbind(res,as.tbl(read.csv(file=paste0('openmole/exploration/',resfile,'.csv'))))
+}
+
+#res=res[res$averageUtility<quantile(res$averageUtility,c(0.95),na.rm = T)&!is.na(res$averageUtility),]
 res2=res
 
 res$mutationRateF = paste0('beta *"="*',res$mutationRate)
@@ -159,10 +171,11 @@ summary(prcomp(m))
 ############
 # calibration
 
-resprefix = 'CALIBRATION_GRID_20200429_220019'
+#resprefix = 'CALIBRATION_GRID_20200429_220019';finalgen='9000'
+resprefix = 'CALIBRATION_GRID_20200618_103726';finalgen='27000'
 resdir = paste0(Sys.getenv('CS_HOME'),'/UrbanEvolution/Results/EvolutionInnovation/',resprefix,'/');dir.create(resdir)
 
[…]

### 2020-06-20 Models/EvolutionInnovation/analysis.R
all_calib_results

@@ -94,10 +94,9 @@ for(indic in indics){
 #resprefix= '20200429_215249_GRID_GRID'
 #resfiles=c('20200429_215249_GRID_GRID')
 resprefix= '20200619_GRID_GRID'
-resfiles=c(#'20200429_215249_GRID_GRID',
-           '20200618_140526_GRID_GRID','20200618_164030_GRID_GRID','20200618_190830_GRID_GRID','20200618_213402_GRID_GRID',
+resfiles=c('20200618_140526_GRID_GRID','20200618_164030_GRID_GRID','20200618_190830_GRID_GRID','20200618_213402_GRID_GRID',
            '20200619_001006_GRID_GRID','20200619_025115_GRID_GRID','20200619_052000_GRID_GRID','20200619_084724_GRID_GRID',
-           '20200619_112519_GRID_GRID')
+           '20200619_112519_GRID_GRID','20200619_140601_GRID_GRID')
 
 resdir = paste0(Sys.getenv('CS_HOME'),'/UrbanEvolution/Results/EvolutionInnovation/',resprefix,'/');dir.create(resdir)
 
@@ -107,7 +106,7 @@ for(resfile in resfiles[2:length(resfiles)]){
 res <- rbind(res,as.tbl(read.csv(file=paste0('openmole/exploration/',resfile,'.csv'))))
 }
 
-#res=res[res$averageUtility<quantile(res$averageUtility,c(0.95),na.rm = T)&!is.na(res$averageUtility),]
+res=res[res$averageUtility<quantile(res$averageUtility,c(0.95),na.rm = T)&!is.na(res$averageUtility),]
 res2=res
 
 res$mutationRateF = paste0('beta *"="*',res$mutationRate)
@@ -116,13 +115,14 @@ res$earlyAdoptersRateF = paste0('r[0]*"="*',res$earlyAdoptersRate)
 for(indic in indics){
   for(distrib in unique(res$utilityDistribution)){
       for(newInnovationHierarchy in unique(res$newInnovationHierarchy)){
+        for(utilityStd in unique(res$utilityStd)){
         
-        g=ggplot(res[res$utilityDistribution==distrib&res$newInnovationHierarchy==newInnovationHierarchy,],
+        g=ggplot(res[res$utilityDistribution==distrib&res$newInnovationHierarchy==newInnovationHierarchy&res$utilityStd==utilityStd,],
                  aes_string(x="gravityDecay",y=indic,group="innovationDecay",color="innovationDecay")
         )
[…]

### 2020-06-22 Docs/Papers/EvolutionInnovation/paper/paper.tex
cr_paper

@@
 $^2$UPS CNRS 3611 ISC-PIF\\
-$^3$UMR CNRS 8504 G{\'e}ographie-cit{\'e}s\\
+$^3$UMR CNRS 8504 G{\'e}ographie-cit{\'e}s\medskip\\
 juste.raimbault@polytechnique.edu} % email of corresponding author
@@
 %Promising work. For the future, it will be interesting to see how the model can reproduce observed properties of different cities.
-%TODO discussion real system
+% discussion real system - OK
 
@@
 It would have been very interesting to see how this would work for a model of a real city - there are an abundance of datasets and models (even cited by the author) that would not have been difficult to integrate.
-%TODO apply model to real city (more discussion)
+% apply model to real city (more discussion) - OK
 
 %In addition, it would be very useful to include a discussion on how modeling innovation at a microscopic scale (or in-between, eg at a level of suburbs) changes the results of the simulations.
-%TODO discussion on microscopic scale
+% discussion on microscopic scale - OK
 
@@
 %14. one may prefer in the compromise the points just before the transition which correspond to a local diffusion setting. Thus, local diffusion correspond to intermediate
-%TODO typo
+% typo - OK
 %15. urban genome can be used to simulated urban dynamics
@@
 %1. You need to deal with text formatting as some texts are being cut out at the end of the line (observed on paragraph 1 on page 5).
-%TODO pb cropping fig2 - OK?
+% pb cropping fig2 - OK?
 %2. Use of compromise and compromize in the same article, (kind of linear compromize between utility … prefer in the compromise the …)
-%TODO typo
+% typo - OK
 %3. It is not clear why the growth rate is fixed to 0.005 and the cited paper (Raimbault, 2020c) does not justify the use of that value.
-%TODO justify fixed growth rate
+% justify fixed growth rate - OK
 %4.
@@
 As it is, it's hard to read.
-%TODO facet names
+% facet names - OK
 %5.
@@
 I don’t think ‘(bottom raw) ’ is correct either.
-%TODO clarify
+% clarify - OK
 %Your arguments are strong and I hope my comments will help to fine-tune your article to make it more readable. Good luck with your submission.
@@
 Explorations of the model on synthetic systems of cities show the role of spatial interaction and innovation diffusion ranges on measures of diversity and utility, and the existence of intermediate ranges yielding an optimal utility.
-Multi-objective optimization shows how the model produces a compromise between utility and diversity.
+Multi-objective optimization shows how the model produces a compromize between utility and diversity.
 This model paves the way towards more elaborated formalizations of urban evolution.
@@
 % (iii) Innovation diffusion - in ALife studies ; - in urban systems
-%\cite{Eletreby5664} diffusion/evolving
 
@@
 One aspect of open-ended evolution are mechanisms endogeneously producing novelty.
+The other way around, artificial societies approaches have also been pinned as a privileged entry for understanding the dynamics of innovation \citep{zenobia2009artificial}.
 In the case of urban systems, \cite{hagerstrand1968innovation} had already highlighted the role of a hierarchical diffusion of innovation between cities in their trajectories.
 The aforementioned evolutionary urban theory suggests that innovation cycles and their hierarchical diffusion is a possible explanation for the properties of urban scaling laws \citep{pumain2006evolutionary}.
+\cite{paulus2004co} provides empirical support for this approach by unveiling the co-evolution between economic functions of urban areas for the French systems of cities, which are closely related to the dynamics of innovation.
 Evolutionary economics also study regional systems of innovation, how it diffuses and the potential existence of spatial spillovers \citep{uyarra2010evolutionary}.
 At the microscopic scale, several agent-based models of innovation diffusion highlight the role of social and geographical networks, both of which are embedded into urban systems \citep{zhang2019empirically}.
-Thus, the diffusion of innovation is a privileged entry into urban evolutionary processes.
+\cite{daude2004apports} proposes for example to understand the diffusion of an agricultural innovation within a geographical system using an agent-based model.
+Thus, the diffusion of innovation is a privileged entry into urban evolutionary processes, as it is furthermore tightly linked with other dimensions of urban systems such as the social and economic dimensions \citep{sonis1995territorial}.
 
@@
 more innovative cities are more attractive.
-This stage is also deterministic, following  with $P_i(t) - P_i(t-1) = w_I\cdot \sum_j \frac{V_{ij}}{<V_{ij}>}$ where $w_I$ is a fixed growth rate fixed to $w_I = 0.005$ in experiments, and where the interaction potential is defined by
+This stage is also deterministic, following  with $P_i(t) - P_i(t-1) = w_I\cdot \sum_j \frac{V_{ij}}{<V_{ij}>}$ where $w_I$ is a fixed growth rate, and where the interaction potential is defined by
 \begin{equation}
@@
 The urban evolution is stopped after a fixed number of steps $t_f$.
+Note that other stopping criteria such as reaching a total population, a maximal population for the largest city, or a maximal number of innovations introduced, could be included for other applications of the model.
 
@@
 We take $N=30$ cities, $P_{max} = 10^5$ and $t_f = 50$ in our experiments, corresponding to a regional or small national system of cities, on timescales of the order of a century.
+According to these approximates scales, we fix $w_I = 0.005$ in experiments which gives a maximal possible growth factor of $(1.005)^{50} = 1.28$ over the whole period, in consistence with mature urban systems having relatively slow population growth \citep{pumain1982dynamique}.
 
@@
 	\centering
-	\includegraphics[width=0.49\linewidth]{figures/averageDiversity-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_distriblog-normal.png}
-	\includegraphics[width=0.49\linewidth]{figures/averageInnovation-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_distriblog-normal.png}
-	\includegraphics[width=0.49\linewidth]{figures/averageUtility-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_distriblog-normal.png}
-	\includegraphics[width=0.49\linewidth]{figures/finalHierarchy-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_distriblog-normal.png}
+	\includegraphics[width=0.49\linewidth]{figures/averageDiversity-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_utilityStd1_distriblog-normal}
+	\includegraphics[width=0.49\linewidth]{figures/averageInnovation-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_utilityStd1_distriblog-normal.png}
+	\includegraphics[width=0.49\linewidth]{figures/averageUtility-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_utilityStd1_distriblog-normal.png}
+	\includegraphics[width=0.49\linewidth]{figures/finalHierarchy-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_utilityStd1_distriblog-normal.png}
 	\caption{Values of indicators obtained with the grid model exploration.
@@
 We first describe synthetic setup parameters which are fixed during our experiments, and then parameters varied during experiments.
-We give associated process, indicative range and default value for each parameter.\label{tab:params}}
-	\hspace{-0.5cm}
+We give associated process, indicative range and default value for each parameter.\label{tab:params}}\medskip
+	\hspace{-1cm}
 	\footnotesize
@@
 	Initial population & $P_{max}$ & System of cities & $[10^4 ; 10^7]$ & $10^5$\\
-	Simulation steps & $t_f$ & Temporal scale & $[10 ; 100]$ & $50$\\\hline
+	Simulation steps & $t_f$ & Temporal scale & $[10 ; 100]$ & $50$\\
+	Growth rate & $w_I$ & Pop. growth & $[0.001 ; 0.01]$ & $0.005$\\\hline
 	Gravity range & $d_G$ & Crossover & $[0 ; 2]$ & $1$\\
@@
 As in the case of normal distributions, a 95\% confidence interval of size $\sigma_2$ is achieved for $n\simeq 64$ runs, a number of 50 runs is satisfying for further experiments.
-%TODO new number of runs
 
@@
 We find that the type of distribution for utility has small effects for all indicators and across most parameter values, but particular cases are more interesting in the log-normal case, which we comment further.
-Similarly, the effect of innovation hierarchy $\alpha_I$ and of innovation utility standard deviation $\sigma_U$ do not induce strong qualitative differences, so we discuss model behavior obtained at $\alpha_I = 1$ and $\sigma_U = 1$.
+Similarly, the effect of innovation hierarchy $\alpha_I$ and of innovation utility standard deviation $\sigma_U$ do not seem to induce strong qualitative differences, so we discuss model behavior obtained at $\alpha_I = 1$ and $\sigma_U = 1$.
 
@@
 Regarding the behavior of diversity (top left panel of Fig.~\ref{fig:exploration}), we find that it always increases for larger population growth spatial interaction distances, meaning that broader population exchanges will induce a higher diversity in innovations, probably through the higher innovation dynamics induced.
-This increase however reaches a plateau around half of the width of the world ($d_G = 0.5$) which is more obvious for higher values of mutation rates (bottom raw): higher intrinsic innovation dampens the role of spatial interactions.
+This increase is however slowed down around half of the width of the world ($d_G = 0.5$), which is more obvious with higher values of mutation rates for which a plateau is reached (bottom row, in particular the sub-panel for $r_0=0.1$ and $\beta = 0.5$): higher intrinsic innovation dampens the role of spatial interactions.
 An increase of innovation interaction distance $d_I$ on the contrary monotonously decreases diversity, consistently with the idea of local innovation niches which can be overridden by competing innovations of higher utility coming from further away.
@@
 Such a setting would yield a higher chance of introducing innovations with a high utility, while a globalized system would concentrate on a single dominating city and be overall less performant.
+
+%%%%%%%%%%%%
+\begin{figure}[t]
+	\centering
+	\includegraphics[width=\linewidth,trim={6cm 6cm 0 6cm}]{figures/corrmat_indics.png}
+	\caption{Correlation matrix between indicators, estimated on the full grid experiment.
+Confidence intervals for correlations estimated with the Fisher method are tiny and not distinguishable in the plot.\label{fig:cormat}}
+\end{figure}
+%%%%%%%%%%%%
 
@@
 
-%%%%%%%%%%%%
-\begin{figure}[t]
-	\centering
-	\includegraphics[width=\linewidth,trim={6cm 6cm 0 6cm}]{figures/corrmat_indics.png}
-	\caption{Correlation matrix between indicators, estimated on the full grid experiment.
-Confidence intervals for correlations estimated with the Fisher method are tiny and not distinguishable in the plot.\label{fig:cormat}}
-\end{figure}
-%%%%%%%%%%%%
-
 \subsection{Model optimization}
@@
 We apply thus a bi-objective genetic algorithm on the model, trying to maximize simultaneously utility and diversity.
-The algorithm is the NSGA2 algorithm \citep{deb2002fast}, which we run for 10,000 generations with a population of 200 individuals on a Island distribution scheme.
+The algorithm is the NSGA2 algorithm \citep{deb2002fast}, which we run for 50,000 generations with a population of 200 individuals and an Island distribution scheme.
 The algorithm is integrated into the OpenMOLE software, and the stochastic aspect is internally tackled by using an embedding strategy, i.e.
-by adding the number of repetitions as an additional objective to find compromise points between the number of run needed and their statistical accuracy.
+by adding the number of repetitions as an additional objective to find compromize points between the number of run needed and their statistical accuracy.
 
 We show in Fig.~\ref{fig:pareto} the obtained points, which are close to a Pareto front.
-We find two regimes, the first corresponding to the upper points in the plot for higher utilities but lower diversities, which should correspond to the intermediate optimal regimes identified before, and in which some kind of linear compromize between utility and diversity exists: increasing global utility is done at the cost of reducing diversity.
+We include the points with at least 20 stochastic samples.
+We find three regimes, the first corresponding to the upper points in the plot for higher utilities but lower diversities, which should correspond to the intermediate optimal regimes identified before, and in which some kind of linear compromize between utility and diversity exists (but with a very flat slope for utility): increasing global utility is done at the cost of reducing diversity.
 Within this first regime, two very different parameter setting coexist, one with high innovation diffusion (light blue, points with higher utility), the other in dark blue with a more local innovation diffusion.
 This means that reducing the span of innovation diffusion will increase diversity as one could expect.
-A second part of the Pareto front, obtained after a sharp transition, allows increasing the diversity but at the price of a much lower utility (points on the bottom right).
-These points are all close to equivalent, and one may prefer in the compromise the points just before the transition which correspond to a local diffusion setting.
-Thus, local diffusion correspond to intermediate compromises, while broad diffusion corresponds to extremes in the Pareto front.
-In a nutshell, this optimization exercise is interesting both to show how the model produces compromises, but also how it could be used in practice for innovation policies applied to systems of cities.
+A second part of the Pareto front corresponds to intermediate points with lower utility but higher diversity, and is exclusively composed of local innovation parameters.
+The third part of the Pareto front, obtained after a sharp transition, allows increasing significantly the diversity compared to the first part, but at the price of a much lower utility (points on the bottom right).
+These points are all close to equivalent, and one may prefer to consider as a compromize the extreme points which correspond to a broad diffusion setting, but this last regime is also mixed with local innovation.
+Thus, local diffusion correspond to intermediate compromizes, while broad diffusion corresponds to extremes in the Pareto front (first and third regimes).
+In a nutshell, this optimization exercise is interesting both to show how the model produces compromizes, but also how it could be used in practice for innovation policies applied to systems of cities.
 
@@
 \citep{hidalgo2007product} show in terms of industrial production by countries that these industrial spaces are highly dimensional in terms of product produced and interdependencies between countries and types of production, increasing the path-dependency in economic trajectories.
-The investigation of patent data from a semantic viewpoint also shows the highly dimensional nature of technological innovation \citep{bergeaud2017classifying}.
+The investigation of patent data from a semantic viewpoint also shows the highly dimensional nature of technological innovation \citep{raimbault2016investigating,bergeaud2017classifying}.
 A direct extension of our model would consist in having a matrix genome instead of a vector one.
-Innovations would occur across several dimensions which can correspond to industrial or technological domains, but also social and cultural innovation and innovations in terms of infrastructures which condition the way systems of cities evolve.
+Innovations would occur across several dimensions which can correspond to industrial or technological domains, but also social and cultural innovation and innovations in terms of infrastructures which condition the way systems of cities evolve \citep{raimbault2018caracterisation}.
 Each dimension should have particular innovation rules depending on its nature, and dependencies across dimensions could be introduced, implementing the possible emergence of technological co-evolution niches beside the spatial co-evolution niches.
@@
 It however only included accurate initial populations and distance matrix, but no empirical data on innovation waves which were also synthetically constructed.
-One important aspect of this application to real systems of cities is thus a test of the performance of this model in predicting possible trajectories for systems of cities.
+One important aspect of this application to real systems of cities is thus a test of the performance of this model in predicting possible population trajectories for systems of cities.
+Several obstacles however remain for a more complete application of the innovation diffusion urban evolution model.
+Harmonized datasets with consistent definition of urban entities, in particular which can be compared between different urban systems and long time scales, are crucial to study the dynamics of urban systems as done for example by \cite{pumain2015multilevel}.
+Innovation measures are not included in such datasets, and existing innovation proxys are limited in time and space, as for example the US Patent Office database for which inventors have been localized \citep{morrison2017disambiguation}.
+Then, how to link innovation proxy measures with model processes, in other words how to parametrize or calibrate the model on real data, remains an open question.
+Finally, how various urban systems with very different histories and at different stages of development would fit model assumptions, in particular the fact that innovation waves partly drive growth rates of cities, remains to be empirically investigated \citep{pumain2009innovation}.
 
@@
 the macroscopic scale since cities are the basic agents.
-One can expect processes of innovation diffusion to behave differently at the microscopic scale or at intermediate scales of companies or innovation clusters % lit spillovers?
+One can expect processes of innovation diffusion to behave differently at the microscopic scale or at intermediate scales of companies or innovation clusters \cite{audretsch2004knowledge}.
+Other types of agents and processes may be more significant than spatial interaction at this finer scales, for example with a diffusion through social networks \citep{zhang2019empirically}.
 Similarly, a single temporal scale was included, although two dynamics of spatial interaction and innovation diffusion are effectively combined.
@@
 Adapting this approach in the case of urban evolution through innovation diffusion would be an interesting potential development.
-The aspects to be clarified
 
@@
 We have introduced a simple model of urban evolution integrating effectively the evolutionary processes of transmission, transformation and evolving sub-systems, through innovation diffusion and spatial interactions.
-Model exploration yield complex behavior while multi-objective optimization shows the potentiality for the model to produce compromises between utility and diversity in the system of cities.
+Model exploration yield complex behavior while multi-objective optimization shows the potentiality for the model to produce compromizes between utility and diversity in the system of cities.
 This work is thus a first step towards more elaborated models of urban evolution.

### 2020-06-22 Models/EvolutionInnovation/analysis.R

@@ -91,12 +91,12 @@ for(indic in indics){
 }
 
 #
-#resprefix= '20200429_215249_GRID_GRID'
-#resfiles=c('20200429_215249_GRID_GRID')
-resprefix= '20200619_GRID_GRID'
-resfiles=c('20200618_140526_GRID_GRID','20200618_164030_GRID_GRID','20200618_190830_GRID_GRID','20200618_213402_GRID_GRID',
-           '20200619_001006_GRID_GRID','20200619_025115_GRID_GRID','20200619_052000_GRID_GRID','20200619_084724_GRID_GRID',
-           '20200619_112519_GRID_GRID','20200619_140601_GRID_GRID')
+resprefix= '20200429_215249_GRID_GRID'
+resfiles=c('20200429_215249_GRID_GRID')
+#resprefix= '20200619_GRID_GRID'
+#resfiles=c('20200618_140526_GRID_GRID','20200618_164030_GRID_GRID','20200618_190830_GRID_GRID','20200618_213402_GRID_GRID',
+#           '20200619_001006_GRID_GRID','20200619_025115_GRID_GRID','20200619_052000_GRID_GRID','20200619_084724_GRID_GRID',
+#           '20200619_112519_GRID_GRID','20200619_140601_GRID_GRID')
 
 resdir = paste0(Sys.getenv('CS_HOME'),'/UrbanEvolution/Results/EvolutionInnovation/',resprefix,'/');dir.create(resdir)
 
@@ -173,7 +173,7 @@ summary(prcomp(m))
 # calibration
 
 #resprefix = 'CALIBRATION_GRID_20200429_220019';finalgen='9000'
-resprefix = 'CALIBRATION_GRID_20200618_103726';finalgen='27000'
+resprefix = 'CALIBRATION_GRID_20200618_103726';finalgen='50000'
 resdir = paste0(Sys.getenv('CS_HOME'),'/UrbanEvolution/Results/EvolutionInnovation/',resprefix,'/');dir.create(resdir)
 
 res <- as.tbl(read.csv(file=paste0('openmole/calibration/',resprefix,'/population',finalgen,'.csv')))

### 2020-11-25 Docs/Seminars/20201209_UTorontoUGP/abstract.tex
abstract_ugp_seminar

\documentclass[10pt]{article}

\usepackage[margin=3cm]{geometry}

\title{Modeling urban evolution and co-evolution in systems of cities\medskip\\
\textit{Urban Genome Project Seminar}\\
\textit{University of Toronto}\\
\textit{December 9th 2020}}

\author{Juste Raimbault$^{1,2,3,\ast}$\medskip\\
$^{1}$ Center for Advanced Spatial Analysis, University College London\\
$^{2}$ UPS CNRS 3611 Complex Systems Institute Paris Ile-de-France\\
$^{3}$ UMR CNRS 8504 G{\'e}ographie-cit{\'e}s\medskip\\
$^{\ast}$ \texttt{juste.raimbault@polytechnique.edu}
}

\date{}

\pagenumbering{gobble}

\begin{document}

\maketitle

Systems of cities are adaptive complex systems, which have been theorised and modelled from many viewpoints and by different disciplines. Understanding the processes driving their dynamics is a crucial challenge for sustainable urban and territorial planning. The concept of urban evolution has proved relevant in this regard. This presentation summarises recent results obtained within the frame of the Evolutionary Urban Theory, a geographical approach to urban systems dynamics developed for more than 20 years by Denise Pumain and collaborators. We first review the main theoretical assumptions underlying this theory and results obtained with different simulation models in previous works. We also highlight the role of new model exploration and validation practices and tools developed in that context, implemented in the OpenMOLE platform.

We then describe a model of urban evolution positioned in this approach, in which the diffusion of innovations between cities captures transformation processes (mutations) and transmission processes (diffusion), using two coupled spatial interaction models. Explorations of the model on synthetic systems of cities show the role of spatial interaction and innovation diffusion ranges on measures of diversity and utility, and the existence of intermediate interaction ranges yielding an optimal utility. Multi-objective optimization shows how the model produces a compromise between utility and diversity.

The last part of the presentation focuses on the concept of co-evolution in urban systems, proposing a specific definition, a method to characterise it based on circular causations, and an application to the co-evolution between transportation networks and territories. For this particular case, we summarise results from co-evolution models at the mesoscopic scale (co-evolution between urban form and road networks) and at the macroscopic scale (co-evolution between cities and urban networks). We finally discuss perspectives towards more complex and multi-scalar models of urban evolution.

\end{document}

### 2020-12-09 Docs/Seminars/20201209_UTorontoUGP/communication/comment.tex
slides_seminar_ugp

\input{header_slides.tex}

\begin{document}

\title[A Critique of Urban Genome Project]{A Critique of Urban Genome Project}
\author[Raimbault]{J.~Raimbault$^{1,2,3\ast}$\\\medskip
$^{\ast}$\texttt{j.raimbault@ucl.ac.uk}
}

\institute[UCL]{$^{1}$Center for Advanced Spatial Analysis, University College London\\
$^{2}$UPS CNRS 3611 Complex Systems Institute Paris\\
$^{3}$UMR CNRS 8504 G{\'e}ographie-cit{\'e}s
}

\date[July 18th 2020]{Urban Genome Project seminar\\
University of Toronto, School of Cities\\
December 9th 2020
}

\frame{\maketitle}

\AtBeginSection[]
{
	\frame{
		\tableofcontents[currentsection, hideallsubsections]
	}
	\addtocounter{framenumber}{-1}
}

%

\section{General comments}

\sframe{Modeling urban evolution}{

% from social evol to urban evol?

}

\section{Specific comments}

% Linear reading

\subsection{Part I}

\subsubsection{Introduction}

% - cities major settlement form? cf Mike: before cities - after? - transition?
% - cit new Science of cities: not sure Mike meant in such an integrative way - cf preprint Elsa?
% - no consensus: better this way? cd urban perspective. Can it be a true unified explanatory model?
% beg. p4: theories and models?
% p5: concept vs variable - what exactly is a variable here?
% Q: can it be "everything"? "different form of urban life" => examples with no evolutionary processes?
% on the need of interdisciplinarity: particularly agree - urb persp

\sframe{Part I - Introduction}{

\begin{itemize}
	\item \cite{batty2013new} one view on urban systems; integrative approaches see \cite{lobo2020urban}
	\item Link and distinction between theories and models? see \cite{livet2010ontology} \cite{raimbault2017applied}
	\item Should urban evolution encompass ``everything urban''? $\rightarrow$ examples on the opposite of evolutionary processes?
	\item On the need for interdisciplinarity and multiple perspectives, totally agree: see \cite{pumain2020conclusion}
\end{itemize}

}

\subsubsection{Literature review}

% Urban ecology : // industrial symbiosis?
% Complex adaptive systems: indeed morphogenesis models are not models of urban evolution - confusion in concepts.
% Scaling: allometric growth: do not cite Chen - ! diff in time / static - Q of ergodicity

% bio -> social: units become themselves more complex, heterogenous, fuzzy.

\sframe{Part I - Literature review}{

\begin{itemize}
	\item On urban ecology, see also industrial symbiosis: analogy or equivalence? \cite{raimbault2020spatial}
	\item On Complex Adaptive Systems approach: indeed morphogenesis models are not evolution models - confusion of concepts in the literature? (as in evolutionary economic geography in general? \cite{schamp2010notion})
	\item On Scaling Theories: allometric growth/scaling is linked to scaling but not the same concept - \cite{chen2009analytical} \cite{lang2019reinvestigating}; issue of non-ergodicity \cite{pumain2012multi}
	\item When switching from biological to social evolution, evolution units become more complex, heterogenous, fuzzy.
\end{itemize}

}

\subsubsection{Model outline}

% disagreements analogy or more universal process? -> importance of understanding epistemo aspects of transfers of concepts / universal processes (structural analogy of models/theories?) / nature of interdisciplinarity
% p16: what is a sociocultural entity is fuzzy (at this stage?)

% Proposing physical entities as elementary units is challenging:
%.  - link form and function?
%.  - indeed encode information on how to best organize space, but multi-obj optim by multiple agents (so could not be "simple results of selection"?) -> OK p19 "rational without necessarily being optimized to the interests of any human actors"
%.  - def of elementary unit, where to draw the boundary in terms of nature and delimitation - how to define the system - scales?
%.  - functions / informational processes (cit info in complex systems) may be missed? -> ok function ~ encoded: "groups and activities towards which they are oriented" "a way of physically organizing space for some sets of activities and groups" "Be made out of this stuff arranged in this way, for doing these things for groups of people like this"
%   - Q: def of "socially organized"?
%  - Q of scale: micro level of representation only? - idea of a "molar unit": not sure it is useful/necessary?
%  - levels of interaction and strong emergence "After all, people do not interact with such abstract entities" - they do: admin entities, state, etc.

\sframe{Model outline}{

\begin{itemize}
	\item Evolution as an analogy or as underlying general processes? $\rightarrow$ transfer of concepts/theories/models between disciplines - nature of interdisciplinarity
	\item What exactly is a ``socio-cultural entity'' is still fuzzy at this stage
\end{itemize}

Proposing ``physical'' urban entities is challenging on several points:

\begin{itemize}
	\item Link form-function? (included in definition of formeme)
	\item Selected as optimizing spatial organization - but fitness functions in social/urban environments are tied to multiple actors at multiple scales, with multiple objectives	
	\item How to draw the boundary for units (nature, scale, delimitation), how to define the system
	\item Information essential in complex systems (included in the formeme) \cite{haken2006information}
	\item Why necessarily a ``molar unit''? Can it span multiple scales?
	\item Compatibility with weak/strong emergence? (people interact with higher level entities: state, municipal council) \cite{bedau2000open}
\end{itemize}

}

\subsubsection{Conclusion}

% origin of cities? yes and no -> cf ALife strongly interested in origins of life - these must be evolutionary (prebiotic theories: chemistry evolution) (see comment on Walker, transitions/information)
% reductionism/determinism: totally agree (complexity of evolution)
% not "teleological": should always be for some agents/utility function? (cf Monod \cite{monod2014hasard}
% from the bottom-up - daily life: agree
% mimicry of bio evol? totally agree - cf comment on transfer of concepts between disciplines
% not anti planning: complexity of managing/ designing complex systems from the bottom up: cf field of morphogenetic engineering \cite{doursat2012morphogenetic}

\sframe{Part I: conclusion}{

\begin{itemize}
	\item Origin and cities: 
	\item Not ``teleological''
\end{itemize}

}

\subsection{Part II}

\subsubsection{Introduction}

% signal -> signals and boundaries

\section{Discussion}

% level of complexity/evolution? - successive evolutions - necessaries? // talk ALife Krakauer https://scholar.google.com/citations?user=R-K-FOwAAAAJ&hl=fr&oi=ao : ? - inevitability of life / intelligence
% \cite{cronin2016beyond} Sara Walker  life as a phase transition, life/non-life (biosignatures)
% "without an understanding of what life is, how can we approach understanding its origins" \cite{walker2017re}

% multidimensionality?

% biological metaphors / transfer of concepts: ex symbiosis, morphogenesis: epistemo work on transfer of concepts?

% Ambitious new way of revisiting link between form and function: // disc morphogenesis

% spatio-temporal aspect of evolution / co-evolution niches (cit. Holland?)

% autopoiesis: Q of system def and boundaries

\sframe{Discussion}{

}

%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[allowframebreaks]
\frametitle{References}
\bibliographystyle{apalike}
\bibliography{biblio}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2020-12-09 Docs/Seminars/20201209_UTorontoUGP/communication/slides.tex
slides_seminar_ugp

\input{header_slides.tex}

\begin{document}

\title[Urban evolution and co-evolution]{Modeling urban evolution and co-evolution in systems of cities}
\author[Raimbault]{J.~Raimbault$^{1,2,3\ast}$\\\medskip
$^{\ast}$\texttt{j.raimbault@ucl.ac.uk}
}

\institute[UCL]{$^{1}$Center for Advanced Spatial Analysis, University College London\\
$^{2}$UPS CNRS 3611 Complex Systems Institute Paris\\
$^{3}$UMR CNRS 8504 G{\'e}ographie-cit{\'e}s
}

\date[July 18th 2020]{Urban Genome Project seminar\\
University of Toronto, School of Cities\\
December 9th 2020
}

\frame{\maketitle}

\AtBeginSection[]
{
	\frame{
		\tableofcontents[currentsection, hideallsubsections]
	}
	\addtocounter{framenumber}{-1}
}

%\sframe{Contents}{
%
%	\tableofcontents
%
%}

\section{Urban evolutionary theory}

% Systems of cities are adaptive complex systems, which have been theorised and modelled from many viewpoints and by different disciplines. Understanding the processes driving their dynamics is a crucial challenge for sustainable urban and territorial planning. The concept of urban evolution has proved relevant in this regard. This presentation summarises recent results obtained within the frame of the Evolutionary Urban Theory, a geographical approach to urban systems dynamics developed for more than 20 years by Denise Pumain and collaborators. We first review the main theoretical assumptions underlying this theory and results obtained with different simulation models in previous works. We also highlight the role of new model exploration and validation practices and tools developed in that context, implemented in the OpenMOLE platform.

\sframe{Urban evolution: Urban systems and Artificial Life}{

\begin{center}
	\includegraphics[width=\linewidth]{../../../Communications/ALife2020/figures/corealife_zoom.png}
\end{center}

{\footnotesize \textit{Citation network of ALife studies of urban systems} \cite{raimbault2020cities} arXiv:2002.12926}

\medskip

\textbf{Transfer of concepts: } Urban morphogenesis, bio-inspired design, urban ecology, autopoiesis \cite{batty2009centenary}

}

\sframe{Urban Evolution}{

\justify

\textbf{Urban evolution} extending cultural evolution, cities as agents with their proper genome and evolutionary dynamics?

\medskip

\begin{center}
	\begin{columns}
	\begin{column}{0.45\linewidth}
	\includegraphics[width=0.9\textwidth]{../../../Communications/ALife2020/figures/evoltheory_scales.png}
	\end{column}
	\begin{column}{0.3\linewidth}
	\includegraphics[width=\linewidth]{figures/evolth_feedbacks.png}

	\end{column}
	\end{columns}
\end{center}

\footnotesize
\textit{An evolutionary urban theory considering cities as systems within systems of cities; co-evolving urban systems in which interactions are crucial \cite{pumain2018evolutionary} \cite{pumain1997pour} \cite{pumain2008socio}}

}

\sframe{The series of Simpop models \cite{pumain2012multi}}{

\begin{columns}

	\begin{column}{0.5\textwidth}

	\centering
	\includegraphics[width=0.75\textwidth]{../../../../../OpenMole/courses/spatialsens/figures/simpop1.png}

	\footnotesize
\textit{Simpop 1 model \cite{sanders1997simpop}}

	\medskip
	\hrule
	\centering

	\medskip

	\includegraphics[width=0.55\textwidth,height=0.3\textheight]{figures/simpop2.png}

	\footnotesize

	\textit{Simpop 2 model \cite{bretagnolle2006theory}}

	\end{column}
	\vrule{}
	\begin{column}{0.5\textwidth}

	\centering

	\includegraphics[width=0.9\textwidth]{figures/simpopseries.png}

	\footnotesize

	\textit{From SimpopLocal to SimpopClim \cite{pumain2012multi}}

	\medskip

	\hrule

	\medskip

	\includegraphics[width=0.5\textwidth]{../../../../../OpenMole/courses/spatialsens/figures/setup_synth_1_tick100.png}

	\footnotesize
	\textit{SimpopNet model \cite{schmitt2014modelisation}}

	\end{column}

\end{columns}

}

\sframe{Towards new practices: ERC Geodivercity}{

% presentation generale de Geodivercity

\begin{columns}
\begin{column}{0.4\textwidth}
	\centering
	\includegraphics[width=\textwidth]{../../../../../OpenMole/courses/spatialsens/figures/urban-dynamics-simulation-models-geodivercity.png}
\end{column}
\begin{column}{0.6\textwidth}

	$\rightarrow$ Recurrent stylized facts on main systems of cities

	$\rightarrow$ Construction of simulation models (with an explicative purpose)

	$\rightarrow$ Tools and methods to explore simulation models

	\bigskip

	%\includegraphics[width=\textwidth]{../../../../../OpenMole/courses/spatialsens/figures/openmole.png}
	\includegraphics[height=0.13\textheight]{../../../Communications/ALife2020/figures/iconOM.png}
	\includegraphics[height=0.13\textheight]{../../../Communications/ALife2020/figures/openmole.png}

\end{column}
\end{columns}

}

\sframe{Theories and models}{

Historical succession of epistemologies in the case of systems of cities
\cite{varenne2018theories}:

\medskip

\begin{enumerate}
    \item Deduction from theory (top-down): Christaller
	\item Induction from the empirical (bottom-up): Berry
	\item Towards an abductive epistemology (iterative interaction theoretical-empirical): Pumain
\end{enumerate}

\medskip

$\rightarrow$ simulation allows synthesis

}

\sframe{Geographical systems and complexity}{

\textit{Necessity of simulation models in geography induced by complexities of these systems ?}

\bigskip

\begin{itemize}
	\item Ontological complexity \cite{pumain2003approche}
	\item Dynamical complexity: non-ergodicity and path-dependency \cite{pumain2012urban}
	\item Complexity and co-evolution \cite{raimbault2019modeling}
	\item Complexity and emergence \cite{bedau2002downward}
\end{itemize}

}

\sframe{Construction of Knowledge across Domains}{

\centering

\includegraphics[height=0.9\textheight]{../../../../../OpenMole/courses/spatialsens/figures/openmoleslide}

\nocite{baffi:tel-01389347}
\nocite{pumain2008socio}
\nocite{reuillon2013openmole}
\nocite{cottineau2015modular}
\nocite{swerts2017database}
\nocite{pumain1997pour}
\nocite{pumain2010theorie}

}

\sframe{Bibliometrics of the Evolutionary Urban Theory}{

\begin{center}
\includegraphics[height=0.7\textheight]{../../../../../OpenMole/courses/spatialsens/figures/evolurbantheory.pdf}
\end{center}

\textit{Citation network analysis of key publications}

\smallskip

\footnotesize
Raimbault, J. (2017). An Applied Knowledge Framework to Study Complex Systems. In Complex Systems Design \& Management (pp. 31-45).

\nocite{raimbault2017applied}

}

\sframe{OpenMOLE software}{

\justify

Urban simulation models integrated into the OpenMOLE open source software for model exploration and validation

\begin{center}
\includegraphics[height=0.13\textheight]{../../../Communications/ALife2020/figures/iconOM.png}
\includegraphics[height=0.13\textheight]{../../../Communications/ALife2020/figures/openmole.png}
\end{center}

\textit{Enables seamlessly (i) model embedding; (ii) access to HPC resources; (iii) exploration and optimization algorithms}

\medskip

\url{https://openmole.org/}

\medskip

\textbf{Applications for the summer school} (May 30th-June 4th 2021) are open!  \url{https://exmodelo.org/}

\medskip

\footnotesize

Reuillon, R., Leclaire, M., \& Rey-Coyrehourcq, S. (2013). OpenMOLE, a workflow engine specifically tailored for the distributed exploration of simulation models. Future Generation Computer Systems, 29(8), 1981-1990.

\nocite{reuillon2013openmole}

}

\sframe{Calibration of SimpopLocal model}{

\footnotesize
\textit{SimpopLocal model calibrated with distributed NSGA2 on grid}

%\medskip
%Formalising the expectations
%\textit{Formalising the expectations as indicators: }

\begin{center}
	\includegraphics[width=0.8\textwidth]{../../../../../OpenMole/courses/spatialsens/figures/slocal_fitness.png}
\end{center}

\smallskip

\tiny

Schmitt, C., Rey-Coyrehourcq, S., Reuillon, R., \& Pumain, D. (2015). Half a billion simulations: Evolutionary algorithms and distributed computing for calibrating the SimpopLocal geographical model. Environment and Planning B: Planning and Design, 42(2), 300-315.

\nocite{schmitt2015half}

}

\sframe{SimpopLocal and calibration profile}{

\textit{Computes the best calibration at fixed steps along one dimension.}

 \medskip

 \includegraphics[width=0.48\textwidth]{../../../../../OpenMole/courses/spatialsens/figures/PdiffusionZ.png}
 \includegraphics[width=0.48\textwidth]{../../../../../OpenMole/courses/spatialsens/figures/InnovationLife.png}

 \bigskip

 \footnotesize

Reuillon, R., Schmitt, C., De Aldama, R., \& Mouret, J. B. (2015). A new method to evaluate simulation models: the calibration profile (cp) algorithm. Journal of Artificial Societies and Social Simulation, 18(1), 12.

 \nocite{reuillon2015new}

}

\sframe{Unicity of mechanisms: multi-modeling}{

\footnotesize
\textit{Automate the confrontation of alternative hypothesis / mechanisms}

\medskip

	\begin{center}
	\includegraphics[width=0.9\textwidth]{../../../../../OpenMole/courses/spatialsens/figures/simfamily.png}
	\end{center}

\tiny

Cottineau, C., Reuillon, R., Chapron, P., Rey-Coyrehourcq, S., \& Pumain, D. (2015). A modular modelling framework for hypotheses testing in the simulation of urbanisation. Systems, 3(4), 348-377.

\nocite{cottineau2015modular}

}

\sframe{Multi-modeling (64 models)}{

%64 models

\centering

\includegraphics[width=0.6\textwidth]{../../../../../OpenMole/courses/spatialsens/figures/marius_complexification.png}

}

\sframe{Exemple of concurrent hypothesis}{

%Exchange mechanism:

%Market based
%Centrally planned

\textbf{Exchange mechanism: } market vs centralized

\medskip

%City growth mechanism:
%Purely inter-urban interactions
%Influenced by environmental situation

\textbf{City growth: } interurban interactions vs environmental situation

\bigskip
\bigskip

\centering

\includegraphics[width=0.5\textwidth]{../../../../../OpenMole/courses/spatialsens/figures/system_mec.png}

}

\sframe{Calibration of model family}{

%
\textit{Compute the best set of parameters for all 64 models.}

\bigskip

\centering

\includegraphics[width=\textwidth]{../../../../../OpenMole/courses/spatialsens/figures/varius_big.png}

% http://shiny.parisgeo.cnrs.fr/VARIUS/

}

\sframe{Other example of multi-modeling}{

%\textit{Benchmark of growth models for systems of cities}

\begin{center}
\includegraphics[width=0.8\textwidth]{../../../../../UrbanGrowth/Docs/Papers/EvolutionaryTheory/paper/figures/Fig5}
\end{center}

\footnotesize

Raimbault, J.; Denis, E.; Pumain, D. (2020). Empowering Urban Governance through Urban Science: Multi-Scale Dynamics of Urban Systems Worldwide. Sustainability 2020, 12, 5954.

\nocite{Raimbault_2020}

}

\sframe{PSE algorithm on the MARIUS model}{

\textit{Diversity of urban systems dynamics produced by the MARIUS model}

\begin{center}
\includegraphics[width=0.5\textwidth]{../../../../../OpenMole/courses/spatialsens/figures/pse_marius.png}
\end{center}

\footnotesize

Chérel, G., Cottineau, C., \& Reuillon, R. (2015). Beyond corroboration: Strengthening model validation by looking for unexpected patterns. PloS one, 10(9), e0138212.

\nocite{10.1371/journal.pone.0138212}

}

\sframe{Spatial sensitivity analysis}{

% space matters

\begin{center}
	\includegraphics[width=0.7\linewidth]{figures/spacematters.png}
\end{center}

\medskip

\tiny

Raimbault, J., Cottineau, C., Le Texier, M., Le Nechet, F., \& Reuillon, R. (2019). Space Matters: Extending Sensitivity Analysis to Initial Spatial Conditions in Geosimulation Models. Journal of Artificial Societies and Social Simulation, 22(4).

\nocite{raimbault2019space}

}

\sframe{Spatial sensitivity analysis: synthetic data}{

% spatial synthetic data

\begin{center}
	\includegraphics[width=0.28\linewidth]{../../../../../UrbanDynamics/Docs/Seminars/20201007_CasaST/synthdata.jpg}\hspace{0.2cm}
	\includegraphics[width=0.6\linewidth]{../../../../../UrbanDynamics/Docs/Seminars/20201007_CasaST/spatialsens_calib.png}
\end{center}

\bigskip

\footnotesize

Raimbault, J. (2019). Second-order control of complex systems with correlated synthetic data. Complex Adaptive Systems Modeling, 7(1), 1-19.
\nocite{raimbault2019second}

\medskip

Raimbault, J., \& Perret, J. (2019). Generating urban morphologies at large scales. In Artificial Life Conference Proceedings (pp. 179-186).
\nocite{raimbault2019generating}

}

\section{A model of urban evolution based on innovation diffusion}

% We then describe a model of urban evolution positioned in this approach, in which the diffusion of innovations between cities captures transformation processes (mutations) and transmission processes (diffusion), using two coupled spatial interaction models. Explorations of the model on synthetic systems of cities show the role of spatial interaction and innovation diffusion ranges on measures of diversity and utility, and the existence of intermediate interaction ranges yielding an optimal util- ity. Multi-objective optimization shows how the model produces a compromise between utility and diversity.

\sframe{Innovation diffusion}{

%\justify

$\rightarrow$ \textbf{Innovation diffusion} is a crucial process in artificial life evolutionary systems and open-ended evolution \cite{bedau2000open}

\medskip

$\rightarrow$ Artificial societies used to study the dynamics of innovation \cite{zenobia2009artificial}

\medskip

$\rightarrow$ Innovations diffuse hierarchically in systems of cities \cite{hagerstrand1968innovation}, potential explanation of urban scaling laws \cite{pumain2006evolutionary}

\bigskip

\textit{Innovation diffusion as a privileged entry to understand urban evolution}

}

\sframe{Model objective}{

$\rightarrow$ Concepts of urban evolution do not necessarily capture essential processes (transmission and transformation) in the literature

\medskip

$\rightarrow$ Need for simple models with explicit urban genome at the system of cities scale

\bigskip

\textbf{Research objective: }

\medskip

\textit{Describe and systematically explore an urban evolution model based on innovation diffusion, for urban dynamics at the macroscopic scale}

\bigskip
\bigskip

\footnotesize

Raimbault, J. (2020). A model of urban evolution based on innovation diffusion. Artificial Life Conference Proceedings 2020 NO. 32, 500-508.

}

\sframe{Model rationale}{

\begin{itemize}
	\item Agents are cities, macroscopic scale (regional, country, continental) and long time scales (century)
	\item Cities characterized by their size in terms of population; genome as adoption proportions of innovations (social or technological) for each city (one single dimension to simplify)
	\item Following \cite{favaro2011gibrat}, attractivity of cities due to level of innovation drive their population growth through spatial interactions; innovation diffuse through an other spatial interaction model \cite{fotheringham1989spatial}
	\item Mutations occur in cities as new innovations appear
\end{itemize}

}

\sframe{Model description}{

% step by step scheme ?

\begin{pspicture}(0,-2.94)(11.17,2.94)
\definecolor{colour0}{rgb}{1.0,0.2,0.0}
\definecolor{colour1}{rgb}{0.2,0.8,0.0}
\definecolor{colour2}{rgb}{0.0,0.8,1.0}
\pscircle[linecolor=black, linewidth=0.04, dimen=outer](0.8,0.26){0.8}
\pscircle[linecolor=black, linewidth=0.04, dimen=outer](3.44,-1.9){0.38}
\pscircle[linecolor=black, linewidth=0.04, dimen=outer](4.64,2.08){0.5}
\pscircle[linecolor=black, linewidth=0.04, dimen=outer](6.72,-2.12){0.48}
\rput[bl](0.12,1.1){\textit{City 1}}
\rput[bl](4.16,2.64){\textit{City 2}}
\rput[bl](2.88,-2.66){\textit{City 3}}
\rput[bl](6.16,-2.94){\textit{City 4}}
\pscircle[linecolor=black, linewidth=0.04, dimen=outer](7.56,2.48){0.44}
\pscircle[linecolor=black, linewidth=0.04, dimen=outer](7.56,2.38){0.34}
\pscircle[linecolor=black, linewidth=0.04, dimen=outer](7.56,2.28){0.22}
\rput[bl](8.18,2.18){Population}
\psframe[linecolor=black, linewidth=0.04, fillstyle=solid,fillcolor=colour0, dimen=outer](7.66,1.56)(7.4,1.22)
\psframe[linecolor=black, linewidth=0.04, fillstyle=solid,fillcolor=colour1, dimen=outer](7.66,1.74)(7.4,1.54)
\psframe[linecolor=black, linewidth=0.04, fillstyle=solid,fillcolor=colour2, dimen=outer](7.66,1.94)(7.4,1.72)
\rput[bl](8.14,1.46){Innovation genome}
\psframe[linecolor=black, linewidth=0.04, fillstyle=solid,fillcolor=colour0, dimen=outer](0.82,0.2)(0.56,-0.38)
\psframe[linecolor=black, linewidth=0.04, fillstyle=solid,fillcolor=colour0, dimen=outer](4.72,2.2)(4.52,1.72)
\psframe[linecolor=black, linewidth=0.04, fillstyle=solid,fillcolor=colour0, dimen=outer](6.74,-1.94)(6.54,-2.44)
\psframe[linecolor=black, linewidth=0.04, fillstyle=solid,fillcolor=colour0, dimen=outer](3.48,-1.8)(3.3,-2.16)
\psframe[linecolor=black, linewidth=0.04, fillstyle=solid,fillcolor=colour1, dimen=outer](0.82,0.58)(0.56,0.18)
\only<2->{
\psframe[linecolor=black, linewidth=0.04, fillstyle=solid,fillcolor=colour1, dimen=outer](4.72,2.4)(4.52,2.18)
\psframe[linecolor=black, linewidth=0.04, fillstyle=solid,fillcolor=colour1, dimen=outer](3.48,-1.6)(3.3,-1.82)
\psframe[linecolor=black, linewidth=0.04, fillstyle=solid,fillcolor=colour1, dimen=outer](6.74,-1.76)(6.54,-1.98)
\psline[linecolor=colour1, linewidth=0.04, linestyle=dashed, dash=0.17638889cm 0.10583334cm, arrowsize=0.05291667cm 2.0,arrowlength=1.4,arrowinset=0.0]{<-}(4.08,2.08)(1.4,0.84)
\psline[linecolor=colour1, linewidth=0.04, linestyle=dashed, dash=0.17638889cm 0.10583334cm, arrowsize=0.05291667cm 2.0,arrowlength=1.4,arrowinset=0.0]{<-}(6.16,-2.1)(1.48,-0.14)
\psline[linecolor=colour1, linewidth=0.04, linestyle=dashed, dash=0.17638889cm 0.10583334cm, arrowsize=0.05291667cm 2.0,arrowlength=1.4,arrowinset=0.0]{<-}(3.06,-1.74)(1.2,-0.44)
\psline[linecolor=colour1, linewidth=0.04, linestyle=dashed, dash=0.17638889cm 0.10583334cm, arrowsize=0.05291667cm 2.0,arrowlength=1.4,arrowinset=0.0]{<-}(7.98,0.5)(7.1,0.5)
\rput[bl](8.14,0.4){Innovation diffusion}
}
\only<3->{
\psline[linecolor=black, linewidth=0.08, arrowsize=0.05291667cm 2.0,arrowlength=1.4,arrowinset=0.0]{<->}(1.56,0.5)(4.18,1.74)
\psline[linecolor=black, linewidth=0.16, arrowsize=0.05291667cm 2.0,arrowlength=1.4,arrowinset=0.0]{<->}(1.6,0.04)(6.2,-1.88)
\psline[linecolor=black, linewidth=0.08, arrowsize=0.05291667cm 2.0,arrowlength=1.4,arrowinset=0.0]{<->}(4.88,1.62)(6.46,-1.6)
\psline[linecolor=black, linewidth=0.04, arrowsize=0.05291667cm 2.0,arrowlength=1.4,arrowinset=0.0]{<->}(1.44,-0.32)(3.16,-1.54)
\psline[linecolor=black, linewidth=0.04, arrowsize=0.05291667cm 2.0,arrowlength=1.4,arrowinset=0.0]{<->}(3.86,-1.94)(6.2,-2.22)
\psline[linecolor=black, linewidth=0.04, arrowsize=0.05291667cm 2.0,arrowlength=1.4,arrowinset=0.0]{<->}(7.06,0.96)(7.96,0.96)
\psline[linecolor=black, linewidth=0.08, arrowsize=0.05291667cm 2.0,arrowlength=1.4,arrowinset=0.0]{<->}(3.7,-1.5)(4.62,1.5)
\rput[bl](8.12,0.84){Population growth}
}
\only<4->{
\psframe[linecolor=black, linewidth=0.04, fillstyle=solid,fillcolor=colour2, dimen=outer](0.82,0.94)(0.56,0.56)
\pscircle[linecolor=colour2, linewidth=0.04, linestyle=dashed, dash=0.17638889cm 0.10583334cm, dimen=outer](0.81,0.27){0.72}
\pscircle[linecolor=colour2, linewidth=0.04, linestyle=dashed, dash=0.17638889cm 0.10583334cm, dimen=outer](7.6,-0.08){0.26}
\rput[bl](8.14,-0.18){Mutation}
}
\end{pspicture}

}

\sframe{Model formalization}{

\footnotesize

At each time step, with $P_i \left(t\right)$ population, $\delta_{c,i}\left(t\right)$ genome, $u_c$ utility of innovation, $p_{c,i,t}$ share of total population adopting innovation $c$ in city $i$
\begin{enumerate}
	\item Crossover through the diffusion of innovations
	\[
	\delta_{c,i,t} \textrm{ = } \frac{\sum_j p_{c,j,t-1}^{\frac{1}{u_c}} \cdot \exp{\left(-\frac{d_{ij}}{d_I}\right)}}{\sum_c \sum_j p_{c,j,t-1}^{\frac{1}{u_c}} \cdot \exp{\left(-\frac{d_{ij}}{d_I}\right)}}	
	\]
	\item Population growth through spatial interactions $P_i\left(t\right) - P_i\left(t-1\right) \textrm{ = } w_I\cdot \sum_j \frac{V_{ij}}{<V_{ij}>}$ with
	\[ 
	V_{ij} \textrm{ = } \frac{P_{i}\left(t-1\right) \cdot P_{j}\left(t-1\right)}{\left(\sum_k P_k\left(t-1\right)\right)^2} \cdot \exp{\left(-\frac{d_{ij}}{d_G} \cdot \prod_c \delta_{c,i,t}^{\phi_{c,t}}\right)}
	\]
	and $\phi_{c,t} \textrm{ = } \sum_i \delta_{i,c,t}\cdot P_i\left(t-1\right) /\sum_{i,c} \delta_{i,c,t}\cdot P_{i}\left(t-1\right)$
	\item Mutations with innovations introduced with probability $\beta \cdot \left(P_i \left(t\right) / \max_k P_k \left(t\right)\right)^{\alpha_I}$ and an initial penetration rate $r_0$; new utility $u_c$ randomly distributed (normal or log-normal) with average current average utility and standard deviation a given parameter $\sigma_U$
\end{enumerate}

}

\sframe{Model indicators}{

\begin{itemize}
	\item Average diversity
	\[
	D \textrm{ = } \frac{1}{t_f \textrm{+} 1} \sum_{t\textrm{=}0}^{t_f} \left(1 - \sum_{i,c} \left(p_{c,i,t}\right)^2 \right)	
	\]
	\item Average utility
	\[
	U \textrm{ = } \frac{1}{t_f \textrm{ + } 1} \sum_{t \textrm{=}0}^{t_f} \sum_{i,c} \delta_{c,i,t} u_c 
	\]
	\item Innovatitivity 
	\[
	I \textrm{ = } \frac{\max c}{N\cdot \left(t_f \textrm{ + } 1\right)}
	\]
	\item Population trajectories, summarized by final hierarchy \cite{raimbault2020unveiling}
\end{itemize}

}

\sframe{Synthetic configurations}{

Model applied on synthetic systems of cities (so that conclusions are independent of geographical contingencies \cite{raimbault2019space}):

\medskip

\begin{itemize}
	\item random positions and rank-size hierarchy $P_i \left(0\right) \textrm{ = } \frac{P_{max}}{i^{\alpha_0}}$ with $\alpha_0 \textrm{ = } 1.0$ and $P_{max} \textrm{ = } 100,000$
	\item regional urban system scale: $N \textrm{ = } 30$ cities
	\item simulated for $t_f \textrm{ = } 50$ macroscopic time steps (order of magnitude of a century)
\end{itemize}

}

\sframe{Model parameters}{

\centering

	\begin{tabular}{|l|l|l|l|l|l|}
	\hline
	Parameter & Not. & Process & Range & Def. \\ \hline
	Number of cities & $N$ & Spatial scale & $[10 ; 100]$ & $30$\\
	Initial hierarchy & $\alpha_0$ & System of cities & $[0.5 ; 2.0]$ & $1$\\
	Initial population & $P_{max}$ & System of cities & $[10^4 ; 10^7]$ & $10^5$\\
	Simulation steps & $t_f$ & Temporal scale & $[10 ; 100]$ & $50$\\
	Growth rate & $w_I$ & Pop. growth & $[0.001 ; 0.01]$ & $0.005$\\\hline
	Gravity range & $d_G$ & Crossover & $[0 ; 2]$ & $1$\\
	Innovation range & $d_I$ & Crossover & $[0 ; 2]$ & $1$\\
	Innovation rate & $\beta$ & Mutation & $[0 ; 1]$ & $0.5$\\
	Innovation hierarchy & $\alpha_I$ & Mutation & $[0 ; 2]$ & $1$\\
	Innov. utility std. & $\sigma_U$ & Mutation & $\left[ 0.7;2\right]$ & $1$\\
	Penetration rate & $r_0$ & Mutation & $\left[0.1;0.9\right]$ & $0.5$\\
	Utility type & - & Mutation & \{n;ln\}& ln \\
	\hline
	\end{tabular}

% Model implemented in \texttt{scala}; relatively large parameter space

}

\sframe{Statistical consistency}{

\begin{itemize}
	\item Latin Hypercube Sampling of 100 parameter points, 1000 replications for each
	\item Sharpe ratios have high values for all indicators and all parameters (minimum 1.7 for utility)
	\item Average and median relative distances defined as $\Delta_{ij} \textrm{ = } 2\frac{\left|\mu_i - \mu_j \right|}{\sigma_i \textrm{ + } \sigma_j}$ larger than one for all indicators: 50 repetitions in further experiments
\end{itemize}

}

\sframe{Model exploration: diversity}{

Grid sampling of the parameter space (23,168 points, 50 replications) with a finer grid on $d_G$ and $d_I$; plots shown at $\alpha_I \textrm{ = } 1$ and $\sigma_U \textrm{ = } 1$

\medskip

\begin{center}
\includegraphics[width=0.7\linewidth]{../../../Communications/ALife2020/figures/averageDiversity-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_utilityStd1_distriblog-normal}
\end{center}

\smallskip

\footnotesize
\textit{Diversity increases with interaction span with a plateau behavior, decreases with innovation diffusion span}

}

\sframe{Model exploration: innovation and hierarchy}{

\begin{center}
\includegraphics[width=0.47\linewidth]{../../../Communications/ALife2020/figures/averageInnovation-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_utilityStd1_distriblog-normal.png}
\includegraphics[width=0.47\linewidth]{../../../Communications/ALife2020/figures/finalHierarchy-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_utilityStd1_distriblog-normal.png}
\end{center}

\smallskip

\footnotesize

\textit{Systems with more interactions and diffusion are less unequal and innovate more}

}

\sframe{Model exploration: utility}{

\begin{center}
\includegraphics[width=0.8\linewidth]{../../../Communications/ALife2020/figures/averageUtility-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_utilityStd1_distriblog-normal.png}
\end{center}

\footnotesize

\textit{Piecewise behavior for low innovation rates; maximum as a function of $d_G$ for high innovation: emergence of regional innovation clusters?}

}

\sframe{Correlations}{

\begin{center}
	\includegraphics[width=0.6\linewidth]{../../../Communications/ALife2020/figures/corrmat_cropped.png}	
\end{center}

\footnotesize

\textit{Correlation matrix estimated over the whole exploration: innovation and population are not strongly correlated; 91\% of variance on first two components}

}

\sframe{Model optimization}{

\begin{center}
	\includegraphics[width=0.6\linewidth]{../../../Communications/ALife2020/figures/paretoDiversity-Utility_colorinnovationDecay.png}
\end{center}

\footnotesize

\textit{NSGA2 algorithm to simultaneously optimize utility and diversity: emergence of three compromize regimes; intermediate regime with low level of innovation diffusion}

}

\sframe{Discussion}{

\textbf{Empirical and theoretical implications}

\smallskip

$\rightarrow$ Global integration of cities is not necessarily optimal in terms of overall utility

\smallskip

$\rightarrow$ Urban evolution simulation model including explicit evolution processes and an urban genome

\medskip

\textbf{Future work and extensions}

\smallskip

$\rightarrow$ Multi-dimensional urban genome to capture multi-dimensionality of urban dynamics \cite{hidalgo2007product}

\smallskip

$\rightarrow$ Application to real systems of cities \cite{2020arXiv200510007R}: patent data as possible proxy for innovation dynamics \cite{bergeaud2017classifying}

\smallskip

$\rightarrow$ Processes at other scales, towards multi-scale models \cite{raimbault:halshs-02351722}

%  A simple model of urban evolution capturing complex dynamics at the macroscopic scale through the diffusion of innovations

%$\rightarrow$ An ALife approach to the simulation of urban systems: ``Cities as they could be''

}

\section{Co-evolution in urban systems}

% The last part of the presentation focuses on the concept of co-evolution in urban systems, proposing a specific definition, a method to characterise it based on circular causations, and an application to the co-evolution between transportation networks and territories. For this particular case, we summarise results from co-evolution models at the mesoscopic scale (co-evolution between urban form and road networks) and at the macroscopic scale (co-evolution between cities and urban networks).

\sframe{Interactions between networks and territories}{

\justify

\begin{center}
\includegraphics[width=0.45\linewidth]{../../../../../CityNetwork/Docs/Communications/2019/TQGDebates2019/figures/accessp_withbridge_prd_EN.png}
\hspace{0.1cm}
\includegraphics[width=0.52\linewidth]{../../../../../CityNetwork/Docs/Communications/2019/TQGDebates2019/figures/avgaccess_facet.png}

\end{center}

\medskip

%\vspace{-0.5cm}

%\begin{justify}
\textit{Accessibility as part of complex processes of co-evolution between transportation networks and territories.}

%\end{justify}

\nocite{raimbault2018evolving}

\medskip

\tiny

Raimbault, J. (2019). Evolving accessibility landscapes: mutations of transportation networks in China. In Aveline-Dubach, N., ed. \textit{Pathways of sustainable urban development across China - the cases of Hangzhou, Datong and Zhuhai}, pp 89-108. Imago. ISBN:978-88-94384-71-0

}

\sframe{Definition of co-evolution \cite{raimbault2019modeling}}{

\footnotesize

  \textbf{Objects: }

  \begin{itemize}
  	\item Cities and territories seen from the \textit{Evolutionary Urban Theory} viewpoint
  	\item Transportation networks as realisation of ``transactional projects'', following the \textit{Territorial Theory of Networks} \cite{dupuy1987vers}
  \end{itemize}

 \bigskip

\uncover<2->{

\textbf{Processes: }

\textit{A three-level definition for co-evolution: } 
\begin{enumerate}
	\item \textcolor{blue}{at the agent level}
	\item \textcolor{green}{at the agent population level (niches)}
	\item \textcolor{red}{at the global system level}
\end{enumerate}

}

\bigskip

\uncover<3->{

\textbf{Corresponding approaches: }

\begin{enumerate}
	\item \textcolor{blue}{Empirical studies (microscopic level)}
	\item \textcolor{green}{Urban morphogenesis modeling (niche level)}
	\item \textcolor{red}{Urban evolutionary theory models (macroscopic level)}
\end{enumerate}

\smallskip

\tiny
Raimbault, J. (2019). Modeling interactions between transportation networks and territories: a co-evolution approach. arXiv preprint arXiv:1902.04802.

}

}

\sframe{Measuring co-evolution}{

\begin{center}
\includegraphics[width=\linewidth]{../../../../../CityNetwork/Docs/Seminars/20180109_LAET/figures/causality_regimes}
\end{center}

\medskip

\footnotesize

Raimbault, J. (2017). Identification de causalit{\'e}s dans des donn{\'e}es spatio-temporelles. SAGEO 2017 Proceedings. \textit{Translated as arXiv:1709.08684}
\nocite{raimbault2017identification}

}

\sframe{Illustration of the method}{

\centering

\includegraphics[height=0.9\textheight]{../../../../../CityNetwork/Docs/Seminars/20180109_LAET/figures/causality_twovars.pdf}

}

\sframe{Method validation}{

\includegraphics[width=\linewidth]{../../../../../CityNetwork/Docs/Seminars/20180109_LAET/figures/4-2-2-fig-causalityregimes-arma.jpg}

\medskip

\textit{Synthetic data: auto-regressive process with lag 2, cross terms parametrized by random $(a_1,a_2) \in [-0.1,0.1]$.}

}

\sframe{Application on model of \cite{raimbault2014hybrid}}{

\includegraphics[width=\textwidth]{../../../../../CityNetwork/Docs/Seminars/20180109_LAET/figures/regimes_1.png}

}

\sframe{Application on \textit{Grand Paris Express}}{

\begin{center}
\includegraphics[width=0.55\textwidth]{../../../../../CityNetwork/Docs/Seminars/20180109_LAET/figures/reseaux}
\includegraphics[width=0.45\textwidth]{../../../../../CityNetwork/Docs/Seminars/20180109_LAET/figures/1-2-1-fig-casestudies-empiricalres}
\end{center}

\smallskip

\textit{Anticipated effects on the new infrastructure on real estate prices}

}

\sframe{Modeling contributions}{

	\textbf{Macroscopic scale:}
	\begin{itemize}
		\item Interaction models between cities including transportation networks		
		$\rightarrow$ \textit{Evidence of network effects; exploration of interaction regimes}
	\end{itemize}

	\bigskip

	\uncover<2->{
	\textbf{Mesoscopic scale:}
	\begin{itemize}
		\item Morphogenesis model coupling urban form and network

		 $\rightarrow$ \textit{Complementarity of multiple processes; calibration at the first and second order}
		\item Exploration of an extended LUTI model including transportation governance
	\end{itemize}
}

}

\sframe{Form and function in territories}{

\textit{Which ontology to include more complex functional properties in mesoscopic morphogenesis models?}

\medskip

$\rightarrow$ Territorial systems as the strong coupling between territories and (potential and realized) networks \cite{dupuy1987vers}.

\medskip

$\rightarrow$ Networks convey functional notions of centralities and accessibility, among others; have furthermore proper topological properties.

}

\sframe{A Morphogenesis Model of co-evolution}{

\justify

\vspace{-1cm}

$\rightarrow$ Coupled grid population distribution and vector transportation network, following the core of \cite{raimbault2014hybrid}

\medskip

$\rightarrow$ Local morphological and functional variables determine a patch-value, driving new population attribution through preferential attachment; combined to population diffusion (reaction-diffusion processes studied before)

\medskip

$\rightarrow$ Network growth is also driven by morphological, functional and local network measures, following diverse heuristics corresponding to different processes (multi-modeling)

\medskip

\textit{Local variables and network properties induce feedback on both, thus a strong coupling capturing the \textbf{co-evolution}}

\bigskip

{
\tiny

Raimbault, J. (2019). An urban morphogenesis model capturing interactions between networks and territories. In The Mathematics of Urban Morphology (pp. 383-409). Birkhäuser, Cham.

\medskip

Raimbault, J. (2018). Multi-modeling the morphogenesis of transportation networks. In Artificial Life Conference Proceedings (pp. 382-383).

}

}

\sframe{Model: Specification}{

\includegraphics[width=\textwidth]{../../../../../CityNetwork/Docs/Communications/2019/TQGDebates2019/figures/coevol_mesocoevol}

}

\sframe{Network Generation}{

At fixed time steps :

\begin{enumerate}
	\item Add new nodes preferentially to new population and connect them
	\item \justify Variable heuristic for new links, among: nothing, random, gravity-based deterministic breakdown, gravity-based random breakdown (from \cite{schmitt2014modelisation}), cost-benefits (from \cite{louf2013emergence}), biological network generation (based on \cite{tero2010rules})
\end{enumerate}

\medskip

\centering

\frame{\includegraphics[height=0.31\textwidth]{../../../../../CityNetwork/Docs/Communications/2019/TQGDebates2019/figures/coevol_example-bio-process-1}}\hspace{0.2cm}
\frame{\includegraphics[height=0.31\textwidth]{../../../../../CityNetwork/Docs/Communications/2019/TQGDebates2019/figures/coevol_example-bio-process-1-tick80}}

\footnotesize

\textit{Intermediate stage for biological network generation}

}

\sframe{Generated Urban Shapes: Urban Form}{

\centering

\frame{\includegraphics[width=0.28\textwidth]{../../../../../CityNetwork/Docs/Communications/2019/TQGDebates2019/figures/coevol_example_synthsetup}}\hspace{0.1cm}
\frame{\includegraphics[width=0.28\textwidth]{../../../../../CityNetwork/Docs/Communications/2019/TQGDebates2019/figures/coevol_example_form-accessonly}}\hspace{0.1cm}
\frame{\includegraphics[width=0.28\textwidth]{../../../../../CityNetwork/Docs/Communications/2019/TQGDebates2019/figures/coevol_example_form-droadonly}}\\\vspace{0.1cm}
\frame{\includegraphics[width=0.28\textwidth]{../../../../../CityNetwork/Docs/Communications/2019/TQGDebates2019/figures/coevol_example_form-bwonly}}\hspace{0.1cm}
\frame{\includegraphics[width=0.28\textwidth]{../../../../../CityNetwork/Docs/Communications/2019/TQGDebates2019/figures/coevol_example_form-closenessonly}}\hspace{0.1cm}
\frame{\includegraphics[width=0.28\textwidth]{../../../../../CityNetwork/Docs/Communications/2019/TQGDebates2019/figures/coevol_example_form-poponly}}

\footnotesize\textit{In order: setup; accessibility driven; road distance driven; betweenness driven; closeness driven; population driven.}

}

\sframe{Generated Urban Shapes: Network}{

\centering

\frame{\includegraphics[width=0.28\textwidth]{../../../../../CityNetwork/Docs/Communications/2019/TQGDebates2019/figures/coevol_example_nw-connection}}\hspace{0.1cm}
\frame{\includegraphics[width=0.28\textwidth]{../../../../../CityNetwork/Docs/Communications/2019/TQGDebates2019/figures/coevol_example_nw-random}}\hspace{0.1cm}
\frame{\includegraphics[width=0.28\textwidth]{../../../../../CityNetwork/Docs/Communications/2019/TQGDebates2019/figures/coevol_example_nw-gravity}}\\\vspace{0.1cm}
\frame{\includegraphics[width=0.28\textwidth]{../../../../../CityNetwork/Docs/Communications/2019/TQGDebates2019/figures/coevol_example_nw-rndbrkdwn}}\hspace{0.1cm}
\frame{\includegraphics[width=0.28\textwidth]{../../../../../CityNetwork/Docs/Communications/2019/TQGDebates2019/figures/coevol_example_nw-cost}}\hspace{0.1cm}
\frame{\includegraphics[width=0.28\textwidth]{../../../../../CityNetwork/Docs/Communications/2019/TQGDebates2019/figures/coevol_example_nw-bio}}

\footnotesize\textit{In order: connection; random; deterministic breakdown; random breakdown; cost-driven; biological.}

}

\sframe{Results : Network Heuristics}{

\justify

\textit{Comparison of feasible space for network indicators with fixed density}

\bigskip

\includegraphics[width=0.52\textwidth,height=0.6\textheight]{../../../../../CityNetwork/Docs/Communications/2019/TQGDebates2019/figures/coevol_feasible_space_withreal_pca_bymorph}
\includegraphics[width=0.43\textwidth,height=0.6\textheight]{../../../../../CityNetwork/Docs/Communications/2019/TQGDebates2019/figures/coevol_distance_real_bymorph}

\footnotesize

\textit{(Left) Feasible spaces by morphological class and network heuristic; (Right) Distribution of distances to topologies of real networks}

}

\sframe{Results : Calibration}{

\justify

\vspace{-0.5cm}

Calibration (model explored with OpenMole~\cite{reuillon2013openmole}, $\sim 10^6$ model runs) at the first order on morphological and topological objectives, and on correlations matrices.

\bigskip

\begin{columns}
\column{0.4\textwidth}
\centering
\includegraphics[width=\textwidth]{../../../../../CityNetwork/Docs/Communications/2019/TQGDebates2019/figures/coevol_pca_allobjs}
\column{0.2\textwidth}
\includegraphics[width=\textwidth]{../../../../../CityNetwork/Docs/Communications/2019/TQGDebates2019/figures/coevol_pca_morpho_byheuristic}\\
\includegraphics[width=\textwidth]{../../../../../CityNetwork/Docs/Communications/2019/TQGDebates2019/figures/coevol_pca_network_byheuristic}
\column{0.4\textwidth}
\includegraphics[width=\textwidth]{../../../../../CityNetwork/Docs/Communications/2019/TQGDebates2019/figures/coevol_corrs-distrib_rhoasize4}

\end{columns}

\footnotesize\textit{(Left) Full indicator space; (Middle) Morphological and Topology, by network heuristic; (Right) Distance distribution for cumulated distance for indicators and correlations.}

}

\sframe{Results: Causality Regimes}{

\footnotesize
\textit{Unsupervised learning on lagged correlations between local variables unveils a diversity of causality regimes}

$\rightarrow$ Link between \emph{co-evolution regime} and morphogenetic properties of the urban system

\medskip

\begin{center}
\includegraphics[width=0.52\textwidth,height=0.55\textheight]{../../../../../CityNetwork/Docs/Communications/2019/TQGDebates2019/figures/coevol_centertrajs}
\includegraphics[width=0.4\textwidth,height=0.55\textheight]{../../../../../CityNetwork/Docs/Communications/2019/TQGDebates2019/figures/coevol_cluster-params}
\end{center}

\footnotesize

\textit{(Left) Lagged correlation profiles of cluster centers; (Right) Distribution of regimes across parameter space}

}

\sframe{Discussion}{

\justify

\vspace{-1cm}

\textbf{Implications}

$\rightarrow$ This rather simple model reproduces most of existing urban forms in Europe for both population distribution and road network: which intrinsic dimension to the urban system and its morphological aspect ?

$\rightarrow$ Ability to reproduce static correlations and a variety of dynamical lagged correlation regimes suggests that the model captures some of the processes of co-evolution

\bigskip

\textbf{Developments}

$\rightarrow$ Towards a dynamical calibration? Need for dynamical data

$\rightarrow$ Investigate the link between spatial non-stationarity and non-ergodicity through simulation by the model

$\rightarrow$ Compare network generation in a ``fair'' way (correcting for additional parameters, open question for models of simulation)

}

\sframe{Co-evolution of cities and networks}{

% simpopnet generator

\textit{Exploration of co-evolution regimes for the SimpopNet model}

\medskip

\begin{center}
\includegraphics[height=0.65\textheight]{../../../../../OpenMole/courses/spatialsens/figures/simpopnet_Fig1.png}
\end{center}

\tiny

Raimbault, J. (2020). Unveiling co-evolutionary patterns in systems of cities: a systematic exploration of the SimpopNet model. In Theories and Models of Urbanization (pp. 261-278). Springer, Cham.
\nocite{raimbault2020unveiling}

}

\sframe{Macroscopic co-evolution model}{

%\centering

\begin{center}
\includegraphics[width=0.95\textwidth]{../../../../../OpenMole/courses/spatialsens/figures/macrocoevol_en.png}
\end{center}

\tiny

Raimbault, J. (2020). Indirect evidence of network effects in a system of cities. Environment and Planning B: Urban Analytics and City Science, 47(1), 138-155.

\smallskip

Raimbault, J. (2020). Modeling the co-evolution of cities and networks. In Niel, Z., Rozenblat, C., eds. \textit{Handbook of Cities and Network}, Edwar Elgar Publishing, \textit{in press}. arXiv:1804.09430

\nocite{raimbault2020indirect}
\nocite{raimbault2018modeling}

}

\sframe{Model application}{

\begin{center}
\includegraphics[height=0.65\textheight]{../../../../../NetworksTerritories/CoevolutionNwTerritories/Docs/Papers/HierarchyCoevolution/chapter/figures/Fig1.png}\hspace{0.2cm}
\includegraphics[height=0.65\textheight]{../../../../../NetworksTerritories/CoevolutionNwTerritories/Docs/Papers/HierarchyCoevolution/chapter/figures/Fig4.png}
\end{center}

\smallskip

\tiny

Raimbault, J. (2020). Hierarchy and co-evolution processes in urban systems. arXiv preprint arXiv:2001.11989. Forthcoming in Fen-Chong, J. ed. Centralit{\'e}s et hi{\'e}rarchies des r{\'e}seaux et des territoires, ISTE Editions.
\nocite{raimbault2020hierarchy}

}

\sframe{Co-evolution regimes}{

\begin{center}

	\includegraphics[width=0.9\linewidth]{../../../../../CityNetwork/Docs/Seminars/20190617_PrixThese/figures/macro-regimes.png}
	\end{center}

	\medskip

\vspace{-0.4cm}

	\textit{Multiple co-evolution regimes unveiled for synthetic configurations}

}

\section{Discussion}

% We finally discuss perspectives towards more complex and multi-scalar models of urban evolution.

\sframe{Towards more complex models}{

\textit{The LUTECIA model includes transportation governance for network growth \cite{le2015modeling}}

\medskip

\centering
\includegraphics[width=\linewidth]{../../../../../CityNetwork/Docs/Seminars/20190617_PrixThese/figures/meso-lutecia.jpg}

}

\sframe{Towards more complex models}{

\textit{Including multinational transportation governance into the macroscopic co-evolution model}

\medskip

\begin{center}
\includegraphics[width=\linewidth]{../../../../../Governance/Docs/Communications/2020/CCS2020/communication/figures/ex_baseline_full.png}
\end{center}

\smallskip

\footnotesize

[Presentation yesterday at Conference on Complex Systems 2020]

}

\sframe{Towards multi-scale models}{

\justify

\textit{Crucial aspect of processes at multiple scales and feedback between these \cite{pumain2008socio}; need to be taken into account to build sustainable territorial policies \cite{rozenblat2018conclusion}}

\bigskip

$\rightarrow$ Empirical analysis of contradictory sustainability indicators on endogenous European mega-city regions \cite{raimbault2019multi}

\medskip

$\rightarrow$ A parsimonious multi-scalar urban growth model coupling spatial interactions at the macroscopic scale with reaction-diffusion models for urban form at the mesoscopic scale in \cite{raimbault:halshs-02351722}

\medskip

$\rightarrow$ Include physical transportation network in macroscopic co-evolution models \cite{raimbault2020hierarchy}

}

\sframe{Conclusion}{

\footnotesize

\textit{Urban evolution and co-evolution as a powerful paradigm to model, understand, sustainably manage urban systems}

\medskip

\footnotesize

\justify

\textbf{Open repositories for models} \texttt{https://github.com/JusteRaimbault/CityNetwork}

\texttt{https://github.com/JusteRaimbault/UrbanEvolution}

\medskip

\textbf{Acknowledgments}: \textit{European Grid Infrastructure}; UKCRIC DAFNI; Urban Dynamics Lab Grant EPSRC EP/M023583/1

\medskip

\textbf{References}

\tiny

Raimbault, J. (2017). Identification of causalities in spatio-temporal data. SAGEO 2017 Proceedings. arXiv:1709.08684
\nocite{raimbault2017identification}

\smallskip

Raimbault, J. (2018). Calibration of a density-based model of urban morphogenesis. PloS one, 13(9), e0203516.
\nocite{raimbault2018calibration}

\smallskip

Raimbault, J. (2018). Multi-modeling the morphogenesis of transportation networks. In Artificial Life Conference Proceedings (pp. 382-383).
\nocite{raimbault2018multi}

\smallskip

Raimbault, J. (2019). An urban morphogenesis model capturing interactions between networks and territories. In The Mathematics of Urban Morphology (pp. 383-409). Birkhäuser, Cham.

\nocite{raimbault2019urban}

\smallskip

Raimbault, J. (2019). Modeling interactions between transportation networks and territories: a co-evolution approach. arXiv preprint arXiv:1902.04802.
\nocite{raimbault2019modeling}

\smallskip

Raimbault, J. (2020). Unveiling co-evolutionary patterns in systems of cities: a systematic exploration of the SimpopNet model. In Theories and Models of Urbanization (pp. 261-278). Springer, Cham.
\nocite{raimbault2020unveiling}

\smallskip

Raimbault, J. (2020). Indirect evidence of network effects in a system of cities. Environment and Planning B: Urban Analytics and City Science, 47(1), 138-155.
\nocite{raimbault2020indirect}

\smallskip

Raimbault, J. (2020). Modeling the co-evolution of cities and networks. Handbook of cities and networks, Rozenblat C., Niel Z., eds. (in press) arXiv:1804.09430.
\nocite{raimbault2018modeling}

\smallskip

Raimbault, J. (2020). Hierarchy and co-evolution processes in urban systems. arXiv preprint arXiv:2001.11989. Forthcoming in Fen-Chong, J. ed. Centralit{\'e}s et hi{\'e}rarchies des r{\'e}seaux et des territoires, ISTE Editions.
\nocite{raimbault2020hierarchy}

}

%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[allowframebreaks]
\frametitle{References}
\bibliographystyle{apalike}
\bibliography{biblio}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2020-12-09 Docs/Seminars/20201209_UTorontoUGP/communication/slides.sh

@@ -1,0 +1,2 @@
+pdfcompress slides.pdf UGPSeminar20201209_UToronto_Raimbault.pdf
+

### 2020-12-09 Docs/Seminars/20201209_UTorontoUGP/communication/comment.tex
comments

@@
 
-\sframe{Modeling urban evolution}{
+\sframe{General comments}{
 
 % from social evol to urban evol?
+
+\begin{itemize}
+	\item Link with geographical/economic  stylized facts at multiple scales seems to be weak - urban theory construction is generally conditioned by available data
+	\item Constructing a formal model of urban evolution is not a guarantee of explanation (``answering the why''): the actual processes corresponding to the semantics will capture the causal structure of systems
+	\item Definitions of a city, of urban systems, etc. are not constructed in strong link with the theory and with empirical observations
+	\item The model formal development, which can be specified in many ways them , is a good thing as it provides a general framework; but as it is now it leaves too much doubt on how to actually measure these processes, the signals, the diffusion of information, etc.
+	\item Does not seem to allow endogeneity and open-ended evolution
+\end{itemize}
 
@@
 
-\sframe{Model outline}{
+\sframe{Part I - Model outline}{
 
@@
 \begin{itemize}
-	\item Origin and cities: 
-	\item Not ``teleological''
+	\item Origin of cities: ALife strongly interest in Origins of Life, these must have been evolutionary (from chemical evolution to biological evolution): emergence of life as a phase transition \cite{cronin2016beyond}
+	\item Reductionism and determinism: totally agree: complexity of evolution, order out of chaos, frozen accidents
+	\item Not ``teleological'' but teleonomical: always some agents with some objective functions \cite{monod2014hasard}
+	\item From the bottom-up - daily urban life: importance of emergence
+	\item Not mimic biological evolution: see comment on transfer of concepts between disciplines
+	\item Not anti-planning: complexity of managing/designing complex systems from the bottom-up: see the field of morphogenetic engineering \cite{doursat2012morphogenetic}
 \end{itemize}
@@
 
-\subsubsection{Introduction}
-
 % signal -> signals and boundaries
 
+\sframe{Part II - Model}{
+
+\begin{itemize}
+	\item Difference between urban genome and phenotype?
+How does the genome express itself?
+The genome is defined as a set of formeme, which are already ``expressed'' through the function $f$ to define uses.
+Is the expression the link between $U$ and $H$?
+Is phenotype the signature?
+	\item The sets $P, A, G$ are fixed - does it mean that a given instantiation of the model is not able to endogenously introduce new categories, and thus not able to exhibit open-ended evolution?
+\cite{taylor2016open}
+	\item Tying the genome to a world is a good idea e.g. to account for non-ergodicity, explore ``what-if'' worlds (``Cities as they could be'' \cite{raimbault2020cities})
+	\item The genome is somehow independent of function $H$ and information propagation $S$ - could we include these in the genome, or is it a fundamental divergence point with biological evolution?
+\end{itemize}
+
+}
+
+\sframe{Part II - Characteristics, similarity, trajectories}{
+
+\begin{itemize}
+	\item To what extent already specifying precise signal characteristics corresponds to model implementation - could it stay generic?
+	\item Defining distances between genomes is indeed crucial for phylogenetic studies; estimating distances between organisms (based on proteins alignement for example) is a very difficult problem in biology; would the same difficulties transfer to urban evolution, or would some known historical data/spatial models help directly construct the phylogenies?
+(section 6 defines the tree in a way).
+	\item The link between genome distance and spatial distance is crucial in specifying the model of urban evolution (as for example in \cite{raimbault2020model})
+	\item To what extent are activity costs and recoding dependent on context and not only on the formeme?
+\end{itemize}
+
+}
+
+\subsection{Part III}
+
+\sframe{Part III}{
+
+\begin{itemize}
+	\item Innovations in formeme are physical innovations - to what extent does the model capture the evolution of concepts/ideas/knowledge? It may have very different dynamics.
+	\item Environmental variations are a way to capture spatial interactions/diffusion through signals.
+	\item Spatial interactions are also crucial in the different selection processes detailed.
+	\item On the concept of niche, what is the link between spatial niche (as it is used here) and evolutionary niche (local fitness optimum)?
+\end{itemize}
+
+}
+
+\sframe{Part IV}{
+
+\begin{itemize}
+	\item Could 6 years be too short to observe meaningful evolutionary processes?
+	\item Are numerical values for distances directly interpretable?
+	\item Is it possible to link longitudinal distance iwth transversal distance? (Extrapolate some divergence time?)
+\end{itemize}
+
+}
+
 \section{Discussion}
+
+\sframe{Successive evolutions and complexity}{
 
@@
 
-% multidimensionality?
+Emergence of life as a succession of phase transitions in an information propagation process \cite{cronin2016beyond}
+
+\medskip
+
+``\textit{Without an understanding of what life is, how can we approach understanding its origins}'' \cite{walker2017re}
+
+\medskip
+
+$\rightarrow$ Succession of chemical, biological, cultural evolution, with increasing complexities
+
+\medskip
+
+Should urban evolution be considered as a new phase transition? (see \cite{batty2018inventing}) What is its main support then? (autocatalytic sets, intelligence, ?)
+
+}
+
+\sframe{Epistemology of concept transfer}{
 
@@
 
+Differentiation between using metaphors/analogies, transferring concepts and models, and unveiling universal models and theories across disciplines
+
+\medskip
+
+$\rightarrow$ Building a theory of urban evolution should be related to an epistemological work on the nature of interdisciplinarity, the transfer of concepts.
+
+\medskip
+
+$\rightarrow$ Link with reflexivity and complexity itself \cite{raimbault2020relating}
+
+}
+
+\sframe{Morphogenesis, form and function}{
+
 % Ambitious new way of revisiting link between form and function: // disc morphogenesis
+
+A core element of the proposed model is the link between form and function (formeme including form and usages)
+
+\medskip
+
+$\rightarrow$ Revisit the concept of urban morphogenesis? \cite{batty2009centenary} Endogenize this link within the model? More general concept than biological morphogenesis?
+
+}
+
+\sframe{Autopoiesis: system boundaries?}{
 
@@
 
-\sframe{Discussion}{
+Question of system definition and system boundaries is not very clear in the proposed model (beyond taking into account geographical areas $c$)
+
+\medskip
+
+System boundaries are crucial in evolution / co-evolution (evolution occurs in spatio-temporal domains with isolation): notion of niche
+
+\medskip
+
+$\rightarrow$ Link with the concept of autopoiesis? \textit{Network of processes self-sustaining itself} \cite{bourgine2004autopoiesis} Are urban systems autopoietic?
 
 }
+
+\sframe{Open-ended evolution}{
+
+The sets used to define formemes are fixed in the model description: does the model apply to very long time scales, across multiple time scales, spatial scales (minimal resolution in mentioned in II, what about maximal resolution?), ontological scales?
+
+\medskip
+
+$\rightarrow$ Capacity of the model to endogeneize the emergence of new elements, new innovations, to capture open-ended evolution? \cite{taylor2016open}
+
+}
+
+% multidimensionality? -> ok encoded in formeme?
 

### 2021-03-14 Docs/Communications/2021/ALife2021/paper.tex
extendedabstract_alife21

\documentclass[letterpaper]{article}

\usepackage{natbib,alifeconf}  %% The order is important

% *****************
%  Requirements:
% *****************
%
% - All pages sized consistently at 8.5 x 11 inches (US letter size).
% - PDF length <= 8 pages for full papers, <=2 pages for extended
%    abstracts (not including citations).
% - Abstract length <= 250 words.
% - No visible crop marks.
% - Images at no greater than 300 dpi, scaled at 100%.
% - Embedded open type fonts only.
% - All layers flattened.
% - No attachments.
% - All desired links active in the files.

% Note that the PDF file must not exceed 5 MB if it is to be indexed
% by Google Scholar. Additional information about Google Scholar
% can be found here:
% http://www.google.com/intl/en/scholar/inclusion.html.

% If your system does not generate letter format documents by default,
% you can use the following workflow:
% latex example
% bibtex example
% latex example ; latex example
% dvips -o example.ps -t letterSize example.dvi
% ps2pdf example.ps example.pdf

% For pdflatex users:
% The alifeconf style file loads the "graphicx" package, and
% this may lead some users of pdflatex to experience problems.
% These can be fixed by editing the alifeconf.sty file to specify:
% \usepackage[pdftex]{graphicx}
%   instead of
% \usepackage{graphicx}.
% The PDF output generated by pdflatex should match the required
% specifications and obviously the dvips and ps2pdf steps become
% unnecessary.

% Note:  Some laser printers have a serious problem printing TeX
% output. The use of ps type I fonts should avoid this problem.

\title{Morphogenesis, evolution and co-evolution of cities}
\author{Juste Raimbault$^{1,2,3}$\\
\mbox{}\\
$^1$Center for Advanced Spatial Analysis, University College London, London, UK\\
$^2$UPS CNRS 3611 ISC-PIF, Paris, France\\
$^3$UMR CNRS 8504 G{\'e}ographie-cit{\'e}s, Paris, France\\
juste.raimbault@polytechnique.edu} % email of corresponding author

% For several authors from the same institution use the same number to
% refer to one address.
%
% If the names do not fit well on one line use
%         Author 1, Author 2 ... \\ {\Large\bf Author n} ...\\ ...
%
% If the title and author information do not fit in the area
% allocated, place \setlength\titlebox{<new height>} after the
% \documentclass line where <new height> is 2.25in

\begin{document}
\maketitle

\begin{abstract}
% Abstract length should not exceed 250 words
  Urban systems have a complexity in their own but still have been described and modelled using analogies and models imported from other disciplines. In particular, biological metaphors have been widely used in urban planning and design. We argue that fruitful transfer of concepts between biology and urban science can be achieved within the Artificial Life interdisciplinary framework. We illustrate this idea by synthesising a recent stream of research focusing on urban morphogenesis, urban evolution and urban co-evolution. In each case, novel definitions specific to the urban case were developed, while keeping the most proximity to the original disciplines. In that context, we suggest that many more interdisciplinary bridges could potentially be constructed through such disciplinary transfers going beyond purely methodological ones.
\end{abstract}

% keywords
% Urban morphogenesis; Urban evolution; Co-evolution; Interdisciplinarity

\section{Introduction}

The emergence of cities is tightly linked to the complexification of societies, and can be understood as one further step in cultural evolution \citep{sjoberg1965origin}. To what extent this additional layer of complexity lead to urban systems that themselves evolve with proper rules and processes remains an open question. Properties such as scaling which can be exhibited across successive level of emergence, from biological to social and urban systems \citep{youn2018scaling}, suggest at least some analogies between these different scales. In that context, biological metaphors have been used to model, understand, and design cities \citep{batty2009centenary}. Several disciplines have developed theories and models to understand the dynamics of cities \citep{pumain2020conclusion}, among which Artificial Life approaches have shown some relevance.

\cite{gershenson2013living} suggests that the application of concepts imported from the study of living systems can bring solutions to urban issues, such as traffic congestion, logistics, or more generally sustainability. Methods such as fractals can be used to quantify urban form \citep{chen2010modeling}. Urban traffic can also be modelled using ALife methods \citep{yoshioka2017macroscopic}. \cite{raimbault2020cities} shows a citation map of the scientific landscape loosely related to ALife in the study of urban systems, ranging from cellular automata models of urban growth to evolutionary computing applied to urban design or optimisation.

Beyond using similar methodologies or techniques, what is a common feature of the study of complex systems, the application of biological metaphors to the study of cities is one crucial feature in which ALife approaches have contributed to urban science, such as urban morphogenesis. The field of urban ecology, and more particularly industrial symbiosis \citep{chertow2007uncovering}, in which companies are understood as ecosystems, is one other example.

The purpose of this contribution is to illustrate that effective transfer of concepts, which go beyond metaphors with new definitions and an adaptation of concepts, can occur between biology and urban science, and be fruitful to understand the dynamics of urban systems. We furthermore claim that the ALife interdisciplinary framework is a privileged scientific background to operate such transfers. To do so, we propose to synthesise a recent stream of research bridging ALife and urban science, which originated with the study of co-evolutionary processes in urban systems \citep{raimbault2018caracterisation}. In that context, different concepts were transferred at different scales: urban morphogenesis at the scale of the city itself, and urban evolution and co-evolution at the scale of the system of cities.

\section{Urban morphogenesis}

Urban morphogenesis can be in it simplest sense understood as the growth of urban form, at the scale of a city or an urban area. \cite{raimbault2014hybrid} introduce for example a cellular automaton coupled with a dynamical road network, which produces a variety of dynamical regimes of urban growth \citep{raimbault2017identification}. \cite{raimbault2019generating} compare and calibrate generative models at the district scale. \cite{raimbault2018multi} show the complementarity of multiple heuristics to simulate the morphogenesis of transportation networks. The concept of urban morphogenesis can also be made closer to its biological counterpart by showing how reaction-diffusion models can accurately reproduce existing urban forms, as shown by \cite{raimbault2018calibration} for urban areas in Europe. Finally, by constructing a definition of urban morphogenesis as \emph{the strong coupling between the growth of urban form and the emergence of urban functions}, \cite{raimbault2018caracterisation} introduces an original viewpoint in which the transfer from biology is crucial. Following this positioning, the morphogenesis model studied by \cite{raimbault2019urban} uses transportation network to integrate a functional aspect into the dynamics of urban growth. Processes of urban morphogenesis can be seen as occurring within territorial niches \citep{holland2012signals}, what suggests the relevance of studying urban evolution within and between these niches at larger scales.

\section{Urban evolution}

The idea of evolving entities within urban systems directly evokes theories of cultural evolution \citep{mesoudi2016cultural}, as social and cultural processes are occurring within cities, which are furthermore highlighted as incubators of social change \citep{pumain2019complexity}. The concept of urban evolution however goes beyond this, as cities themselves can be seen as evolving entities. \cite{pumain1997pour} proposed an \emph{evolutionary theory of cities} to understand the dynamics of urban systems, in the sense of adaptive complex systems. \cite{pumain2017urban} synthesise the empirical and modeling results which were obtained in the direct heritage of this positioning.

Approaches of urban evolution with a definition closer to its biological and social counterparts however remains to be constructed. A partial proposition for a model of urban evolution is done by \cite{raimbault2020model}. Extending the model of \cite{favaro2011gibrat}, the main ingredients are innovations which propagate between cities. Therein, transmission processes are captured through diffusion, while transformation processes occur as mutations when new innovations are created. A one-dimensional urban genome is thus the adoption of each innovation in each city. While this model remains stylised and limited, first in the fact that other urban dimensions are not accounted for, secondly as innovations all compete on the same dimension, it however provides a definition and operationalisation of urban evolution which includes the formal processes required for evolution (transmission, transformation, isolation of sub-systems for differentiation) \citep{durham1991coevolution}.

\section{Co-evolution in urban systems}

Finally, one concept for which a relevant transfer can be achieved between biology and urban science is the concept of co-evolution. \cite{raimbault2018caracterisation} introduces a multi-level definition, with at the intermediate level the possibility of \emph{statistically co-evolving population of urban entities within territorial niches}. More concretely, it is applied to the co-evolution between transportation networks of cities, and hypothesises that a circular relation between some properties of cities and some of transportation networks can be observed within some geographical regions. A method to characterise it on spatio-temporal data was introduced by \cite{raimbault2017identification}. At the mesoscopic scale of urban areas, the morphogenesis model of \cite{raimbault2019generating} effectively captures such a co-evolution between indicators of urban morphology and the topology of road networks. \cite{le2015modeling} propose a model strongly coupling land-use dynamics with transportation infrastructure dynamics, integrating network governance agents. At the macroscopic scale of the system of cities, \cite{raimbault2018modeling} study a co-evolution model between cities and inter-city networks, and show that diverse regimes of co-evolution can be produced. \cite{raimbault2020hierarchy} extends this model with a more precise representation of transportation networks. These different applications show how this transfer of concept is fruitful to the understanding of urban dynamics.

\section{Discussion}

Several further transfers and developments can be suggested following this synthesis. A main research direction also strongly inspired by ALife and complexity is the construction of multi-scale models of urban evolution and urban dynamics. Previous efforts have remained limited as they did not used distinct ontologies and strong coupling between scales. For example, \cite{batty2005agents} introduces a common formalisation using cellular automata to simulate urban dynamics from the neighbourhood scale to the system of cities scale. \cite{murcio2015urban} study a model simulating urban migrations migrations at different spatial ranges. A first step towards such models has been recently proposed by \cite{raimbault2021strong}.

Such multi-scale models of urban dynamics would be essential tools for a sustainable territorial planning \citep{rozenblat2018conclusion}. Similarly, the benchmarking and comparison of multiple models for urban dynamics, as illustrated by \cite{raimbault2020empowering} for multiple urban systems worldwide, is a crucial step to ensure the robustness of policies.

Finally, many concepts originating from biology and studied across disciplines in the field of Artificial Life, remain relevant candidates for a strong transfer of concept and an operationalisation within urban science. For example, biomimicry has been put forward as an effective tool for urban design and architecture \citep{taylor2017art}. The concept of autopoiesis \citep{bourgine2004autopoiesis}, which is tightly linked with morphogenesis but also cognition, remains to be investigated in the case of urban systems. Urban computing and collective intelligence, as research fields such as smart cities and digital twins are booming \citep{batty2018digital}, may be linked to evolutionary computation, and more generally to studies of collective computation in biological systems. Altogether, we suggest that the integration of ALife into urban science and underlying transfer of concepts between disciplines, are relevant research directions for both.

%\section{Acknowledgements}
%

\footnotesize
\bibliographystyle{apalike}
\bibliography{biblio} % replace by the name of your .bib file

\end{document}

%\begin{figure}[t]
%\begin{center}
%\includegraphics{}
%\caption{}
%\label{fig1}
%\end{center}
%\end{figure}

%\begin{table}[h]
%\center{
%\begin{tabular}{|c|c|c|c|}\hline
%Name & Result & Bonus $b_i$ & Difficulty\\ \hline\hline
%Echo & I/O   & 1 & --\\
%Equals &$\neg(A\ {\rm xor}\ B)$&6& 4 \\ \hline
%\end{tabular}
%}
%\vskip 0.25cm
%\caption{}
%\end{table}

### 2021-10-21 Docs/Communications/2021/CCS2021/poster/poster.tex
poster_CCS

\documentclass{beamer}
  \usepackage[english]{babel}
  \usepackage[utf8]{inputenc}
  \usepackage{times}
  \usepackage{amsmath,amsthm, amssymb, latexsym,ragged2e}
  \boldmath
  \usepackage{multicol}

  \usetheme{Poster}
  \usepackage[orientation=portrait,size=a0,scale=1.4]{beamerposter}

% Urban systems have their own complexity and specific characteristics, but still have been described and modelled using analogies and models imported from other disciplines. In particular, biological metaphors have been widely used in urban planning and design. We argue that fruitful transfer of concepts between biology and urban science can be achieved within a complexity interdisciplinary framework. We illustrate this idea by synthesising a recent stream of research focusing on urban morphogenesis, urban evolution and urban co- evolution. First, urban morphogenesis can be understood as the growth of urban form, at the scale of a city or an urban area. [1] compares and calibrates morphogenesis models at the district scale. [2] shows for urban areas in Europe that reaction-diffusion models can accurately reproduce existing urban forms. Then, the concept of urban evolution can be related to cultural evolution in urban systems, but also to cities themselves as evolving entities. [3] proposed an evolutionary theory of cities to understand the dynamics of urban systems, in the sense of adaptive complex systems. In that context, a model of urban evolution is introduced by [4] in which the main ingredients are innovations propagating between cities. Finally, [5] introduces a multi-level definition for a concept of urban co- evolution, with at the intermediate level the possibility of statistically co-evolving population of urban entities within territorial niches. A method to characterise it on spatio-temporal data was introduced by [6]. At the mesoscopic scale of urban areas, the morphogenesis model of [7] effectively captures such a co-evolution between indicators of urban morphology and the topology of road networks. At the macroscopic scale of the system of cities, [8] study a co- evolution model between cities and inter-city networks, and show that diverse regimes of co- evolution can be produced. We suggest that other several interdisciplinary bridges could be relevant to urban science. For example, biomimicry has been put forward as an effective tool for urban design and architecture, while the concept of autopoiesis has not been applied yet to urban systems. Urban computing and collective intelligence, as research fields such as smart cities and digital twins are emerging, may also be linked to studies of collective computation in biological systems. Altogether, understanding urban systems as complex systems implies an interdisciplinary integration of concepts, methods and models.

  \title{Morphogenesis, Evolution and Co-evolution of Cities}
  \author[juste.raimbault@polytechnique.edu]{J. Raimbault$^{1,2,3}$}
  \institute[]
  {$^1$ CASA, UCL; $^2$ UPS CNRS 3611 ISC-PIF; $^3$ UMR CNRS 8504 Géographie-cités\vspace{1cm}}
  \date{}

  \logo{
  \hfill
  \includegraphics[height=8cm]{figures/ucl-logo-colours-notext.jpg}
  \includegraphics[height=8cm]{figures/casa_logo.jpeg}
  %\includegraphics[height=8cm]{figures/isc}
  %\includegraphics[height=8cm]{figures/geocite}
  \hfill\hfill
  }

  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
  \begin{document}
  \begin{frame}{}

    \vfill
    \begin{columns}[t]
      \begin{column}{.49\textwidth}

      \vspace{-1cm}

        \begin{block}{Context}
        \vspace{-1cm}
        \begin{columns}[t]
        \begin{column}{.95\textwidth}
          \begin{itemize}         
          \item \justify Urban complexity and interdisciplinarity \cite{pumain2005cumulativite}
          \end{itemize}
          \bigskip
          \begin{itemize} 
          \item \begin{justify} Urban science and Artificial Life: applications of biological concepts in planning and design \cite{batty2009centenary}
          \end{justify}

          \end{itemize}

          \bigskip
          \begin{justify}
          $\rightarrow$ \textit{Urban morphogenesis, urban evolution and urban co-evolution as concepts to understand urban dynamics and sustainability}
          \end{justify}

          \end{column}
          \end{columns}
        \end{block}

         \begin{block}{Urban morphogenesis}
         % Urban morphogenesis can be in it simplest sense understood as the growth of urban form, at the scale of a city or an urban area. Raimbault et al. (2014) introduce for example a cellular automaton coupled with a dynamical road net- work, which produces a variety of dynamical regimes of ur- ban growth (Raimbault, 2017). Raimbault and Perret (2019) compare and calibrate generative models at the district scale. Raimbault (2018d) show the complementarity of multiple heuristics to simulate the morphogenesis of transportation networks. The concept of urban morphogenesis can also be made closer to its biological counterpart by showing how reaction-diffusion models can accurately reproduce existing urban forms, as shown by Raimbault (2018a) for urban areas in Europe. Finally, by constructing a definition of urban morphogenesis as the strong coupling between the growth of urban form and the emergence of urban functions, Raim- bault (2018b) introduces an original viewpoint in which the transfer from biology is crucial. Following this positioning, the morphogenesis model studied by Raimbault (2019) uses transportation network to integrate a functional aspect into the dynamics of urban growth. Processes of urban morpho- genesis can be seen as occurring within territorial niches (Holland, 2012), what suggests the relevance of studying urban evolution within and between these niches at larger scales.

        %\vspace{-1cm}
        \begin{columns}[t]
        \begin{column}{.95\textwidth}
        \vspace{-2cm}

        %\begin{justify}
          \textbf{Definition of urban morphogenesis} \cite{raimbault2018caracterisation}: \textit{strong coupling between the dynamics of urban form and function}

          \bigskip

          \textbf{Generative models} for urban morphogenesis:

          \begin{itemize}
          		\item cellular automaton and road network \cite{raimbault2014hybrid}
          		\item building configurations \cite{raimbault2019generating}
          		\item reaction-diffusion model for population density \cite{raimbault2018calibration}
          \end{itemize}

		%\end{justify}

		\begin{center}

		\includegraphics[width=0.63\textwidth]{figures/spatialsens_calib.png}
		\hspace{0.5cm}
		\includegraphics[width=0.29\textwidth]{figures/mesobench_Fig3.png}

		\end{center}

		{\small\textit{Left: buildings \cite{raimbault2019generating}; Right: population \cite{raimbault2020comparison}}}

          \end{column}
          \end{columns}
        \end{block}

         \begin{block}{Urban evolution}
       \vspace{-2cm}
        \begin{columns}[t]
        \begin{column}{.95\textwidth}

        \vspace{0.5cm}

        \textit{Urban evolutionary theory introduced by \cite{pumain1997pour}}

  		\vspace{1cm}

        An \textbf{explicit urban evolution model} described by \cite{raimbault2020model}:
		\begin{itemize}
			\item Spatial diffusion of innovations captures transmission processes
			\item Spatial interaction models drive innovation and population growth
			\item Urban genome as the distribution of different innovations
			\item Mutation as the emergence of new innovations
		\end{itemize}

        \bigskip

        % trim=0cm 20cm 0cm 0cm trim does not work
        \begin{center}
        \includegraphics[width=\textwidth]{figures/urbanevol_Fig2.png}

        {\small\textit{Processes in the urban evolution model \cite{raimbault2020spatial}}}

        \end{center}

          \end{column}
          \end{columns}
        \end{block}

      \end{column}

      \begin{column}{.49\textwidth}

      \vspace{-1cm}

       \begin{block}{Co-evolution in urban systems}
        \vspace{-1.5cm}
          \begin{columns}[t]
        \begin{column}{.95\textwidth}

       %\begin{justify}

       \textbf{Definition of urban co-evolution} \cite{raimbault2018caracterisation}: \textit{statistically co-evolving population of urban entities within territorial niches}

       \vspace{1cm}

       \textbf{Characterisation method} based on circular Granger causalities between proxy variables

       \bigskip

       \begin{center}
       \includegraphics[width=0.87\textwidth]{figures/carac_regimes_1.png}

       {\small \textit{Diverse causality regimes in a morphogenesis model \cite{raimbault2017identification}}}
       \end{center}

       \vspace{1cm}

       \textbf{Application} to transportation networks and territories:

       \bigskip

       \begin{itemize}
       		\item At the \textbf{mesoscopic} scale (urban areas), co-evolution of urban morphology and road network topology \cite{raimbault2019urban}
       \end{itemize}
       \begin{center}
       \includegraphics[width=0.87\textwidth]{figures/meso_multimodeling.png}

       {\small \textit{Multiple network growth heuristics \cite{raimbault2018multi}}}

		\end{center}

        \vspace{1cm}

        \begin{itemize}
        	\item Land-use transport model and transportation governance at the \textbf{urban region} scale \cite{raimbault2021introducing}	
        \end{itemize}

        \vspace{1cm}

        \begin{itemize}
        	\item At the \textbf{macroscopic} scale (systems of cities), co-evolution between cities and inter-city transport networks \cite{raimbault2021modeling}
        \end{itemize}

       \vspace{1cm}

       \textbf{Possible extensions:} application to other urban dimensions (mobility and accessibility e.g.), integration with explicit urban evolution models

       %\end{justify}

                  \end{column}
          \end{columns}
        \end{block}

        \begin{block}{Discussion}
         \begin{columns}[t]
        \begin{column}{.95\textwidth}

		\vspace{-1.5cm}

		\begin{itemize}
			\item \justify Potential transfer of other concepts to urban systems: biomimicry, autopoiesis, collective computation
		\end{itemize}

		\begin{itemize}
			\item \justify Towards multi-scale models for urban systems

		\end{itemize}

		\begin{itemize}
			\item \justify Application to sustainable policies at multiple spatial and time scales \cite{rozenblat2018conclusion}
		\end{itemize}

          \end{column}
          \end{columns}
        \end{block}

      \end{column}
    \end{columns}

    %\begin{columns}[t]
    %  \begin{column}{\linewidth}

    \begin{block}{References}
        {\tiny
        \vspace{-1cm}
        \begin{multicols}{4}
          \bibliographystyle{apalike}
          \bibliography{biblio}
        \end{multicols}
          }
        \end{block}
    %\end{column}
    %\end{columns}

  \end{frame}
\end{document}

### 2021-11-29 Models/EvolutionInnovation/src/main/scala/urbanevolution/RunUrbanEvolution.scala
add_flows_as_indicator

@@ -31,5 +31,6 @@ object RunUrbanEvolution extends App {
   val finalHierarchy = Statistics.slope(result.populations.getCol(50).flatValues)._1
   println("Final pop hierarchy = "+finalHierarchy)
   println("Delta hierarchy pop = "+(finalHierarchy - Statistics.slope(result.populations.getCol(0).flatValues)._1))
+  println("Average gravity flow = "+result.averageGravityFlow)
 
 }

### 2021-11-29 Models/EvolutionInnovation/src/main/scala/urbanevolution/Statistics.scala

@@ -18,7 +18,7 @@ object Statistics {
 
 
   def slope(values: Array[Double]): (Double,Double) = {
-    def distribution: Array[Double] = values.filter(_ > 0).sorted(Ordering.Double.reverse)
+    def distribution: Array[Double] = values.filter(_ > 0).sorted(Ordering.Double.TotalOrdering.reverse)
     def distributionLog: Array[Array[Double]] = distribution.zipWithIndex.map { case (q, i) => Array(log(i + 1), log(q)) }
     val simpleRegression = new SimpleRegression(true)
     simpleRegression.addData(distributionLog)

### 2021-11-29 Models/EvolutionInnovation/src/main/scala/urbanevolution/UrbanEvolution.scala

@@ -41,9 +41,9 @@ case class UrbanEvolution(
     * run for model calibration
     * @return
     */
-  def run: (Double, Double, Double, Double) = {
+  def run: (Double, Double, Double, Double, Double) = {
     val result = UrbanEvolution.run(this)
-    (result.averageDiversity,result.averageUtility,result.averageInnovation,Statistics.slope(result.populations.getCol(dates.length - 1).flatValues)._1)
+    (result.averageDiversity,result.averageUtility,result.averageInnovation,Statistics.slope(result.populations.getCol(dates.length - 1).flatValues)._1, result.averageGravityFlow)
   }
 
 
@@ -78,11 +78,13 @@ object UrbanEvolution {
     * @param populations populations
     * @param innovationUtilities utilities
     * @param innovationShares shares
+    * @param gravityPotentials gravity potentials (could be recomputed from pops, distance weights and innov shares)
     */
   case class InnovationResult(
                                populations: Matrix,
                                innovationUtilities: Seq[Double],
-                               innovationShares: Seq[Matrix]
+                               innovationShares: Seq[Matrix],
+                               gravityPotentials: Seq[Matrix]
                              ) {
 
     /**
@@ -124,6 +126,14 @@ object UrbanEvolution {
       */
     def averageInnovation: Double = {
       innovationUtilities.length.toDouble/(populations.nrows.toDouble*populations.ncols.toDouble)
+    }
+
+    /**
+      * gravity flows averaged in time and over cities
+      * @return
+      */
+    def averageGravityFlow: Double = {
+      gravityPotentials.map(m => m.sum / (m.ncols*m.nrows)).sum / gravityPotentials.size
     }
 
 
@@ -379,6 +389,8 @@ object UrbanEvolution {
     archaicTechno.setMSubmat(0, 0, Array.fill(n)(Array(1.0)))
     innovationProportions.append(archaicTechno)
 
+    val gravityPotentials: ArrayBuffer[Matrix] = new ArrayBuffer[Matrix]
+
 
     for (t <- 1 until p) {
 
@@ -422,6 +434,8 @@ object UrbanEvolution {
       val potsgravity = diagpops %*% DenseMatrix.diagonal(technoFactor) %*% gravityDistanceWeights %*% diagpops
       val meanpotgravity = potsgravity.sum / (n * n)
 
+      gravityPotentials.append(potsgravity)
+
[…]

### 2021-11-29 Models/EvolutionInnovation/sdg.R
res_optim

@@ -1,0 +1,26 @@
+setwd(paste0(Sys.getenv('CS_HOME'),'/UrbanEvolution/Models/EvolutionInnovation'))
+
+library(ggplot2)
+library(dplyr)
+
+source(paste0(Sys.getenv('CS_HOME'),'/Organisation/Models/Utils/R/plots.R'))
+
+
+res <- as_tibble(read.csv('openmole/calibration/CALIBRATION_REPLICATIONS_LOCAL_20211129_115043/population10000.csv'))
+
+
+g <- ggplot(res,aes(x=oppAverageUtility, y = averageGravityFlow, color=gravityDecay, size = innovationDecay))
+g+geom_point(alpha=0.5)
+
+g <- ggplot(res,aes(x=oppAverageUtility, y = averageGravityFlow, color=newInnovationHierarchy, size = innovationDecay, shape=utilityDistribution))
+g+geom_point(alpha=0.5)
+
+
+## several Pareto fronts: initial hierarchy
+res1 <- as_tibble(read.csv('openmole/calibration/CALIBRATION_REPLICATIONS_HIERARCHY_0_5_LOCAL_20211129_165201/population3000.csv'))
+res2 <- as_tibble(read.csv('openmole/calibration/CALIBRATION_REPLICATIONS_HIERARCHY_1_5_LOCAL_20211129_165057/population3000.csv'))
+
+allres = rbind(cbind(res,hierarchy=rep("1",nrow(res))),cbind(res1,hierarchy=rep("0.5",nrow(res1))),cbind(res2,hierarchy=rep("1.5",nrow(res2))))
+
+g <- ggplot(allres,aes(x=oppAverageUtility, y = averageGravityFlow, color=gravityDecay, size = innovationDecay, shape=hierarchy))
+g+geom_point(alpha=0.5)

### 2021-11-30 Models/EvolutionInnovation/openmole/runCalib.sh
res

@@ -1,0 +1,4 @@
+openmole -p ../target/scala-2.13/urbanevolution_2.13-0.1-SNAPSHOT.jar --password-file ~/.oml --script Calibration_innovHierarchy0-5.oms
+openmole -p ../target/scala-2.13/urbanevolution_2.13-0.1-SNAPSHOT.jar --password-file ~/.oml --script Calibration_innovHierarchy1-5.oms
+openmole -p ../target/scala-2.13/urbanevolution_2.13-0.1-SNAPSHOT.jar --password-file ~/.oml --script Calibration_innovHierarchy1.oms
+

### 2021-11-30 Docs/Papers/SDGTradeoffs/arxiv/v1/main.tex
paper_arxiv

\documentclass{article}
\usepackage[utf8]{inputenc}
%\usepackage[margin=2cm]{geometry}
\usepackage{graphicx,hyperref}

\title{Trade-offs between sustainable development goals in systems of cities}

\author{J. Raimbault$^{1,\ast}$ and D. Pumain$^{2}$\medskip\\
$^{1}$ CASA, University College London\\
$^{2}$ G{\'e}ographie-cit{\'e}s, Universit{\'e} Paris 1\medskip\\
$\ast$ \texttt{juste.raimbault@polytechnnique.edu}
}
\date{}

\begin{document}

\maketitle

\begin{abstract}
	Sustainable Development Goals are intrinsically competing, but their embedding into urban systems furthermore emphasises such compromises, due to spatial complexity, the non-optimal nature of such systems, and multi-objective aspects of their agents, among other reasons. We propose in this paper to use a stylised simulation model for systems of cities, focused on innovation diffusion and population dynamics, to show how trade-offs may operate at such a scale. We proceed in particular to a bi-objective optimisation of emissions and innovation utilities, and show that no single urban optimum exists, but a diversity of regimes forming a compromise between the two objectives.
\end{abstract}

\section{Introduction}

The concept of urban optima, in the sense of optimising certain dimensions of urban systems, has been considered from diverse perspectives. In most cases, there does not seem to be clear patterns, neither empirical nor theoretical, of possible simple optimisations of single objectives by urban systems. Some results in urban economics regarding an optimal city size requires to consider a city in a closed system, which is unreasonable from a realistic perspectives \cite{singell1974optimum}. Studies of an optimal urban population density are restricted to economic criteria of wage and productivity \cite{su2017density}. The sustainability of urban forms for CO2 emissions requires considering complex indicators of urban form \cite{le2012urban}. In terms of pollution, empirical results across different urban systems suggest no fixed relationship between city size and emission of pollutants \cite{han2016optimum}. Altogether, this converges with the idea of multiple agents optimising multiple dimensions at different scales \cite{pumain2008socio}.

Sustainable Development Goals (SDGs) are characterised in a similar way by compromises between different dimensions. Urban sustainability, in the sense of the urban aspect of environmental issues \cite{finco2001pathways}, has thus to be understood as trade-offs between multiple objectives \cite{viguie2012trade}. This aspect occurs within subsystems themselves, such as in the case of designing transport networks \cite{sharma2011multiobjective}. Planning and policies must in that context account for such competing objectives \cite{caparros2015optimised}.

We propose in this paper to study trade-offs between different SDGs in systems of cities. We consider systems of cities at the macroscopic scale, and more particularly the dynamics of innovation diffusion and population growth. Using a stylised model for such urban dynamics, we apply a bi-objective optimisation genetic algorithm, to explore how trade-offs can occur in such systems. The rest of this paper is organised as follows: we first recall the assumptions of the system of cities model applied; we then describe results of its optimisation on proxies for two SDGs (innovation utility and emissions); we finally discuss theoretical implications of these results and how further work could include empirical components.

\section{Urban system model}

We work with a stylised model for the dynamics of urban systems at the macroscopic scale. This model is based on innovation diffusion dynamics and their impact on population growth. It was first formulated by \cite{favaro2011gibrat}, within the context of an evolutionary urban theory \cite{pumain1997pour}. A similar agent-based model was used to explore assumptions on the emergence of systems of cities themselves \cite{schmitt2015half}. A modified version was described by \cite{raimbault2020model} as an urban evolution model, including an urban genome shared and mutated across cities. As this particular version can furthermore be setup on stylised systems of cities, we use it in our multi-objective optimisation approach.

We describe the model below, but without detailed equations for the purpose of staying concise. See \cite{raimbault2020model} for the full description.

\subsection{Model setup}

The simulated urban system is composed by cities, characterised at each time step by (i) their population; (ii) their location in the geographical space; (iii) adoption rates by their populations for different innovations (urban genome).

We work on synthetic systems of cities, which are randomly generated given some fixed macro characteristics. This approach allows controlling for example for the role of space, and unentangling intrinsic model dynamics from geographical contingencies \cite{raimbault2019space}. In our case, as emission indicator is linked to inter-city flows, strongly dependent on the geography, averaging over several synthetic systems of cities will thus provide robust results.

Synthetic systems of cities are generated with random locations, an initial rank-size hierarchy which can be tuned (otherwise fixed to a default value of 1, to mirror a Zipf law distribution for city size), and a number of 30 cities. The largest city has initially a population of 100,000 and the model is run for 50 time steps.

\subsection{Model dynamics}

Starting from the initial state, the model updates population and innovation step by step. At each time step (of an order of magnitude of 10 years - the effects are observed on long time scales), the following procedure is used:

\begin{enumerate}
	\item innovations are diffused between cities using a spatial interaction model; innovations with a higher utility will diffuse more;
	\item populations are updated following an other spatial interaction model, with an advantage for cities being more innovative;
	\item new innovations may be invented in cities, following a probability determined by a mutation rate and by population (with a given hierarchy across cities);
	\item if a new innovation emerges, it has an initial penetration share fixed by one parameter, and a utility randomly distributed (normal or log-normal law), with a fixed standard deviation and an average corresponding to the current empirical average of existing innovation utilities.
\end{enumerate}
% note: do not necessarily have an increase of utility: in avg should stay the same with normal, deviate to higher with log-normal?

\subsection{Model parameters}

The parameters left free for optimisation are (i) spatial interaction range for innovation diffusion; (ii) spatial interaction range for population growth; (iii) mutation (innovation) rate; (iv) innovative city selection hierarchy; (v) rate of early adopters; (vi) standard-deviation of new innovation utilities; (vii) type of distribution for new innovation utilities.

\subsection{Optimisation objectives}

We consider the ``innovation'' SDG (goal 9) and the ``climate'' SDG (goal 14) as conflicting objectives. We can expect that a higher economic activity linked to more intensive innovative activities will increase endogenous emissions, but also transport emissions between urban areas, generated by economic and transport flows. The existence of a trade-off remains an hypothesis which will be checked during the optimisation stage.

We consider therefore the two following objectives for model optimisation: (i) aggregated total utility during model dynamics, computed over time and across cities, with shares of each innovations; (ii) total emissions due to transport flows between cities, computed as cumulative population gravity flows.

\section{Results}

\subsection{Implementation}

The model is implemented in \texttt{scala} for performances purposes, using matrix operations to update innovation shares and populations. Source code and simulation results are available on the open git repository of the project at \url{https://github.com/JusteRaimbault/UrbanEvolution}.

Model optimisation is achieved by integrating the model into the OpenMOLE platform \cite{reuillon2013openmole}. This free and open source software facilitates model embedding into a workflow system, distribution of computation into high-performance computing infrastructures, and provides a simple access to state-of-the-art model validation methods.

\subsection{Bi-objective optimisation}

We investigate trade-offs between total innovation utility and emissions, by optimising the model using a bi-objective heuristic with free parameters and indicators detailed above. We use a NSGA2 optimisation algorithm, provided by OpenMOLE, with a population of 100 individuals, for 10,000 generations.

\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{pareto-oppAverageUtility-averageGravityFlow_color-gravityDecay_size-innovationDecay_shape-utilityDistribution.png}
	\caption{Pareto front between the opposite of average utility and cumulated gravity flows. Point color gives population spatial interaction range; point size innovation diffusion range; and point shape the utility distribution.\label{fig:fig1}}
\end{figure}

We show optimisation results, as the final algorithm population, in Fig.~\ref{fig:fig1}. We indeed find a broad Pareto front, confirming the existence of a trade-off in such urban dynamics driven by innovation diffusion. We note two parts of the Pareto front, with fat-tailed distributions for utility distribution (log-normal) giving the upper part of the front corresponding to situations with a higher utility but which are more emission intensive. Within this subfront, population spatial interaction are rather local, while a more local innovation diffusion yields less emitting configurations. A similar aspect is observed for the normal distribution subfront, with a U-shaped value of population spatial interactions when going through the front: a more integrated system in terms of population migration produces by itself an intermediate compromise.

\subsection{Conditional optimisation}

% other experiments: conditional to initial hierarchy? innovation hierarchy? -> both useful for policies -> compare Pareto fronts
% Q: normalise by total pop?

\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{pareto-oppAverageUtility-averageGravityFlow_VARYINGHIERARCHY_color-gravityDecay_size-innovationDecay.png}
	\caption{Pareto fronts, with initial population hierarchy fixed at different values (point shape).\label{fig:fig2}}
\end{figure}

We now turn to experiments which could potentially provide policy insights. We run the same optimisation as before, but changing the initial population hierarchy of cities. To put it simply, we investigate how trade-offs change in different hypothetical systems of cities, ranging from highly hierarchical (Zipf exponent of 1.5) to a more balanced system (exponent of 0.5). We show results in Fig.~\ref{fig:fig2}. We find that the higher the hierarchy, the less flat the front. Overall, the less hierarchical system dominates the others (but this comparison remains limited as total population is different across systems). Furthermore, the size of the front is the smallest with the lowest hierarchy, meaning that this system is indeed closer to some global optimum.

\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{pareto-oppAverageUtility-averageGravityFlow_VARYINGINNOVHIERARCHY_color-gravityDecay_size-innovationDecay.png}
	\caption{Pareto fronts, with innovation hierarchy fixed at different values (point shape).\label{fig:fig3}}
\end{figure}

We finally show in Fig.~\ref{fig:fig3} a similar conditional optimisation, run by changing the fixed value of innovation hierarchy. This corresponds in terms of policies, to either letting innovation aggregate into larger metropolises (scaling with a high exponent value \cite{pumain2006evolutionary}), or regulating and providing incentives to enhance innovation into smaller and medium-sized cities. We also find that balanced policies provides a more optimal front (they can be compared in this case). Furthermore, this lowest hierarchy corresponds to much higher absolute values of total utility, going against the narrative of a higher value innovation produced by large cities only. Points for the two other fronts are rather close, corresponding to a lower sensitivity when the scaling exponent is larger than one.

\section{Discussion}

We have shown, in a stylised model of urban population and innovation dynamics, that trade-offs between transport emissions and total innovation utility emerge from model dynamics. This has theoretical implications, confirming the general non-optimising nature of urban systems and the predominance of trade-offs across different urban dimensions. Our results from conditional optimisation suggest that less hierarchical systems, both regarding initial population hierarchy and innovation hierarchy, provide more optimal Pareto fronts. This could have implications for policies such as innovation incentives, to avoid a too strong concentration into larger cities. More empirical investigations remain however needed.

Extending this theoretical and stylised work towards more empirical and data-grounded applications raises several issues. First, how to quantify spatial proxies for innovation, to either parametrise the initial configuration, or to calibrate model trajectories in terms of innovation diffusion, remains a difficult question. The use of patent data provides such insights \cite{griliches200713}, but the lack of harmonised and spatialised open patent database limits this perspective. Some initiatives are currently working towards this goal, such as \cite{bergeaud2021patentcity}. Furthermore, more realistic indicators for emissions, both transport and endogenous ones, would be also needed, for example by estimating them through a link with existing emissions databases such as EDGAR \cite{olivier1994emission}.

To conclude, we have provided a first stylised insight into trade-offs between SDGs in systems of cities at the macroscopic scale, which can be applied from a theoretical viewpoint to validate or unvalidate urban theories, and be used as a basis towards more practical application towards sustainable long-term territorial policies \cite{rozenblat2018conclusion}.

\newpage

\begin{thebibliography}{}

\bibitem[Bergeaud and Verluise, 2021]{bergeaud2021patentcity}
Bergeaud, A. and Verluise, C. (2021).
\newblock Patentcity: A century of innovation: New data and facts.

\bibitem[Caparros-Midwood et~al., 2015]{caparros2015optimised}
Caparros-Midwood, D., Barr, S., and Dawson, R. (2015).
\newblock Optimised spatial planning to meet long term urban sustainability
  objectives.
\newblock {\em Computers, Environment and Urban Systems}, 54:154--164.

\bibitem[Favaro and Pumain, 2011]{favaro2011gibrat}
Favaro, J.-M. and Pumain, D. (2011).
\newblock Gibrat revisited: An urban growth model incorporating spatial
  interaction and innovation cycles.
\newblock {\em Geographical Analysis}, 43(3):261--286.

\bibitem[Finco and Nijkamp, 2001]{finco2001pathways}
Finco, A. and Nijkamp, P. (2001).
\newblock Pathways to urban sustainability.
\newblock {\em Journal of Environmental Policy \& Planning}, 3(4):289--302.

\bibitem[Griliches, 2007]{griliches200713}
Griliches, Z. (2007).
\newblock {\em 13. Patent Statistics as Economic Indicators: A Survey}.
\newblock University of Chicago Press.

\bibitem[Han et~al., 2016]{han2016optimum}
Han, L., Zhou, W., Pickett, S.~T., Li, W., and Li, L. (2016).
\newblock An optimum city size? the scaling relationship for urban population
  and fine particulate (pm2. 5) concentration.
\newblock {\em Environmental Pollution}, 208:96--101.

\bibitem[Le~N{\'e}chet, 2012]{le2012urban}
Le~N{\'e}chet, F. (2012).
\newblock Urban spatial structure, daily mobility and energy consumption: a
  study of 34 european cities.
\newblock {\em Cybergeo: European Journal of Geography}.

\bibitem[Olivier et~al., 1994]{olivier1994emission}
Olivier, J., Bouwman, A., Van~der Maas, C., and Berdowski, J. (1994).
\newblock Emission database for global atmospheric research (edgar).
\newblock {\em Environmental Monitoring and Assessment}, 31(1):93--106.

\bibitem[Pumain, 1997]{pumain1997pour}
Pumain, D. (1997).
\newblock Pour une th{\'e}orie {\'e}volutive des villes.
\newblock {\em L'Espace g{\'e}ographique}, pages 119--134.

\bibitem[Pumain, 2008]{pumain2008socio}
Pumain, D. (2008).
\newblock The socio-spatial dynamics of systems of cities and innovation
  processes: a multi-level model.
\newblock In {\em The Dynamics of Complex Urban Systems}, pages 373--389.
  Springer.

\bibitem[Pumain et~al., 2006]{pumain2006evolutionary}
Pumain, D., Paulus, F., Vacchiani-Marcuzzo, C., and Lobo, J. (2006).
\newblock An evolutionary theory for interpreting urban scaling laws.
\newblock {\em Cybergeo: European Journal of Geography}.

\bibitem[Raimbault, 2020]{raimbault2020model}
Raimbault, J. (2020).
\newblock A model of urban evolution based on innovation diffusion.
\newblock In {\em Artificial Life Conference Proceedings}, pages 500--508. MIT
  Press.

\bibitem[Raimbault et~al., 2019]{raimbault2019space}
Raimbault, J., Cottineau, C., Le~Texier, M., Le~Nechet, F., and Reuillon, R.
  (2019).
\newblock Space matters: Extending sensitivity analysis to initial spatial
  conditions in geosimulation models.
\newblock {\em Journal of Artificial Societies and Social Simulation}, 22(4).

\bibitem[Reuillon et~al., 2013]{reuillon2013openmole}
Reuillon, R., Leclaire, M., and Rey-Coyrehourcq, S. (2013).
\newblock Openmole, a workflow engine specifically tailored for the distributed
  exploration of simulation models.
\newblock {\em Future Generation Computer Systems}, 29(8):1981--1990.

\bibitem[Rozenblat and Pumain, 2018]{rozenblat2018conclusion}
Rozenblat, C. and Pumain, D. (2018).
\newblock Conclusion: Toward a methodology for multi-scalar urban system
  policies.
\newblock {\em International and Transnational Perspectives on Urban Systems},
  page 385.

\bibitem[Schmitt et~al., 2015]{schmitt2015half}
Schmitt, C., Rey-Coyrehourcq, S., Reuillon, R., and Pumain, D. (2015).
\newblock Half a billion simulations: Evolutionary algorithms and distributed
  computing for calibrating the simpoplocal geographical model.
\newblock {\em Environment and Planning B: Planning and Design},
  42(2):300--315.

\bibitem[Sharma and Mathew, 2011]{sharma2011multiobjective}
Sharma, S. and Mathew, T.~V. (2011).
\newblock Multiobjective network design for emission and travel-time trade-off
  for a sustainable large urban transportation network.
\newblock {\em Environment and Planning B: Planning and Design},
  38(3):520--538.

\bibitem[Singell, 1974]{singell1974optimum}
Singell, L.~D. (1974).
\newblock Optimum city size: Some thoughts on theory and policy.
\newblock {\em Land Economics}, 50(3):207--212.

\bibitem[Su et~al., 2017]{su2017density}
Su, H., Wei, H., and Zhao, J. (2017).
\newblock Density effect and optimum density of the urban population in china.
\newblock {\em Urban Studies}, 54(7):1760--1777.

\bibitem[Vigui{\'e} and Hallegatte, 2012]{viguie2012trade}
Vigui{\'e}, V. and Hallegatte, S. (2012).
\newblock Trade-offs and synergies in urban climate policies.
\newblock {\em Nature Climate Change}, 2(5):334--337.

\end{thebibliography}

\end{document}

### 2021-11-30 Models/EvolutionInnovation/sdg.R

@@ -5,22 +5,41 @@ library(dplyr)
 
 source(paste0(Sys.getenv('CS_HOME'),'/Organisation/Models/Utils/R/plots.R'))
 
+resprefix = 'CALIBRATION_REPLICATIONS_LOCAL_20211129_115043'
+res <- as_tibble(read.csv(paste0('openmole/calibration/',resprefix,'/population10000.csv')))
+resdir = paste0(Sys.getenv('CS_HOME'),'/UrbanEvolution/Results/EvolutionInnovation/',resprefix,'/');dir.create(resdir)
 
-res <- as_tibble(read.csv('openmole/calibration/CALIBRATION_REPLICATIONS_LOCAL_20211129_115043/population10000.csv'))
-
-
-g <- ggplot(res,aes(x=oppAverageUtility, y = averageGravityFlow, color=gravityDecay, size = innovationDecay))
+g <- ggplot(res,aes(x=oppAverageUtility, y = averageGravityFlow, color=gravityDecay, size = innovationDecay, shape=utilityDistribution))
 g+geom_point(alpha=0.5)
+ggsave(file=paste0(resdir,'pareto-oppAverageUtility-averageGravityFlow_color-gravityDecay_size-innovationDecay_shape-utilityDistribution.png'),width=22,height=18,units='cm')
 
-g <- ggplot(res,aes(x=oppAverageUtility, y = averageGravityFlow, color=newInnovationHierarchy, size = innovationDecay, shape=utilityDistribution))
-g+geom_point(alpha=0.5)
+
+#g <- ggplot(res,aes(x=oppAverageUtility, y = averageGravityFlow, color=newInnovationHierarchy, size = innovationDecay, shape=utilityDistribution))
+#g+geom_point(alpha=0.5)
 
 
 ## several Pareto fronts: initial hierarchy
-res1 <- as_tibble(read.csv('openmole/calibration/CALIBRATION_REPLICATIONS_HIERARCHY_0_5_LOCAL_20211129_165201/population3000.csv'))
-res2 <- as_tibble(read.csv('openmole/calibration/CALIBRATION_REPLICATIONS_HIERARCHY_1_5_LOCAL_20211129_165057/population3000.csv'))
+res1 <- as_tibble(read.csv('openmole/calibration/CALIBRATION_REPLICATIONS_HIERARCHY_0_5_LOCAL_20211129_165201/population10000.csv'))
[…]

### 2022-08-30 Docs/Papers/ALifeCities/arxiv/v1/paper.tex
reviews_alife2020_alifecities

--- 2020-04-22 Docs/Papers/ALifeCities/paper/paper.tex
@@
 % \cite{pattee1995artificial} epistemic cut: simulation / hard or wet ALife / nature of computation % not relevant here?
-
-\cite{doi:10.1098/rstb.2019.0329}
 
@@
 \begin{mdframed}
-\includegraphics[width=\linewidth]{figures/core.png}
+\includegraphics[width=\linewidth]{core.png}
 \end{mdframed}
 \begin{mdframed}
-\includegraphics[width=\linewidth]{figures/corealife_zoom.png}
+\includegraphics[width=\linewidth]{corealife_zoom.png}
 \end{mdframed}
@@
 \begin{center}
-\includegraphics[width=0.49\linewidth]{figures/citation_all.png}
-\includegraphics[width=0.49\linewidth]{figures/citation_alife.png}
+\includegraphics[width=0.49\linewidth]{citation_all.png}
+\includegraphics[width=0.49\linewidth]{citation_alife.png}
 \caption{Citation flows between largest communities, for the full network (Left) and the ALife subnetwork (Right).
@@
 \footnotesize
-\bibliographystyle{apalike}
-\bibliography{biblio} % replace by the name of your .bib file
+%\bibliographystyle{apalike}
+%\bibliography{biblio} % replace by the name of your .bib file
+
+\begin{thebibliography}{}
+
+\bibitem[Adamowski and Karapataki, 2010]{adamowski2010comparison}
+Adamowski, J. and Karapataki, C. (2010).
+\newblock Comparison of multivariate regression and artificial neural networks
+  for peak urban water-demand forecasting: evaluation of different ann learning
+  algorithms.
+\newblock {\em Journal of Hydrologic Engineering}, 15(10):729--743.
+
+\bibitem[Aguilar et~al., 2014]{aguilar2014past}
+Aguilar, W., Santamar{\'\i}a-Bonfil, G., Froese, T., and Gershenson, C. (2014).
+\newblock The past, present, and future of artificial life.
+\newblock {\em Frontiers in Robotics and AI}, 1:8.
+
+\bibitem[Akbarzadeh et~al., 2006]{akbarzadeh2006towards}
+Akbarzadeh, A., Frahm, J.-M., Mordohai, P., Clipp, B., Engels, C., Gallup, D.,
+  Merrell, P., Phelps, M., Sinha, S., Talton, B., et~al. (2006).
+\newblock Towards urban 3d reconstruction from video.
+\newblock In {\em Third International Symposium on 3D Data Processing,
+  Visualization, and Transmission (3DPVT'06)}, pages 1--8. IEEE.
+
+\bibitem[Banzhaf et~al., 2016]{banzhaf2016defining}
+Banzhaf, W., Baumgaertner, B., Beslon, G., Doursat, R., Foster, J.~A.,
+  McMullin, B., De~Melo, V.~V., Miconi, T., Spector, L., Stepney, S., et~al.
+  (2016).
+\newblock Defining and simulating open-ended novelty: requirements, guidelines,
+  and challenges.
+\newblock {\em Theory in Biosciences}, 135(3):131--161.
+
+\bibitem[Batty, 2018]{batty2018inventing}
+Batty, M. (2018).
+\newblock {\em Inventing future cities}.
+\newblock Mit Press.
+
+\bibitem[Batty and Marshall, 2009]{batty2009centenary}
+Batty, M. and Marshall, S. (2009).
+\newblock Centenary paper: The evolution of cities: Geddes, abercrombie and the
+  new physicalism.
+\newblock {\em Town Planning Review}, 80(6):551--574.
+
+\bibitem[Batty and Xie, 1994]{batty1994cells}
+Batty, M. and Xie, Y. (1994).
+\newblock From cells to cities.
+\newblock {\em Environment and planning B: Planning and design},
+  21(7):S31--S48.
+
+\bibitem[Blondel et~al., 2008]{blondel2008fast}
+Blondel, V.~D., Guillaume, J.-L., Lambiotte, R., and Lefebvre, E. (2008).
+\newblock Fast unfolding of communities in large networks.
+\newblock {\em Journal of statistical mechanics: theory and experiment},
+  2008(10):P10008.
+
+\bibitem[Bourgine and Stewart, 2004]{bourgine2004autopoiesis}
+Bourgine, P. and Stewart, J. (2004).
+\newblock Autopoiesis and cognition.
+\newblock {\em Artificial life}, 10(3):327--345.
+
+\bibitem[Bull et~al., 2000]{bull2000meme}
+Bull, L., Holland, O., and Blackmore, S. (2000).
+\newblock On meme--gene coevolution.
+\newblock {\em Artificial life}, 6(3):227--235.
+
+\bibitem[Casti, 1997]{casti1997would}
+Casti, J.~L. (1997).
+\newblock Would-be worlds: toward a theory of complex systems.
+\newblock {\em Artificial Life and Robotics}, 1(1):11--13.
+
+\bibitem[Chen and Cheng, 2010]{chen2010review}
+Chen, B. and Cheng, H.~H. (2010).
+\newblock A review of the applications of agent technology in traffic and
+  transportation systems.
+\newblock {\em IEEE Transactions on intelligent transportation systems},
+  11(2):485--497.
+
+\bibitem[Cottineau et~al., 2017]{cottineau2017diverse}
+Cottineau, C., Hatna, E., Arcaute, E., and Batty, M. (2017).
+\newblock Diverse cities or the systematic paradox of urban scaling laws.
+\newblock {\em Computers, environment and urban systems}, 63:80--94.
+
+\bibitem[Davids, 2002]{davids2002urban}
+Davids, A. (2002).
+\newblock Urban search and rescue robots: from tragedy to technology.
+\newblock {\em IEEE Intelligent systems}, 17(2):81--83.
+
+\bibitem[Doursat et~al., 2012]{doursat2012morphogenetic}
+Doursat, R., Sayama, H., and Michel, O. (2012).
+\newblock {\em Morphogenetic engineering: toward programmable complex systems}.
+\newblock Springer.
+
+\bibitem[D’Acci, 2013]{d2013simulating}
+D’Acci, L. (2013).
+\newblock Simulating future societies in isobenefit cities: Social isobenefit
+  scenarios.
+\newblock {\em Futures}, 54:3--18.
+
+\bibitem[Flack and Krakauer, 2011]{flack2011challenges}
+Flack, J.~C. and Krakauer, D.~C. (2011).
+\newblock Challenges for complexity measures: A perspective from social
+  dynamics and collective social computation.
+\newblock {\em Chaos: An Interdisciplinary Journal of Nonlinear Science},
+  21(3):037108.
+
+\bibitem[Franklin, 2002]{franklin2002embodied}
+Franklin, S. (2002).
+\newblock {\em Embodied progress: A cultural account of assisted conception}.
+\newblock Routledge.
+
+\bibitem[Gonz{\'a}lez et~al., 2015]{gonzalez2015review}
+Gonz{\'a}lez, D., P{\'e}rez, J., Milan{\'e}s, V., and Nashashibi, F. (2015).
+\newblock A review of motion planning techniques for automated vehicles.
+\newblock {\em IEEE Transactions on Intelligent Transportation Systems},
+  17(4):1135--1145.
+
+\bibitem[Ib{\'a}{\~n}ez et~al., 2011]{ibanez2011using}
+Ib{\'a}{\~n}ez, A., Larra{\~n}aga, P., and Bielza, C. (2011).
+\newblock Using bayesian networks to discover relationships between
+  bibliometric indices. a case study of computer science and artificial
+  intelligence journals.
+\newblock {\em Scientometrics}, 89(2):523--551.
+
+\bibitem[IEPSFC, 2018]{iepsfc2018science}
+IEPSFC (2018).
+\newblock Science and the future of cities, report of the international expert
+  panel on science and the future of cities.
+\newblock Technical report.
+
+\bibitem[Kato et~al., 1998]{kato1998alife}
+Kato, N., Okuno, T., Okano, A., Kanoh, H., and Nishihara, S. (1998).
+\newblock An alife approach to modeling virtual cities.
+\newblock In {\em SMC'98 Conference Proceedings. 1998 IEEE International
+  Conference on Systems, Man, and Cybernetics (Cat. No. 98CH36218)}, volume~2,
+  pages 1168--1173. IEEE.
+
+\bibitem[Kato et~al., 2000]{kato2000modeling}
+Kato, N., Okuno, T., Suzuki, R., and Kanoh, H. (2000).
+\newblock Modeling virtual cities based on interaction between cells.
+\newblock In {\em Smc 2000 conference proceedings. 2000 ieee international
+  conference on systems, man and cybernetics.'cybernetics evolving to systems,
+  humans, organizations, and their complex interactions'(cat. no. 0}, volume~1,
+  pages 143--148. IEEE.
+
+\bibitem[Kendoul, 2012]{kendoul2012survey}
+Kendoul, F. (2012).
+\newblock Survey of advances in guidance, navigation, and control of unmanned
+  rotorcraft systems.
+\newblock {\em Journal of Field Robotics}, 29(2):315--378.
+
+\bibitem[Kim and Cho, 2006]{kim2006comprehensive}
+Kim, K.-J. and Cho, S.-B. (2006).
+\newblock A comprehensive overview of the applications of artificial life.
+\newblock {\em Artificial Life}, 12(1):153--182.
+
+\bibitem[Langton, 1986]{langton1986studying}
+Langton, C.~G. (1986).
+\newblock Studying artificial life with cellular automata.
+
+\bibitem[Levinson et~al., 2011]{levinson2011towards}
+Levinson, J., Askeland, J., Becker, J., Dolson, J., Held, D., Kammel, S.,
+  Kolter, J.~Z., Langer, D., Pink, O., Pratt, V., et~al. (2011).
+\newblock Towards fully autonomous driving: Systems and algorithms.
+\newblock In {\em 2011 IEEE Intelligent Vehicles Symposium (IV)}, pages
+  163--168. IEEE.
+
+\bibitem[Leydesdorff, 2001]{leydesdorff2001challenge}
+Leydesdorff, L. (2001).
+\newblock {\em The challenge of scientometrics: The development, measurement,
+  and self-organization of scientific communications}.
+\newblock Universal-Publishers.
+
+\bibitem[Lobo et~al., 2020]{lobo2020urban}
+Lobo, J., Alberti, M., Allen-Dumas, M., Arcaute, E., Barthelemy, M.,
+  Bojorquez~Tapia, L.~A., Brail, S., Bettencourt, L., Beukes, A., Chen, W.-Q.,
+  et~al. (2020).
+\newblock Urban science: Integrated theory from the first cities to sustainable
+  metropolises.
+
+\bibitem[Lucic and Teodorovic, 2002]{lucic2002transportation}
+Lucic, P. and Teodorovic, D. (2002).
+\newblock Transportation modeling: an artificial life approach.
+\newblock In {\em 14th IEEE International Conference on Tools with Artificial
+  Intelligence, 2002.(ICTAI 2002). Proceedings.}, pages 216--223. IEEE.
+
+\bibitem[Lv et~al., 2014]{lv2014traffic}
+Lv, Y., Duan, Y., Kang, W., Li, Z., and Wang, F.-Y. (2014).
+\newblock Traffic flow prediction with big data: a deep learning approach.
+\newblock {\em IEEE Transactions on Intelligent Transportation Systems},
+  16(2):865--873.
+
+\bibitem[Ma et~al., 2009]{ma2009real}
+Ma, Y., Chowdhury, M., Sadek, A., and Jeihani, M. (2009).
+\newblock Real-time highway traffic condition assessment framework using
+  vehicle--infrastructure integration (vii) with artificial intelligence (ai).
+\newblock {\em IEEE Transactions on Intelligent Transportation Systems},
+  10(4):615--627.
+
+\bibitem[Medda et~al., 2009]{medda2009morphogenetic}
+Medda, F., Nijkamp, P., and Rietveld, P. (2009).
+\newblock A morphogenetic perspective on spatial complexity.
+\newblock In {\em Complexity and spatial networks}, pages 51--60. Springer.
+
+\bibitem[Nicosia et~al., 2009]{nicosia2009extending}
+Nicosia, V., Mangioni, G., Carchiolo, V., and Malgeri, M. (2009).
+\newblock Extending the definition of modularity to directed graphs with
+  overlapping communities.
+\newblock {\em Journal of Statistical Mechanics: Theory and Experiment},
+  2009(03):P03024.
+
+\bibitem[Niu et~al., 2016]{niu2016global}
+Niu, J., Tang, W., Xu, F., Zhou, X., and Song, Y. (2016).
+\newblock Global research on artificial intelligence from 1990--2014:
+  Spatially-explicit bibliometric analysis.
+\newblock {\em ISPRS International Journal of Geo-Information}, 5(5):66.
+
+\bibitem[Olsen, 1982]{olsen1982urban}
+Olsen, S. (1982).
+\newblock Urban metabolism and morphogenesis.
+\newblock {\em Urban Geography}, 3(2):87--109.
+
+\bibitem[Openshaw, 1995]{openshaw1995developing}
+Openshaw, S. (1995).
+\newblock Developing automated and smart spatial pattern exploration tools for
+  geographical information systems applications.
+\newblock {\em Journal of the Royal Statistical Society: Series D (The
+  Statistician)}, 44(1):3--16.
+
+\bibitem[Pelechano et~al., 2007]{pelechano2007controlling}
+Pelechano, N., Allbeck, J.~M., and Badler, N.~I. (2007).
+\newblock Controlling individual agents in high-density crowd simulation.
+
+\bibitem[Pumain, 2018]{pumain2018evolutionary}
+Pumain, D. (2018).
+\newblock An evolutionary theory of urban systems.
+\newblock In {\em International and transnational perspectives on urban
+  systems}, pages 3--18. Springer.
+
+\bibitem[Raimbault, 2018]{raimbault2018calibration}
+Raimbault, J. (2018).
+\newblock Calibration of a density-based model of urban morphogenesis.
+\newblock {\em PloS one}, 13(9).
+
+\bibitem[Raimbault, 2019a]{raimbault2019exploration}
+Raimbault, J. (2019a).
+\newblock Exploration of an interdisciplinary scientific landscape.
+\newblock {\em Scientometrics}, 119(2):617--641.
+
+\bibitem[Raimbault, 2019b]{raimbault2019modeling}
+Raimbault, J. (2019b).
+\newblock Modeling interactions between transportation networks and
+  territories: a co-evolution approach.
+\newblock {\em arXiv preprint arXiv:1902.04802}.
+
+\bibitem[Raimbault et~al., 2020]{raimbault2020spatial}
+Raimbault, J., Broere, J., Somveille, M., Serna, J.~M., Strombom, E., Moore,
+  C., Zhu, B., and Sugar, L. (2020).
+\newblock A spatial agent based model for simulating and optimizing networked
+  eco-industrial systems.
+\newblock {\em Resources, Conservation and Recycling}, 155:104538.
+
+\bibitem[Raimbault et~al., 2019]{raimbault2019empowering}
+Raimbault, J., Chasset, P.-O., Cottineau, C., Commenges, H., Pumain, D.,
+  Kosmopoulos, C., and Banos, A. (2019).
+\newblock Empowering open science with reflexive and spatialised indicators.
+\newblock {\em Environment and Planning B: Urban Analytics and City Science},
+  page 2399808319870816.
+
+\bibitem[Raimbault and Perret, 2019]{raimbault2019generating}
+Raimbault, J. and Perret, J. (2019).
+\newblock Generating urban morphologies at large scales.
+\newblock In {\em The 2019 Conference on Artificial Life}, pages 179--186. MIT
+  Press.
+
+\bibitem[Rigas et~al., 2014]{rigas2014managing}
+Rigas, E.~S., Ramchurn, S.~D., and Bassiliades, N. (2014).
+\newblock Managing electric vehicles in the smart grid using artificial
+  intelligence: A survey.
+\newblock {\em IEEE Transactions on Intelligent Transportation Systems},
+  16(4):1619--1635.
+
+\bibitem[Sayama, 2019]{doi:10.1162/isala00134}
+Sayama, H. (2019).
+\newblock Suppleness and open-endedness for social sustainability.
+\newblock {\em Artificial Life Conference Proceedings}, (31):28--29.
+
+\bibitem[Schipperijn et~al., 2010]{schipperijn2010factors}
+Schipperijn, J., Ekholm, O., Stigsdotter, U.~K., Toftager, M., Bentsen, P.,
+  Kamper-J{\o}rgensen, F., and Randrup, T.~B. (2010).
+\newblock Factors influencing the use of green space: Results from a danish
+  national representative survey.
+\newblock {\em Landscape and urban planning}, 95(3):130--137.
+
+\bibitem[Sen-Ching and Kamath, 2004]{sen2004robust}
+Sen-Ching, S.~C. and Kamath, C. (2004).
+\newblock Robust techniques for background subtraction in urban traffic video.
+\newblock In {\em Visual Communications and Image Processing 2004}, volume
+  5308, pages 881--892. International Society for Optics and Photonics.
+
+\bibitem[Siqueiros-García, 2019]{doi:10.1162/isala00135}
+Siqueiros-García, J.~M. (2019).
+\newblock Cities, a conceptual framing for a synthetic perspective.
+\newblock {\em Artificial Life Conference Proceedings}, (31):30--31.
+
+\bibitem[Squazzoni and Casnici, 2013]{squazzoni2013social}
+Squazzoni, F. and Casnici, N. (2013).
+\newblock Is social simulation a social science outstation? a bibliometric
+  analysis of the impact of jasss.
+\newblock {\em Journal of Artificial Societies and Social Simulation},
+  16(1):10.
+
+\bibitem[Takizawa et~al., 2000]{takizawa2000simulation}
+Takizawa, A., Kawamura, H., and Tani, A. (2000).
+\newblock Simulation of urban system``as it could be''by multiagent model.
+\newblock In {\em Computing in Civil and Building Engineering (2000)}, pages
+  1566--1573.
+
+\bibitem[Taylor~Buck, 2017]{taylor2017art}
+Taylor~Buck, N. (2017).
+\newblock The art of imitating life: The potential contribution of biomimicry
+  in shaping the future of our cities.
+\newblock {\em Environment and Planning B: Urban Analytics and City Science},
+  44(1):120--140.
+
+\bibitem[Thrun and Montemerlo, 2006]{thrun2006graph}
+Thrun, S. and Montemerlo, M. (2006).
+\newblock The graph slam algorithm with applications to large-scale mapping of
+  urban structures.
+\newblock {\em The International Journal of Robotics Research},
+  25(5-6):403--429.
+
+\bibitem[Torrens, 2003]{torrens2003automata}
+Torrens, P.~M. (2003).
+\newblock Automata-based models of urban systems.
+\newblock {\em Advanced spatial analysis}, pages 61--79.
+
+\bibitem[Van~Raan and Tijssen, 1993]{van1993neural}
+Van~Raan, A. and Tijssen, R. (1993).
+\newblock The neural net of neural network research: An exercise in
+  bibliometric mapping.
+\newblock {\em Scientometrics}, 26(1):169--192.
+
+\bibitem[Vlahogianni et~al., 2005]{vlahogianni2005optimized}
+Vlahogianni, E.~I., Karlaftis, M.~G., and Golias, J.~C. (2005).
+\newblock Optimized and meta-optimized neural networks for short-term traffic
+  flow prediction: A genetic approach.
+\newblock {\em Transportation Research Part C: Emerging Technologies},
+  13(3):211--234.
+
+\bibitem[Votsis and Haavisto, 2019]{votsis2019urban}
+Votsis, A. and Haavisto, R. (2019).
+\newblock Urban dna and sustainable cities: A multi-city comparison.
+\newblock {\em Frontiers in Environmental Science}, 7:4.
+
+\bibitem[Wang, 2007]{wang2007toward}
+Wang, F.-Y. (2007).
+\newblock Toward a paradigm shift in social computing: The acp approach.
+\newblock {\em IEEE Intelligent Systems}, 22(5):65--67.
+
+\bibitem[White and Engelen, 1993]{white1993cellular}
+White, R. and Engelen, G. (1993).
+\newblock Cellular automata and fractal urban form: a cellular modelling
+  approach to the evolution of urban land-use patterns.
+\newblock {\em Environment and planning A}, 25(8):1175--1199.
+
+\bibitem[White, 1989]{white1989artificial}
+White, R.~W. (1989).
+\newblock The artificial intelligence of urban dynamics: Neural network
+  modeling of urban structure.
+\newblock In {\em Papers of the Regional Science Association}, volume~67, pages
+  43--53. Springer.
+
+\bibitem[Whitelaw, 2004]{whitelaw2004metacreation}
+Whitelaw, M. (2004).
+\newblock {\em Metacreation: art and artificial life}.
+\newblock Mit Press.
+
+\bibitem[Wu and Silva, 2010]{wu2010artificial}
+Wu, N. and Silva, E.~A. (2010).
+\newblock Artificial intelligence solutions for urban land dynamics: a review.
+\newblock {\em Journal of Planning Literature}, 24(3):246--265.
+
+\bibitem[Zheng et~al., 2014]{zheng2014urban}
+Zheng, Y., Capra, L., Wolfson, O., and Yang, H. (2014).
+\newblock Urban computing: concepts, methodologies, and applications.
+\newblock {\em ACM Transactions on Intelligent Systems and Technology (TIST)},
+  5(3):1--55.
+
+\end{thebibliography}
 

### 2022-08-30 Docs/Papers/ALifeCities/paper/paper.tex
papers
reviews_alife2020_alifecities

@@
-\documentclass[letterpaper]{article}
+\documentclass[10pt]{article}
 
-\usepackage{natbib,alifeconf}  %% The order is important
-
-% *****************
-%  Requirements:
-% *****************
-%
-% - All pages sized consistently at 8.5 x 11 inches (US letter size).
-% - PDF length <= 8 pages for full papers, <=2 pages for extended
-%    abstracts (not including citations).
-% - Abstract length <= 250 words.
-% - No visible crop marks.
-% - Images at no greater than 300 dpi, scaled at 100%.
-% - Embedded open type fonts only.
-% - All layers flattened.
-% - No attachments.
-% - All desired links active in the files.
-
-% Note that the PDF file must not exceed 5 MB if it is to be indexed
-% by Google Scholar. Additional information about Google Scholar
-% can be found here:
-% http://www.google.com/intl/en/scholar/inclusion.html.
-
-% If your system does not generate letter format documents by default,
-% you can use the following workflow:
-% latex example
-% bibtex example
-% latex example ; latex example
-% dvips -o example.ps -t letterSize example.dvi
-% ps2pdf example.ps example.pdf
-
-% For pdflatex users:
-% The alifeconf style file loads the "graphicx" package, and
-% this may lead some users of pdflatex to experience problems.
-% These can be fixed by editing the alifeconf.sty file to specify:
-% \usepackage[pdftex]{graphicx}
-%   instead of
-% \usepackage{graphicx}.
-% The PDF output generated by pdflatex should match the required
-% specifications and obviously the dvips and ps2pdf steps become
-% unnecessary.
-
-% Note:  Some laser printers have a serious problem printing TeX
-% output. The use of ps type I fonts should avoid this problem.
+\usepackage{natbib}
+\usepackage[margin=2cm]{geometry}
 
@@
 
-\title{\vspace{-2cm}Cities as they could be: Artificial Life and Urban Systems}
+\title{Cities as they could be: Artificial Life and Urban Systems}
 \author{Juste Raimbault$^{1,2,3}$\\
@@
 $^3$UMR CNRS 8504 G{\'e}ographie-cit{\'e}s\\
-juste.raimbault@polytechnique.edu} % email of corresponding author
+juste.raimbault@polytechnique.edu}
 
-% For several authors from the same institution use the same number to
-% refer to one address.
+% reviews [01/06/2020]
+% Dear Submitters and Authors,
+
+%Thank you for your submission to ALife 2020. Please see below for reviews on your submissions.
+
+%SUBMISSION: 76
+%TITLE: Cities as they could be: Artificial Life and Urban Systems
+
+%----------------------- REVIEW 1 ---------------------
+%SUBMISSION: 76
+%TITLE: Cities as they could be: Artificial Life and Urban Systems
+%AUTHORS: Juste Raimbault
+
+%----------- Overall evaluation -----------
+%SCORE: 1 (weak accept)
+%----- TEXT:
+%This is a useful synthesis of the existing literature on the crossover between alife and urbanism.
+As populations increasingly move to cities, coming up with new and creative ways of handling this transition will become a more pressing problem, so this analysis is timely.
+
+%The list of possible areas for development is well-motivated and useful.
+
+%With regards to the keywords chosen for the bibliometric analysis - how specific are these?
+I experimented by typing "artificial life urban" into Google Scholar (the engine used here), and got a reasonable number of hits on papers that I would not necessarily expect to consider to be in scope...
+(e.g., "The urban matrix and artificial light restricts the nightly ranging behaviour of Gould's long‐eared bat (Nyctophilus gouldi)") Why not "artificial life" as a phrase, rather than as individual terms?....
+So, my confidence in the analysis is very slightly undermined...
+The author does state that they manually removed some "obviously" irrelevant papers, but this clearly introduces a degree of subjectivity...
+However, the overall results would appear to be reasonably robust...
+
+%I would like the diagrams to be slightly more readable (specifically, Fig 1), although I appreciate the constraints on presentation.
+
+%----------------------- REVIEW 2 ---------------------
+%SUBMISSION: 76
+%TITLE: Cities as they could be: Artificial Life and Urban Systems
+%AUTHORS: Juste Raimbault
+
+%----------- Overall evaluation -----------
+%SCORE: -2 (reject)
+%----- TEXT:
+%This paper presents a bibliometric analysis of literature at the intersection of artificial intelligence, artificial life, and urban systems.
+225,000 papers are identified, and their citation networks are presented.
+Then, some analysis describes communities within these research areas, and identifies 'flows' of citations (i.e.
+implying transference of ideas) between them.
+
+%The paper describes its methodology well, and the analysis provides some interesting things of note.
+However, the scope of the search is quite vague and arbitrary ('urban' and 'artificial intelligence' are included, but not 'city' or 'machine learning', for example) and any actionable insights for ALife researchers are not sufficiently present for me to judge that the paper yet has publishable value for the community.
+
+%Additionally, the paper unfortunately has quite a few language errors that make it difficult to understand in places. I would recommend obtaining feedback from a copy editor.
+
+%----------------------- REVIEW 3 ---------------------
+%SUBMISSION: 76
+%TITLE: Cities as they could be: Artificial Life and Urban Systems
+%AUTHORS: Juste Raimbault
 %
-% If the names do not fit well on one line use
-%         Author 1, Author 2 ... \\ {\Large\bf Author n} ...\\ ...
-%
-% If the title and author information do not fit in the area
-% allocated, place \setlength\titlebox{<new height>} after the
-% \documentclass line where <new height> is 2.25in
+%----------- Overall evaluation -----------
+%SCORE: 1 (weak accept)
+%----- TEXT:
+%Summary:
+%In this paper, the authors explore the Artificial Life and Intelligence methodologies, tools, concepts, and application domains to enhance the understanding and management of cities.
+%Their exploration method consists of literature mapping of nearly 225,000 papers in different research areas related to ALife or/and AI. 
+%The result is a list of distinct communities and domains related to ALife and AI concepts, the relation between them and how each community uses ALife and AI applications and concepts in general and particularly in urban systems.
+After the bibliometric analysis, they also develop several concepts which are considered to be under-explored in the literature.
+
+%Contribution to science:
+%The main contribution of this paper is the Network Graph linking all the communities using ALife and AI concepts and more importantly the sub Network that emphasizes the use of ALife in urban systems.
+%This contribution is interesting because it can be considered as a taxonomy in the field of ALife in Urban Systems.
+
+%The shortcomings in this paper are stated as follows:
+%- Errors in typography, in the abstract line 03 : "We propose in this paper", not "in that paper".
+%- Errors in typography, in the introduction line 19 : "It is", not "is is"
+%- The first figure representing the Network Graph of the citation network for ALife and AI applications to urban systems is not readable, it contains too much text.
+
+%The biggest faults are stated as follows:
+%- Artificial Intelligence concepts are presented briefly without any further detail (e.g, Neural Networks)
+%- The use cases of AI mentioned in the paper are shallow and not clear (e.g, AI can be applied to manage highway traffic)
+%- The use of vague expressions when stating AI which keeps the subject broad and unreachable (e.g, We saw tools and methods, in particular Artificial Intelligence methods, are already widely applied)
+
+%Verdict:
+%It is an interesting work because it provides a good bibliometric analysis and makes clear the existing correlation between communities related to urban systems and using ALife concepts.
+This paper may be considered as a starting point for further exploration and concept development but the scientific contribution is weak.
 
 \begin{document}
+
 \maketitle
@@
 \begin{abstract}
-% Abstract length should not exceed 250 words
 The metaphor of cities as organisms has a long history in urban planning, and a few urban modeling approaches have explicitly been linked to Artificial Life.
@@
 
+% ALife approaches to urban systems
+% Bibliometrics
+% Urban complexity
+
 % \cite{pattee1995artificial} epistemic cut: simulation / hard or wet ALife / nature of computation % not relevant here?
 
-\cite{doi:10.1098/rstb.2019.0329}
+%\cite{doi:10.1098/rstb.2019.0329}
 

### 2023-09-14 Docs/Communications/2023/ECTQG2023/slides.tex
slides_ectqg

\documentclass[english,11pt]{beamer}

\DeclareMathOperator{\Cov}{Cov}
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Proba}{\mathbb{P}}

\newcommand{\Covb}[2]{\ensuremath{\Cov\!\left[#1,#2\right]}}
\newcommand{\Eb}[1]{\ensuremath{\E\!\left[#1\right]}}
\newcommand{\Pb}[1]{\ensuremath{\Proba\!\left[#1\right]}}
\newcommand{\Varb}[1]{\ensuremath{\Var\!\left[#1\right]}}

% norm
\newcommand{\norm}[1]{\| #1 \|}

\newcommand{\indep}{\rotatebox[origin=c]{90}{$\models$}}

\usepackage{mathptmx,amsmath,amssymb,graphicx,bibentry,bbm,babel,ragged2e}

\makeatletter

\newcommand{\noun}[1]{\textsc{#1}}
\newcommand{\jitem}[1]{\item \begin{justify} #1 \end{justify} \vfill{}}
\newcommand{\sframe}[2]{\frame{\frametitle{#1} #2}}

\newenvironment{centercolumns}{\begin{columns}[c]}{\end{columns}}
%\newenvironment{jitem}{\begin{justify}\begin{itemize}}{\end{itemize}\end{justify}}

\usetheme{Warsaw}
\setbeamertemplate{footline}[text line]{}
\setbeamercolor{structure}{fg=purple!50!blue, bg=purple!50!blue}

\setbeamersize{text margin left=15pt,text margin right=15pt}

\setbeamercovered{transparent}

\@ifundefined{showcaptionsetup}{}{%
 \PassOptionsToPackage{caption=false}{subfig}}
\usepackage{subfig}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

\usepackage{multirow}

\makeatother

\begin{document}

\title{Multiscalar models for systems of cities}

\author{J.~Raimbault$^{1,2,3,4,\ast}$\\
\texttt{$\ast$ juste.raimbault@ign.fr}
}

\institute{$^{1}$ LASTIG, IGN-ENSG\\
$^{2}$CASA, UCL\\
$^{3}$UPS CNRS 3611 ISC-PIF\\
$^{4}$UMR CNRS 8504 G{\'e}ographie-cit{\'e}s
}

\date{ECTQG 2023\\\smallskip
September 15th 2023
}

\frame{\maketitle}

\sframe{An evolutionary theory for urban systems}{

% Simulation models for the dynamics of systems of cities have been developed to capture salient stylised facts of such systems and understand underlying processes (Pumain & Reuillon, 2017). These agent-based models simulate the interactions between cities at the macroscopic scale, focusing on different dimensions such as innovation or economic exchanges. They correspond to the highest level of the multiscalar ontology described by Pumain (2011) for urban systems.

\begin{center}
	\begin{columns}
	\begin{column}{0.45\linewidth}
	\includegraphics[width=\textwidth]{../../2020/ALife2020/figures/evoltheory_scales.png}
	\end{column}
	\begin{column}{0.22\linewidth}
	\includegraphics[width=\textwidth]{../../2020/ALife2020/figures/simpop1.png}\\\medskip
	\includegraphics[width=\textwidth]{../../2020/ALife2020/figures/setup_synth_1_tick100.png}
	\end{column}
	\end{columns}
\end{center}

\footnotesize
\textit{An evolutionary urban theory considering cities as systems within systems of cities \cite{pumain2018evolutionary}; Simpop 1 model \cite{sanders1997simpop}; SimpopNet model \cite{schmitt2014modelisation}}

}

\sframe{Multi-modeling urban systems dynamics}{

% sdg tradeoffs, multiobj ersa2022

}

\sframe{Hybrid models}{

% Lutecia and physical network: "semi-multiscale"

}

\sframe{Towards multiscale models}{

% In order to address open issues related to multi-level governance of territorial sustainable transitions, models simulating simultaneously multiple scales are however needed (Rozenblat & Pumain, 2018).

% contribution

}

\sframe{Urban morphology and inter-city interactions}{

% This contribution synthesises recent lessons learnt from the construction and exploration of such strongly coupled multi-scalar models. The first model couples a macroscopic population dynamics model with local urban morphogenesis models within each city (Raimbault, 2021),

% broad description of morph-intgib model: cit ccs19 and preprint

}

\sframe{Model exploration}{

% salient results for urbmorph-intgib

}

\sframe{Multi-scale innovation dynamics}{

% while the second one simulates the diffusion of innovation between urban areas and its impact on urban growth at the macroscopic scale, and innovation cluster dynamics within each area (Raimbault & Pumain, 2023).

}

\sframe{Model exploration}{

}

\sframe{Common aspects from simulation results}{

% First simulation results show that strong emergence appears in a significant number of regimes for the innovation model, and that contradictory objectives across scales can be simultaneously optimised using multi-objective genetic algorithms for both models.

}

\sframe{Teaching 1: ontologies and feedbacks}{

% A few pitfalls have been identified while theoretically constructing the models and empirically exploring their parameter space using advanced model validation techniques with the OpenMOLE software (Reuillon et al., 2013): (i) to obtain a strong coupling between scales, and thus effective multiscalar dynamics rather than fixed effects at the meso scale only for example, both top-down and bottom-up feedbacks have to be included explicitly, and corresponding ontologies and processes must be identified;

}

\sframe{Teaching 2: consistent data parametrisation}{

%ii) the parametrisation or calibration with empirical data is significantly more cumbersome than with single level models, and synthetic systems are a first alternative to explore model behaviour;

}

\sframe{Teaching 3: stochasticity}{

% (iii) the convergence of model indicators regarding stochasticity seems more difficult to obtain, possibly due to non-linear noise propagation between scales;

% TODO histogram/distib for illustrated indicators for the innovation model?

}

\sframe{Teaching 4: quantifying strong emergence}{

% (iv) a crucial aspect remains to quantify the strength of emergence, such that these approaches are not a superfluous complication of simpler dynamics at a single scale – the indicators introduced by (Rosas et al., 2020) are good candidates for such measures but remain to be tested systematically on geosimulation models.

}

\sframe{Discussion}{

% \textbf{Main results:}

\justify

$\rightarrow$

\bigskip

$\rightarrow$

\bigskip

%\textbf{Conclusion: }

}

%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[allowframebreaks]
\frametitle{References}
\bibliographystyle{apalike}
\bibliography{biblio}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2023-09-14 Docs/Communications/2023/ECTQG2023/slides.tex
slides

@@
 \date{ECTQG 2023\\\smallskip
+Session Theoretical Geography 1\\\smallskip
 September 15th 2023
@@
 	\begin{columns}
-	\begin{column}{0.45\linewidth}
+	\begin{column}{0.49\linewidth}
 	\includegraphics[width=\textwidth]{../../2020/ALife2020/figures/evoltheory_scales.png}
 	\end{column}
-	\begin{column}{0.22\linewidth}
+	\begin{column}{0.3\linewidth}
 	\includegraphics[width=\textwidth]{../../2020/ALife2020/figures/simpop1.png}\\\medskip
@@
 
-\footnotesize
-\textit{An evolutionary urban theory considering cities as systems within systems of cities \cite{pumain2018evolutionary}; Simpop 1 model \cite{sanders1997simpop}; SimpopNet model \cite{schmitt2014modelisation}}
+%\footnotesize
+An evolutionary urban theory considering cities as systems within systems of cities \cite{pumain2018evolutionary}; Simpop 1 model \cite{sanders1997simpop}; SimpopNet model \cite{schmitt2014modelisation}
 
@@
 
-\sframe{Multi-modeling urban systems dynamics}{
+\sframe{Multi-modeling urban systems dynamics for sustainability}{
 
 % sdg tradeoffs, multiobj ersa2022
+
+\begin{center}
+	\includegraphics[width=0.51\textwidth]{../../../Papers/SDGTradeoffs/SDGTradeoffs-paper/figures/Fig3.png}
+	\includegraphics[width=0.42\textwidth]{../../../../../SDGTradeoffs/Docs/Communications/2022/FRCCS2022/figures/radar_sdgs.png}
+\end{center}
+
+Trade-offs between sustainable development goals in systems of cities \cite{raimbault2020empowering}; multi-modeling for many-objectives SDGs optimisation \cite{raimbault2022trade}
 
@@
 % Lutecia and physical network: "semi-multiscale"
+
+\begin{center}
+	\includegraphics[width=0.49\textwidth]{../../../../../CityNetwork/Docs/ThesisMemoire/Final/Figures/Final/7-3-3-fig-lutecia-governance.jpg}
+	\includegraphics[width=0.49\textwidth]{../../../../../NetworksTerritories/CoevolutionNwTerritories/Docs/Papers/HierarchyCoevolution/chapter/figuresraw/ex_realphysical_1975_tf.png}
+\end{center}
+
+Simulation models for systems of cities and transportation networks making a slight bridge between scales: transportation governance modeling \cite{raimbault2021introducing}; physical network self-reinforcment \cite{raimbault2020hierarchy}
 
@@
 % contribution
+
+$\rightarrow$ multi-scale simulation models are needed to explore sustainable policies \cite{rozenblat2018conclusion}
+
+\bigskip
+
+$\rightarrow$ multi-scale urban models remain however underexplored in the literature
+
+\bigskip
+\bigskip
+
+\textbf{This contribution: } \textit{to synthesise lessons from two different recent tentatives to build multi-scale urban models with a strong coupling between scales: an urban morphogenesis model \cite{raimbault2021strong} and an innovation diffusion model \cite{raimbault2023innovation}, learnt from the systematic exploration of these models using OpenMOLE \cite{reuillon2013openmole}}
 
@@
 
+\begin{center}
+\includegraphics[width=\textwidth]{../../../../../UrbanGrowth/Docs/Papers/Multiscale/paper/figures/Fig1.jpg}
+\end{center}
+
+A strong coupling explored by \cite{raimbault2021strong} between urban morphogenesis agregation-diffusion processes at the meso scale \cite{raimbault2018calibration} and interactions between cities at the macro scale \cite{raimbault2020indirect}
+
 }
@@
 % salient results for urbmorph-intgib
+
+\begin{center}
+\includegraphics[width=0.58\textwidth]{../../../../../UrbanGrowth/Docs/Papers/Multiscale/paper/figures/Fig5.jpg}
+\includegraphics[width=0.4\textwidth]{../../../../../UrbanGrowth/Docs/Papers/Multiscale/paper/figures/Fig7.jpg}
+\end{center}
+
+\bigskip
+
+(Left) Non-monotonous effect of policy parameters linking scales; (Right) Bi-objective optimisation of contradictory objectives across scales
 
@@
 
+\begin{center}
+	\includegraphics[width=0.75\textwidth]{../../../Papers/InnovationMultiscale/InnovationMultiscale-paper/slides/figures/model_4.png}
+\end{center}
+
+A bi-scale model explored by \cite{raimbault2023innovation}, coupling urban evolution based on innovation diffusion at the macro scale \cite{raimbault2020model}, with intra-urban innovation cluster dynamics between firms for the emergence of new innovations \cite{raimbault2022innovation}; quantification of strong emergence using indicators by \cite{rosas2020reconciling}
+
 }
@@
 \sframe{Model exploration}{
+
+\begin{center}
+\includegraphics[width=0.49\textwidth]{../../../Papers/InnovationMultiscale/InnovationMultiscale-paper/figures/paretoDiversity-Fitness_colordG_sizetheta.png}
+\includegraphics[width=0.49\textwidth]{../../../Papers/InnovationMultiscale/InnovationMultiscale-paper/figures/pse-psi-delta-utility_colorGamma_sizetheta.png}
+\end{center}
+
+\bigskip
+
+(Left) bi-optimisation between contradictory objectives across scales; (Right) diversity of emergence regimes obtained with a diversity search algorithm
 
@@
 % First simulation results show that strong emergence appears in a significant number of regimes for the innovation model, and that contradictory objectives across scales can be simultaneously optimised using multi-objective genetic algorithms for both models.
+
+$\rightarrow$ strong emergence quantified in a high diversity of regimes for the innovation diffusion model, probable (given cross-scale parameter influence, but not quantified) for the urban morphogenesis model
+
+\bigskip
+
+$\rightarrow$ optimisation objectives across scales seem generally contradictory, and can be simultaneously optimised with Pareto trade-offs
 

### 2023-10-12 Docs/Communications/2023/ECTQG2023/slides.tex

@@
 
-\sframe{Hybrid models}{
+\sframe{Hybrid models mixing scales}{
 
@@
 
+$\rightarrow$ models coupling scales with a single direction of inter-scale feedback can be simplified to non-stationary dynamics or fixed effects at a single scale: ``strong'' coupling is necessary
+
+\bigskip
+
+$\rightarrow$ scales have their own distinct ontologies with their own processes
+
+\bigskip
+
+% Q: is aggregation a process? not necessarily - but a feedback between variables
+
+$\rightarrow$ inter-scales processes may be necessary to capture feedbacks and may necessitate their own ontology
+
 }
@@
 %ii) the parametrisation or calibration with empirical data is significantly more cumbersome than with single level models, and synthetic systems are a first alternative to explore model behaviour;
+
+$\rightarrow$ model parametrisation with real data is more complicated: which scales?
+need of consistent datasets in case of multiple scales (cf Geodivercity project); how to consistently setup model states?
+\ldots
+
+\bigskip
+
+$\rightarrow$ synthetic systems of cities are a first way to explore model dynamics independently of geographical contingencies \cite{raimbault2019space}
 
@@
 
-% TODO histogram/distib for illustrated indicators for the innovation model?
+% histogram/distib for illustrated indicators for the innovation model? no time
+
+\begin{center}
+	\includegraphics[width=0.49\textwidth]{../../../../Results/InnovationMultiscale/exploration/20230623_135234_EXPLORATION/psiDiversity_MED-macroGravityDecay_color-macroInnovationDecay_facet-mesoToMacroInnovationThreshold-macroToMesoExchangeMaxUpdate_mesoCrossOverProba05.png}
+	\includegraphics[width=0.49\textwidth]{../../../../Results/InnovationMultiscale/exploration/20230705-20230712_EXPLORATION/psiDiversity-macroGravityDecay_color-macroInnovationDecay_facet-mesoToMacroInnovationThreshold-macroToMesoExchangeMaxUpdate_mesoCrossOverProba05.png}
+\end{center}
+
+% not sure 100? check oms commit! -> 56547b137cb2e8c807511e71efde1c06ba42ef6d : 20 repets only!
+
+$\rightarrow$ statistical convergence of indicators is more difficult to reach (example above for innovation diffusion: 20 vs 10,000 replications, despite ``good-looking'' empirical Sharpe ratios suggesting a low number of repetitions)
+
+\bigskip
+
+$\rightarrow$ non-linearities, fat-tail distribution, noise propagation between scales? More uncertainty-based/Bayesian validation methods?
 
@@
 
+$\rightarrow$ issue of over-complication/overfitting (generic issue for simulation models \cite{raimbault2020indirect}): is the multi-scale aspect necessary?
+
+\bigskip
+
+$\rightarrow$ the stronger the emergence, the more it is needed?
+\cite{bedau2002downward}: for nominal emergence, can be simplified/surrogated; for weak emergence, depends on the cases?
+(cf Schelling model); for strong emergence difficult to be avoided because of the autonomy?
+
+\bigskip
+
+$\rightarrow$ need indicators and methods to systematically quantify emergence in multi-scalar simulated dynamics: cf. information-theory-based indicators proposed by \cite{rosas2020reconciling}
+
+\bigskip
+
+$\rightarrow$ link between strength of emergence and strength of coupling, and other types of complexities (such as \cite{pumain2003approche}'s ``interdisciplinary'' complexity), seems to be an open question, related to the nature of complexities \cite{raimbault2020relating} and complex knowledge/knowledge of the complex
+
 }
 
-\sframe{Discussion}{
+\sframe{Discussion: still a long way to go!}{
 
-% \textbf{Main results:}
+%\textbf{Main results:}
 
@@
 
-$\rightarrow$
+$\rightarrow$ there seem to be methodological regularities when building strongly coupled multi-scale models - to what extent do they depend on the type of system, methodological framework, do they hold with a parametrisation on real data?
 
@@
 
-$\rightarrow$
+$\rightarrow$ subtle compromise between model complication/complexity and the effective capture of multi-scale dynamics
 
@@
 
+$\rightarrow$ application to multi-scalar policies still far, but already possible in a stylised way such as for trade-offs between SDGs (at a single scale for now \cite{raimbault2022trade})
+
+\bigskip
+\bigskip
+
 %\textbf{Conclusion: }
+
+\textbf{Open repositories for presented models:}
+
+\texttt{https://github.com/JusteRaimbault/UrbanGrowth}
+
+\texttt{https://github.com/JusteRaimbault/InnovationMultiscale-model}
 
